

```{r setup, warning = FALSE, message=FALSE}

library(tidyverse)
library(haven)
library(formatR)
library(lubridate)
library(smooth)
library(forecast)
library(scales)
library(kableExtra)
library(ggplot2)
library(readxl)
library(tidyverse)
library(data.table)
library(quantmod)
library(geofacet)
library(janitor)


knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

```

# Data Creation and Cleaning



```{r rename-variables, eval=FALSE}

#Example code below: Read in excel file and rename columns so that it plays well with the other years' files.
revenue_fy22 <- read_xlsx("Fis_Fut_Rev_2022_Final.xlsx") %>% 
  rename(fy = 'FY',
         fund = 'FUND',
         fund_name = 'FUND NAME',
         agency = 'AGENCY',
         agency_name = 'AGENCY NAME',
         source = 'REVENUE SOURCE',
         source_name = 'REV SRC NAME',
         receipts = 'AMOUNT'
  ) 

exp_fy22 <- read_xlsx("Fis_Fut_Exp_2022_Final.xlsx") %>% 
  rename(fy = 'FY',
         fund = 'FUND',
         fund_name = 'FUND NAME',
         agency = 'AGENCY',
         agency_name = 'AGENCY NAME',
         appr_org = 'ORGANIZATION',
         org_name = 'ORGANIZATION NAME',
         obj_seq_type = 'APPROPRIATION',
         wh_approp_name = 'APPROPRIATION NAME',
        # exp_net_xfer = 'NET OF TRANS AMOUNT',
         expenditure = 'EXPENDED'
  ) 

# %>%
#   # these come from ioc_source file after merging
#   mutate(data_source = "exp IOC Aug 2022",
#          object = ,
#          seq = ,
#          type = ,
#          fund_cat = FIND_COLUMN, #create fund_cat column
#          fund_cat_name = FIND_NAME) # create fund_cat_name column



```

 
The code chunk below takes the .dta files for all fiscal years before FY 2022 and binds them together.  Variable names were manually changed by past researchers so that they were consistent across years. If joining all files as excel documents, then use .xlsx extension instead.

For FY 2023 and after, .dta files can be avoided entirely and .csv or excel files will be used. All files before this year had been saved and passed on as .dta files for Stata code before the transition to R in Fall 2022.  For years after fy22: add line of code to bind csv files after binding the dta files together. Variable names must identical to merge files together.

```{r create-rev-csv-FY22, eval=FALSE}
# years after fy22: add line of code to bind csv files after binding the dta files.
setwd("C:/Users/aleaw/OneDrive/Documents/PhD Fall 2021 - Spring 2022/Merriman RA/Fiscal Futures FY2022/Replication-Files/revenue")

allrevfiles22 = list.files(path = "C:/Users/aleaw/OneDrive/Documents/PhD Fall 2021 - Spring 2022/Merriman RA/Fiscal Futures FY2022/Replication-Files/revenue",  pattern = ".dta") %>%  lapply(read_dta) %>% bind_rows
#Fy21: 62294 observations, 13 variables
#FY22: 65094 obs, 13 vars

#write_csv(allrevfiles22, "allrevfiles.csv")




setwd("C:/Users/aleaw/OneDrive/Documents/PhD Fall 2021 - Spring 2022/Merriman RA/Fiscal Futures FY2022/Replication-Files/expenditures")

allexpfiles22 = list.files(path = "C:/Users/aleaw/OneDrive/Documents/PhD Fall 2021 - Spring 2022/Merriman RA/Fiscal Futures FY2022/Replication-Files/expenditures",  pattern = ".dta") %>%  lapply(read_dta) %>% bind_rows

#fy21 213372 observations, 20 variables
# fy22 225587 obs, 21 vars.

#write_csv(allexpfiles22, "allexpfiles.csv")


```


Code below reads in the csv files created in chunks above (allrevfiles.csv and allrexpfiles.csv). These files contain all years of data combined into one file BEFORE any recoding is done. Do not use this file for summing categories because it is just an inbetween step before recoding revenue and expenditure categories. It also does not have funds or agencies tracked consistently over time. Essentially just the raw data for all years combined.

Once confident in the joined file, I saved it as allrevfiles and allexpfiles so that I didn't have to create the file every time I ran the code. 

```{r readCSVs}
# combined in past chunks called create-rev-csv and create-exp-csv

allrevfiles22 <- read_csv("allrevfiles22.csv") #combined but not recoded
allexpfiles22 <- read_csv("allexpfiles22.csv") #combined but not recoded

```


## Inspecting new FY22 data files

Data files for closed years have been obtained from IOC. The numbers of funds, agencies, organizations, and revenue sources below were found by using pivot tables in the codebook files in the FY2022 Box folder. This could also be done using R and grouping new files by fund, agency, source number, source names, etc..

Revenue File:  
- 684 Fund Numbers   
- 80 Agencies  
- 1184 Revenue source numbers
- 1156 revenue source names

Expenditure File:  
- 708 Fund Numbers   
- 107 Agencies   
- 98 Organization Numbers   
- 313 Organization names  

### Finding new agencies and funds

General steps:  

1. Identify new and reused funds for newest fiscal year.    
2. Recode funds to take into account different fund numbers/names over the years. See [Recoding New and Reused Funds] for code chunk that does this.  
3. Update fund_ab_in_2022.xlsx with any changes from previous fiscal year.     




**New Agencies, Funds, and Organizations from Expenditure files:**

- Using the code below, I found 2 agencies, multiple funds, and a couple organizations that had not been specifically mentioned in the funds_ab_in file - AWM, FY2022.
  - Note: One new org name and org number combo has an expenditure of $600 million for pensions payment. Investigate later.  

```{r}

allrevfiles22 <- read_csv("allrevfiles22.csv") #combined but not recoded
allexpfiles22 <- read_csv("allexpfiles22.csv") #combined but not recoded

#### From Expenditure Data #####

# agencies referenced in any year before 2020:
agencies_past <- allexpfiles22 %>% 
  filter(fy < 2022) %>% 
  mutate(agency == as.character(agency)) %>% 
  group_by(agency, agency_name) %>% unique() %>% 
  summarize(expenditure = sum(expenditure, na.rm = TRUE)) %>% 
  drop_na() %>% 
  arrange(agency)
# agencies_past # 146 agencies ever


# agencies in 2022 data:
agencies22 <- allexpfiles22  %>% 
  filter(fy == 2022) %>% 
  mutate(agency == as.character(agency)) %>% 
  group_by(agency, agency_name) %>% 
  summarize(expenditure = sum(expenditure, na.rm = TRUE))
# agencies22 # 107 agencies this year


# 280 and 533 are new agency codes: 
anti_join(agencies22, agencies_past, 
          by = c("agency", "agency_name")) %>% 
  arrange(agency)




funds_past <- allexpfiles22  %>% 
  filter(fy < 2022) %>% 
  mutate(fund == as.character(fund)) %>% 
  group_by(fund, fund_name) %>% 
  summarize(count = n(), Expenditure = sum(expenditure, na.rm = TRUE))  %>% 
  drop_na()

funds22 <- allexpfiles22  %>% 
  filter(fy == 2022) %>% 
  mutate(fund == as.character(fund)) %>% 
  group_by(fund, fund_name) %>%  
  summarize(count = n(), Expenditure = sum(expenditure, na.rm = TRUE)) %>% 
  unique()


# 15 funds were in FY22 data that were not in past data:
anti_join(funds22, funds_past, 
          by = c("fund", "fund_name")) %>% 
  arrange(fund)

# orgs_pastin the past =  916 org groups ever
orgs_past <- allexpfiles22  %>% 
  filter(fy < 2022) %>% 
  mutate(appr_org == as.character(appr_org)) %>% 
  group_by(appr_org, org_name) %>% unique() %>% 
  summarize(Expenditure = sum(expenditure, na.rm = TRUE)) %>% 
  drop_na()

# orgs_past # 916 org groups ever
orgs22 <- allexpfiles22  %>% 
  filter(fy == 2022) %>% 
  mutate(appr_org = as.character(appr_org)) %>% 
  group_by(appr_org, org_name) %>% 
  summarize(Expenditure = sum(expenditure, na.rm = TRUE))
# orgs22 # 396 org groups this year


# 4 org number and org name combos are new for FY2022:
anti_join(orgs22, orgs_past,
          by = c("appr_org", "org_name")) %>% 
  arrange(appr_org)
```


**New Revenue Funds, Sources, and New Agencies:**

```{r  }
#### From Revenue Data ####



# agencies_past # 108 agencies ever
agencies_past <- allrevfiles22  %>% filter(fy < 2022) %>% 
  mutate(agency == as.character(agency)) %>% 
  group_by(agency, agency_name) %>% unique() %>% 
  summarize(Receipts = sum(receipts, na.rm = TRUE)) %>% 
  drop_na()


# agencies22 # 80 agencies this year
agencies22 <- allrevfiles22  %>% 
  filter(fy == 2022) %>% 
  mutate(agency == as.character(agency)) %>% 
  group_by(agency, agency_name) %>% 
  summarize(Receipts = sum(receipts, na.rm = TRUE))

# 0 new agencies in revenue data this year
anti_join(agencies22, agencies_past, by = c("agency", "agency_name")) %>% 
  arrange(agency)


funds_past <- allrevfiles22  %>% 
  filter(fy < 2022) %>% 
  mutate(fund == as.character(fund)) %>% 
  group_by(fund, fund_name) %>% 
  summarize(count = n(), Receipts = sum(receipts, na.rm = TRUE))  %>% 
  drop_na()

funds22 <- allrevfiles22  %>% 
  filter(fy == 2022) %>% 
  mutate(fund == as.character(fund)) %>% 
  group_by(fund, fund_name) %>%  
  summarize(count = n(), Receipts = sum(receipts, na.rm = TRUE)) %>% 
  unique() %>% 
  drop_na()

# 19 revenue funds were in FY22 revenue data that were not in past data 
# some could be small fund name changes
anti_join(funds22, funds_past, by = c("fund", "fund_name")) %>% 
  arrange(fund)



sources_past <- allrevfiles22  %>% 
  filter(fy < 2022) %>% 
  mutate(source == as.character(source)) %>% 
  group_by(source, source_name) %>% 
  summarize(count = n(), Receipts = sum(receipts, na.rm = TRUE))  %>% 
  drop_na()

sources22 <- allrevfiles22  %>% 
  filter(fy == 2022) %>% 
  mutate(source == as.character(source)) %>% 
  group_by(source, source_name) %>% 
  summarize(count = n(), Receipts = sum(receipts, na.rm = TRUE)) %>% 
  unique()


# 20 revenue sources were in FY22 data that were not in past data 
# some could be small source name changes:
anti_join(sources22, sources_past, by = c("source", "source_name")) %>% 
  arrange(source)

```

Sources 2737 through 2756 were not found in the IOC_source file so I added them to `ioc_source_updated22_AWM.xlsx`. They do NOT have a rev_type in the file so each source must be searched on the Comptrollers website. We could ask for a variable key from the comptroller for a list of all revenue sources and their revenue type in the future to make our lives easier. I had to look at the sources within each revenue type to find the ones missing and then manually update our ioc_source file. Most of these were considered type 39, "Licenses, Fees, & Registrations". These were finally updated on January 20th 2023 after totals were created for the FY22 paper. 

> To do: Double check ioc_source file "local" variable. Make sure it is still accurate and that it isn't missing any local sources.  


### Recoding New and Reused Funds

New funds will need to be manually added to the funds_ab_in excel file and determined if they should or should not be included in Fiscal Future calculations.

For funds that were reused once, a 9 replaces the 0 as the first digit. If reused twice, then the first two values are 10.    

- Ex. 0350 --> 9350 because its use changed.   
- Ex. 0367 becomes 10367 because its use has changed twice now. There was fund 0367 originally, then its use changed and it was recoded as 9367, and now it changed again so it is a 10367.  
- Excel file also has alternative ways to name funds (e.g. 0397-A and 0397-B) and variables for the year that the fund stopped being used. These have not been updated consistently over the years but it is useful information when trying to find any coding mistakes from the past.     

New or reused funds revenue file recoding:
```{r recode-rev-funds}
# if first character is a 0, replace with a 9 if its purpose has changed

rev_1998_2022 <- allrevfiles22 %>%
      mutate(fund = ifelse(fy < 2002 & fund %in% c("0730", "0241", "0350", "0367", "0381", "0382", "0526", "0603", "0734", "0913", "0379"), str_replace(fund, "0","9"), fund)) %>%
  
  mutate(fund = ifelse (fy < 2008 & fund %in% c("0027", "0033", "0037", "0058", "0062", "0066", "0075", "0083", "0116", "0119", "0120", "0122", "0148", "0149", "0157", "0158", "0166", "0194", "0201", "0209", "0211", "0217", "0223", "0231", "0234", "0253", "0320", "0503", "0505", "0512", "0516", "0531", "0532", "0533", "0547", "0563", "0579", "0591", "0606", "0616", "0624", "0659", "0662", "0665", "0676", "0710", 
"0068", "0076", "0115", "0119", "0168", "0182", "0199", "0241", "0307", "0506", "0509", "0513"), str_replace(fund, "0","9"), fund)) %>%
  
  mutate(fund = ifelse(fy < 2016 & fund %in% c("0263", "0399", "0409"), str_replace(fund, "0","9"), fund)) %>%
  
  mutate(fund =  ifelse(fy < 2017 & fund == "0364", str_replace(fund, "0","9"), fund)) %>%
  
  mutate(fund =  ifelse(fy < 2018 & fund %in% c("0818", "0767", "0671", "0593", "0578"), str_replace(fund, "0","9"), fund)) %>%
  
  mutate(fund = ifelse(fy>1999 & fy < 2018 & fund == "0231", "10231", fund) ) %>%
  
  mutate(fund = ifelse(fy < 2019 & fund %in% c("0161", "0489", "0500", "0612", "0893", "0766"), str_replace(fund, "0","9"), fund)) %>%
  
  mutate(fund =  ifelse(fy < 2020 & fund %in% c("0254", "0304", "0324", "0610", "0887", "0908", "0939", "0968"), str_replace(fund, "0","9"), fund)) %>%
  
  mutate(fund =  ifelse(fy < 2021 & fund %in% c("0255", "0325", "0348", "0967", "0972"), str_replace(fund, "0","9"), fund) ) %>%
  
   #2022 changes
  mutate(fund = ifelse(fy < 2022 & fund %in% c("0110","0165","0351", "0392", "0393", "0422", "0544", "0628", "0634",  "0656", "0672", "0683", "0723", "0742", "0743"), str_replace(fund, "0","9"), as.character(fund))) %>%  # replaces first 0 it finds with a 9
  mutate(fund = ifelse(fy < 2022 & fund == "0367", "10367", as.character(fund)) # fund reused for 3rd time
)
```

For 2022, new funds included The Essential Government Services fund, Electric Vehicle Rebate, Budget Stabilization fund, and more. New funds were added to the funds_ab_in.xlsx file and recoded funds were addressed in the code chunks below and updated in the same excel file. 


Expenditure recoding: 

```{r recode-exp-funds}
# if first character is a 0, replace with a 9

exp_1998_2022 <- allexpfiles22 %>%
      mutate(fund = ifelse(fy < 2002 & fund %in% c("0730", "0241", "0350", "0367", "0381", "0382", "0526", "0603", "0734", "0913", "0379"), str_replace(fund, "0","9"), fund)) %>%
  
  mutate(fund = ifelse(fy < 2008 & fund %in% c("0027", "0033", "0037", "0058", "0062", "0066", "0075", "0083", "0116", "0119", "0120", "0122", "0148", "0149", "0157", "0158", "0166", "0194", "0201", "0209", "0211", "0217", "0223", "0231", "0234", "0253", "0320", "0503", "0505", "0512", "0516", "0531", "0532", "0533", "0547", "0563", "0579", "0591", "0606", "0616", "0624", "0659", "0662", "0665", "0676", "0710", 
"0068", "0076", "0115", "0119", "0168", "0182", "0199", "0241", "0307", "0506", "0509", "0513"), str_replace(fund, "0","9"), fund)) %>%
  
  mutate(fund = ifelse(fy < 2016 & fund %in% c("0263", "0399", "0409"), str_replace(fund, "0","9"), fund)) %>%
  
  mutate(fund =  ifelse(fy < 2017 & fund == "0364", str_replace(fund, "0","9"), fund)) %>%
  
  mutate(fund =  ifelse(fy < 2018 & fund %in% c("0818", "0767", "0671", "0593", "0578"), str_replace(fund, "0","9"), fund)) %>%
  mutate(fund = ifelse(fy>1999 & fy < 2018 & fund == "0231", "10231", fund) ) %>%
  
  mutate(fund = ifelse(fy < 2019 & fund %in% c("0161", "0489", "0500", "0612", "0893", "0766"), str_replace(fund, "0","9"), fund)) %>%
  
  mutate(fund =  ifelse(fy < 2020 & fund %in% c("0254", "0304", "0324", "0610", "0887", "0908", "0939", "0968"), str_replace(fund, "0","9"), fund)) %>%
  
  mutate(fund =  ifelse(fy < 2021 & fund %in% c("0255", "0325", "0348", "0967", "0972"), str_replace(fund, "0","9"), fund))  %>%
  
  #2022 changes
  mutate(fund = ifelse(fy < 2022 & fund %in% c("0110","0165","0351", "0392", "0393", "0422", "0544", "0628", "0634",  "0656", "0672", "0683","0723", "0742", "0743"), str_replace(fund, "0","9"), as.character(fund))) %>%  # replaces first 0 it finds with a 9
  mutate(fund = ifelse(fy < 2022 & fund == "0367", "10367", as.character(fund)) # fund reused for 3rd time 
  )

```

> The `funds_ab_in.xlsx` file contains all funds that have existed since 1998, if they still exist, indicates if fund numbers have been reused for varying purposes, and is updated yearly with new fund numbers used by the IOC.


```{r create-exp_temp, warning = FALSE, message = FALSE}
funds_ab_in_2022 = readxl::read_excel("C:/Users/aleaw/OneDrive/Documents/PhD Fall 2021 - Spring 2022/Merriman RA/Fiscal Futures FY2022/Replication-Files/funds_ab_in_2022.xlsx")


exp_temp <- exp_1998_2022 %>% 
  arrange(fund, fy) %>%
  filter(expenditure != 0) %>%             # keeps everything that is not zero
# join  funds_ab_in_2021  to exp_temp
 left_join(funds_ab_in_2022, by = "fund")  # matches most recent fund number 


# Agency == 799 for Statutory transfers 
#  Object == 1993 is for Interfund cash transfers  
exp_temp <- exp_temp %>% 
  mutate(transfer = ifelse(org_name == "TRANSFERS", 1, 0),
         trans_agency = ifelse(org_name == "TRANSFERS", str_sub(obj_seq_type,1,3), trans_agency),
         trans_type = ifelse(org_name == "TRANSFERS", str_sub(obj_seq_type,4,9), trans_type)
         )
```


- the initial combined years of data are saved as dataframes named `exp_1998_2022` and `rev_1998_2022`. These are then saved as exp_temp and rev_temp while recoding variables. This is BEFORE category groups are created and cleaned below. `exp_temp` and `rev_temp` are only temporary files; do not use for aggregating totals yet! 


```{r remove-all_obs_df, include=FALSE}
#include = FALSE in the chunk settings run the code but do not include the chunk in the html output

# remove from computer memory to free up space (in case your computer needs it)
rm(allexpfiles22)
rm(allrevfiles22)
rm(exp_fy22)
rm(revenue_fy22)
```

Update Agencies: Some agencies have merged with others or changed names over time.
```{r agencies-exp}
# recodes old agency numbers to consistent agency number
exp_temp <- exp_temp %>% 
  mutate(agency = case_when(
    (agency=="438"| agency=="475" |agency == "505") ~ "440",
    # financial institution &  professional regulation &
     # banks and real estate  --> coded as  financial and professional reg
    agency == "473" ~ "588", # nuclear safety moved into IEMA
    (agency =="531" | agency =="577") ~ "532", # coded as EPA
    (agency =="556" | agency == "538") ~ "406", # coded as agriculture
    agency == "560" ~ "592", # IL finance authority (fire trucks and agriculture stuff)to state fire marshal
    agency == "570" & fund == "0011" ~ "494",   # city of Chicago road fund to transportation
    TRUE ~ (as.character(agency)))) 
```



For aggregating revenue, use the rev_1998_2022 dataframe, join the funds_ab_in_2022 file to it, and then join the ioc_source_type file to the dataset. Remember: You need to update the funds_ab_in and ioc_source_type file every year! 

```{r create-rev-temp, warning = FALSE, message=FALSE}
# fund info to revenue for all years
rev_temp <- inner_join(rev_1998_2022, funds_ab_in_2022, by = "fund") %>% arrange(source)

# need to update the ioc_source_type file every year! 
ioc_source_type <- readxl::read_xlsx("C:/Users/aleaw/OneDrive/Documents/PhD Fall 2021 - Spring 2022/Merriman RA/Fiscal Futures FY2022/Replication-Files/ioc_source_updated22_AWM.xlsx")

rev_temp <- left_join(rev_temp, ioc_source_type, by = "source")
# automatically used source, source name does not match for the join to work using source_name



# recodes old agency numbers to consistent agency number
rev_temp <- rev_temp %>% 
  mutate(agency = case_when(
    (agency=="438"| agency=="475" |agency == "505") ~ "440",
    # financial institution &  professional regulation &
     # banks and real estate  --> coded as  financial and professional reg
    agency == "473" ~ "588", # nuclear safety moved into IEMA
    (agency =="531" | agency =="577") ~ "532", # coded as EPA
    (agency =="556" | agency == "538") ~ "406", # coded as agriculture
    agency == "560" ~ "592", # IL finance authority (fire trucks and agriculture stuff)to state fire marshal
    agency == "570" & fund == "0011" ~ "494",   # city of Chicago road fund to transportation
    TRUE ~ (as.character(agency)))) 
```

# Expenditure & Revenue Categorization

## Modify Expenditure File

__Tax refunds__ 

Aggregate expenditures: Save tax refunds as negative revenue. Code refunds to match the rev_type codes (02=income taxes, 03 = corporate income taxes, 06=sales tax, 09=motor fuel tax, 24=insurance taxes and fees, 35 = all other tax refunds).


```{r tax-refunds}
## negative revenue becomes tax refunds

tax_refund_long <- exp_temp %>%           # fund != "0401" # removes State Trust Funds
  filter(fund != "0401" & (object=="9910"|object=="9921"|object=="9923"|object=="9925")) %>%
  # keeps these objects which represent revenue, insurance, treasurer,and financial and professional reg tax refunds
  mutate(refund = case_when(
    fund=="0278" & sequence == "00" ~ "02", # for income tax refund
    fund=="0278" & sequence == "01" ~ "03", # tax administration and enforcement and tax operations become corporate income tax refund
     fund == "0278" & sequence == "02" ~ "02",
    object=="9921" ~ "21",                # inheritance tax and estate tax refund appropriation
    object=="9923" ~ "09",                # motor fuel tax refunds
    obj_seq_type == "99250055" ~ "06",    # sales tax refund
    fund=="0378" & object=="9925" ~ "24", # insurance privilege tax refund
    fund=="0001" & object=="9925" ~ "35", # all other taxes
      T ~ "CHECK"))                       # if none of the items above apply to the observations, then code them as CHECK 

    
exp_temp <- left_join(exp_temp, tax_refund_long) %>%
  mutate(refund = ifelse(is.na(refund),"not refund", as.character(refund)))

tax_refund <- tax_refund_long %>% 
  group_by(refund, fy)%>%
  summarize(refund_amount = sum(expenditure, na.rm = TRUE)/1000000) %>%
  pivot_wider(names_from = refund, values_from = refund_amount, names_prefix = "ref_") %>%
  mutate_all(~replace_na(.,0)) %>%
  arrange(fy)

tax_refund %>% pivot_longer( ref_02:ref_35, names_to = "Refund Type", values_to = "Amount") %>%
  ggplot()+
  geom_line(aes(x=fy,y=Amount, group = `Refund Type`, color = `Refund Type`))+
  labs(title = "Refund Types", caption = "Refunds are excluded from Expenditure totals and instead subtracted from Revenue totals") + 
  labs(title = "Tax refunds", 
       caption = "Rev_type codes: 02=income taxes, 03=corporate income taxes, 06=sales tax, 09=motor fuel tax, 
       24=insurance taxes and fees, 35 = all other tax refunds." )

# remove the items we recoded in tax_refund_long
exp_temp <- exp_temp %>% filter(refund == "not refund")


```

`tax_refund` amounts are removed from expenditure totals and subtracted from revenue totals (since they were tax refunds).


__Pension Expenditures__

__For yearly expenditure calculations, the state contributions to the pension funds (object = 4431) should be the expenditure included for pensions. If trying to look at the bigger fiscal health picture and include unfunded liabilities and in/out flows, then items like purchase of investments and large spikes that occurred from policy changes should be included. Again, State contributions TO the pension funds are the expenditures BUT an additional graph/discussion on the employer contributions, employee contributions (revenue source), and benefits paid out to employees (object = 4430) should be included and considered for additional context on Illinois' situation.__

State payments to the following pension systems:  

• Teachers Retirement System (TRS)  
- New POB bond in 2019: Accelerated Bond Fund paid benefits in advance as lump sum   
• State Employee Retirement System (SERS)  
• State University Retirement System (SURS)  
• Judges Retirement System (JRS)  
• General Assembly Retirement System (GARS)  

- Includes pension stabilization fund = 0319, object = 1900 and the $300 million investment in FY2022.    
- State pension contributions are largely captured with object=4431. **(State payments into pension fund)**    
  - includes 8 billion payment in 2004 that creates large peak in expenditure graphs   
  - does not capture recent pension stabilization payments   
- Some expenditures with object=4430 are paid for with Pension obligation bond funds (fund == 0825). In past years, some POB funded expenditures were moved to revenue side in the Stata code. We are no longer doing this as of FY2021.    

Modify exp_temp and move all state pension contributions to their own group (901):
```{r pensions}
exp_temp <-  exp_temp %>% 
  arrange(fund) %>%
  mutate(pension = case_when( 
   (object=="4431") ~ 1, # 4431 = easy to find pension payments INTO fund
   
 # (object>"1159" & object<"1166") & fund != "0183" & fund != "0193"   ~ 2, 
   # objects 1159 to 1166 are all considered Retirement by Comptroller, 
  # Excluded - employer contributions from agencies/organizations/etc.

  (object=="1298" &  # Purchase of Investments, Normally excluded
     (fy==2010 | fy==2011) & 
     (fund=="0477" | fund=="0479" | fund=="0481")) ~ 3, #judges retirement OUT of fund
  # state borrowed money from pension funds to pay for core services during 2010 and 2011. 
  # used to fill budget gap and push problems to the future. 
 

 fund == "0319" ~ 4, # pension stabilization fund
                                        TRUE ~ 0) )

table(exp_temp$pension) 



exp_temp %>% filter(pension != 0) %>%
  mutate(pension = as.factor(pension))%>%
  group_by(fy, pension) %>% 
  summarize(expenditure = sum(expenditure, na.rm = TRUE)) %>%
  ggplot(aes(x=fy, y = expenditure, group=pension)) + 
  geom_line(aes(color = pension)) + 

  labs (title = "Pension expenditures", 
  caption = "1 = State contributions INTO pension funds")+
    theme(legend.position = "bottom")


```



```{r pensions-POB}

# special accounting of pension obligation bond (POB)-funded contributions to JRS, SERS, GARS, TRS 

exp_temp <- exp_temp %>% 
  # change object for 2010 and 2011, retirement expenditures were bond proceeds and would have been excluded
  mutate(object = ifelse((pension >0 & in_ff == "0"), "4431", object)) %>% 
  # changes weird teacher & judge retirement system  pensions object to normal pension object 4431
  mutate(pension =  ifelse(pension >0 & in_ff == "0", 6, pension)) %>% # coded as 6 if it was supposed to be excluded. 
  mutate(in_ff = ifelse(pension>0, "1", in_ff))

table(exp_temp$pension) 


# all other pensions objects  codes get agency code 901 for State Pension Contributions
exp_temp <- exp_temp %>% 
  mutate(agency = ifelse(pension>0, "901", as.character(agency)),
         agency_name = ifelse(agency == "901", "State Pension Contributions", as.character(agency_name)))

exp_temp %>% 
 filter(pension > 0) %>%  
  mutate(pension = as.factor(pension)) %>%
  group_by(fy, pension) %>% 
  summarize(expenditure = sum(expenditure, na.rm=TRUE)) %>%
  ggplot(aes(x=fy, y=expenditure, color = pension)) +
  geom_line() + 
  labs (title = "Pension Expenditures", 
  caption = "")

exp_temp %>% 
 filter(pension > 0) %>%  
  group_by(fy) %>% 
  summarize(expenditure = sum(expenditure, na.rm=TRUE)) %>%
  ggplot(aes(x=fy, y=expenditure)) +
  geom_line() + 
  labs (title = "Pension Expenditures")
```



__Drop Interfund transfers__

- object == 1993 is for interfund cash transfers  
- agency == 799 is for statutory transfers  
- object == 1298 is for purchase of investments and is not spending EXCEPT for costs in 2010 and 2011 (and were recoded already to object == "4431"). Over 168,000 observations remain.  

```{r drop-transfers}

transfers_drop <- exp_temp %>% filter(
  agency == "799" | # statutory transfers
           object == "1993" |  # interfund cash transfers
           object == "1298") # purchase of investments

exp_temp <- anti_join(exp_temp, transfers_drop)
exp_temp

transfers_drop %>% filter(fy>2019 & object == 1993) %>% group_by(obj_seq_type) %>% summarize(sum = sum(expenditure)) %>% arrange(-sum)
```


__State employee healthcare costs__


State Employee Health Care = Sum of expenditures for "health care coverage as elected by members per state employees group insurance act." The payments are made from the Health Insurance Reserve Fund. We subtract the share that came from employee contributions. Employee contributions are not considered a revenue source or an expenditure in our analysis.

In FY2013, the Local Government Health Insurance fund was transferred to the department of Central Management Services (agency changes from 478 to 416 in data.)

Employer group insurance contributions for health insurance are excluded to avoid double counting the cost of healthcare. All employer group insurance contributions are coded as object = 1180. BUT the last two fiscal years were coded as 1900 instead of 1180 for lump sums instead of employer contributions    

- anything to do with pandemic federal money categorization? Was a lump sum or transfer used to pay the group contribution and coded differently?   

- This group insurance expenditure coded as 1900 instead of 1180 caused one of the major problems when calculating the fiscal gap for FY21. Stata code originally failed to drop group insurance expenditure in FY21 (since object was = 1900 instead of 1180) and resulted in expenditures being around $2 billion higher than they should have been for Central Management expenditures. This was noticed by GOMB (Alexis Sturm) and resulted in looking deeper into what items are going in and out of health care funds (See [State Employee Healthcare Discussion]).   



Relevant funds:   
Fund = 0457 is "Group insurance premium", in_ff = 1  
Fund = 0193 is "Local govt health insurance reserve", in=ff = 0  
fund = 0477 is "Community College Health Insurance", in=ff = 0.     
- had large amount in early years    
Fund = 0907 = health insurance reserve, in_ff = 1  
Fund = 9939 is "group self-insurers' insolv", in_ff = 1  
Fund = 0940 is Self-Insurers security, in_ff = 0  
Fund = 0739 is Group Workers Comp Pool Insol, in_ff = 1   


- eehc = 0 means it is NOT a state healthcare cost but it is an employer contribution of some type to some fund   
- eehc = 1 means it is a state employee healthcare cost and it is an employer contribution to health insurance   


Of the many different ways I have coded the healthcare costs, most do not have expenditures during 2010, 2011, 2016, 2017.

```{r include=FALSE}
exp_temp_check <- exp_temp %>% 
  mutate(agency = case_when(   # turns specific items into State Employee Healthcare (agency=904)
      fund=="0907" & (agency=="416" & appr_org=="20") ~ "904",   # central management Bureau of benefits using health insurance reserve 
      fund=="0907" & (agency=="478" & appr_org=="80") ~ "904",   # agency = 478: healthcare & family services using health insurance reserve - stopped using this in 2012
    #  fund=="0001" & appr_org=="20" & object=="1900" & agency=="416" & (fy>2002 & fy<2006) ~ "904",
      
    #  fund=="0001" & appr_org=="20" & object=="1900" & agency=="416" & (fy>2020) ~ "904",
      
    #  obj_seq_type == "19000000" & wh_approp_name == "GROUP INSURANCE" & (fy>2020) ~ "904",

      TRUE ~ as.character(agency))) %>%
  
  mutate(agency_name = ifelse(agency == "904", "STATE EMPLOYEE HEALTHCARE", as.character(agency_name)),
         group = ifelse(agency == "904", "904", as.character(agency))) %>% # creates group variable
filter(group == "904") %>% group_by(fy) %>% summarise(healthcare_cost = sum(expenditure))

exp_temp_check

# Looks good, Sept 28 AWM
```

If observation is a group insurance contribution, then the expenditure amount is set to $0 (essentially dropped from analysis).

```{r eehc1}
#if observation is a group insurance contribution, then the expenditure amount is set to $0 (essentially dropped from analysis)

# pretend eehc is named group_insurance_contribution or something like that
# eehc coded as zero implies that it is group insurance
# if eehc=0, then expenditures are coded as zero for group insurance to avoid double counting costs

exp_temp <- exp_temp %>% 
  mutate(eehc = ifelse(
    # group insurance contributions for 1998-2005 and 2013-present
   fund == "0001" & (object == "1180" | object =="1900") & agency == "416" & appr_org=="20", 0, 1) )%>% 
  mutate(eehc = ifelse(
    # group insurance contributions for 2006-2012
    fund == "0001" & object == "1180" & agency == "478" & appr_org=="80", 0, eehc) )%>%
     # group insurance contributions from road fund
  # coded with 1900 for some reason??
    mutate(eehc = ifelse(
      fund == "0011" & object == "1900" & agency == "416" & appr_org=="20", 0, eehc) ) %>%
  
  mutate(expenditure = ifelse(eehc=="0", 0, expenditure)) %>%
  
  mutate(agency = case_when(   # turns specific items into State Employee Healthcare (agency=904)
      fund=="0907" & (agency=="416" & appr_org=="20") ~ "904",   # central management Bureau of benefits using health insurance reserve 
      fund=="0907" & (agency=="478" & appr_org=="80") ~ "904",   # agency = 478: healthcare & family services using health insurance reserve - stopped using this in 2012
      TRUE ~ as.character(agency))) %>%
  mutate(agency_name = ifelse(agency == "904", "STATE EMPLOYEE HEALTHCARE", as.character(agency_name)),
         in_ff = ifelse( agency == "904", 1, in_ff),
         group = ifelse(agency == "904", "904", as.character(agency)))  
# creates group variable

# Default group = agency number

healthcare_costs <- exp_temp %>% filter(group == "904")

healthcare_costs

exp_temp %>% filter(group == "904") %>% group_by(fy) %>% 
  summarise(healthcare_cost = sum(expenditure, na.rm = TRUE)) %>% 
  ggplot() +geom_line(aes(x=fy, y=healthcare_cost)) + labs(title="State Employee Healthcare Costs - Included in Fiscal Futures Model", caption = "Fund 0907 for agencies responsible for health insurance reserve (DHFS & CMS)")


#exp_temp <- anti_join(exp_temp, healthcare_costs) %>% mutate(expenditure = ifelse(object == "1180", 0, expenditure))

#healthcare_costs_yearly <- healthcare_costs %>% group_by(fy, group) %>% summarise(healthcare_cost = sum(expenditure, na.rm = TRUE)/1000000) %>% select(-group)


```

~~This code chunk above for dealing with group insurance means that healthcare costs need to be added to expenditures after other group names are assigned. Then employee contributions/insurance premiums from the revenue side need to be subtracted from the total cost of employee healthcare for the net cost.~~ Do not do this. 


__Local Transfers__

Separate transfers to local from parent agencies that come from DOR(492) or Transportation (494). Treats muni revenue transfers as expenditures, not negative revenue.

The share of certain taxes levied state-wide at a common rate and then transferred to local governments. (Purely local-option taxes levied by specific local governments with the state acting as collection agent are NOT included.)

The six corresponding revenue items are:  

• Local share of Personal Income Tax
  - Individual Income Tax Pass-Through New 2021 (source 2582).
• Local share of General Sales Tax   
• Personal Property Replacement Tax on Business Income  
• Personal Property Replacement Tax on Public Utilities  

• Local share of Motor Fuel Tax 
  - Transportation Renewal Fund 0952  
  

Until Dec 18. 2022, Local CURE was being aggregated into Revenue totals since the agency was the Department of Revenue. However the \$371 million expenditure is for "LOC GOVT ARPA" and the revenue source that is Local CURE is also $371 million. Since it cancels out and is just passed through the state government, I am changing changing the fund_ab_in file so that in_ff=0 for the Local CURE fund. It also inflates the department of revenue expenditures in a misleading way when the expense is actually a transfer to local governments. 


- Dropping Local CURE fund from analysis results in a \$371 million decrease in the department of Revenue (where the Local Government ARPA transfer money). The appropriation for it was over $740 million so some will probably be rolled over to FY23 too.   
- In the FY21 New and Reused Funds word document, 0325 Local CURE is described as _"Created as a federal trust fund. The fund is established to receive transfers from either the disaster response and recovery fund or the state cure fund of federal funds received by the state. These transfers, subject to appropriation, will provide for the administration and payment of grants and expense reimbursements to units of local government. Revenues should be under Federal Other and expenditures under Commerce and Economic Opportunity."_ - I propose changing it to exclude for both (in_ff=0). 


```{r transfers-to-local}
exp_temp <- exp_temp %>% mutate(
  agency = case_when(fund=="0515" & object=="4470" & type=="08" ~ "971", # income tax to local governments
                     fund=="0515" & object=="4491" & type=="08" & sequence=="00" ~ "971", # object is shared revenue payments
                     fund=="0802" & object=="4491" ~ "972", #pprt transfer
                     fund=="0515" & object=="4491" & type=="08" & sequence=="01" ~ "976", #gst to local
                     fund=="0627" & object=="4472"~ "976" , # public transportation fund but no observations exist
                     fund=="0648" & object=="4472" ~ "976", # downstate public transportation, but doesn't exist
                     fund=="0515" & object=="4470" & type=="00" ~ "976", # object 4470 is grants to local governments
                    object=="4491" & (fund=="0188"|fund=="0189") ~ "976",
                     fund=="0187" & object=="4470" ~ "976",
                     fund=="0186" & object=="4470" ~ "976",
                    object=="4491" & (fund=="0413"|fund=="0414"|fund=="0415")  ~ "975", #mft to local
                  fund == "0952"~ "975", # Added Sept 29 2022 AWM. Transportation Renewal MFT
                    TRUE ~ as.character(agency)),
  
  agency_name = case_when(agency == "971"~ "INCOME TAX 1/10 TO LOCAL",
                          agency == "972" ~ "PPRT TRANSFER TO LOCAL",
                          agency == "975" ~ "MFT TO LOCAL",
                          agency == "976" ~ "GST TO LOCAL",
                          TRUE~as.character(agency_name)),
  group = ifelse(agency>"970" & agency < "977", as.character(agency), as.character(group)))

```




```{r drop-local-transfers}
transfers_long <- exp_temp %>% 
  filter((group == "971" |group == "972" | group == "975" | group == "976"))# | fund == "0325")

transfers_long %>% 
  group_by(agency_name, fy) %>% 
  summarize(expenditure = sum(expenditure, na.rm=TRUE) )%>% 
  ggplot() + geom_line(aes(x=fy, y = expenditure, color=agency_name)) + labs(title = "Transfers to Local Governments", caption = "Data Source: Illinois Office of the Comptroller")

transfers <- transfers_long %>%
  group_by(fy, group ) %>%
  summarize(sum_expenditure = sum(expenditure)/1000000) %>%
  pivot_wider(names_from = "group", values_from = "sum_expenditure", names_prefix = "exp_" )

exp_temp <- anti_join(exp_temp, transfers_long)


dropped_inff_0 <- exp_temp %>% filter(in_ff == 0)

exp_temp <- exp_temp %>% filter(in_ff == 1) # drops in_ff = 0 funds AFTER dealing with net-revenue above

```


The Local Transfers from the Personal Property Replacement Tax (fund 802) increased over $2 billion from corporate income taxes alone. Personal property replacement taxes (PPRT) are revenues collected by the state of Illinois and paid to local governments to replace money that was lost by local governments when their powers to impose personal property taxes on corporations, partnerships, and other business entities were taken away.


__Debt Service__


Debt Service expenditures include interest payment on both short-term and long-term debt. We do not include escrow or principal payments. 

**Decision from Sept 30 2022:** We are no longer including short term principal payments as a cost; only interest on borrowing is a cost. Pre FY22 and the FY21 correction, we did include an escrow payment and principle payments as costs but not bond proceeds as revenues. This caused expenditures to be inflated because we were essentially counting debt twice - the principle payment and whatever the money was spent on in other expenditure categories, which was incorrect. 

```{r debt-service}
debt_drop <- exp_temp %>% 
  filter(object == "8841" |  object == "8811")  
# escrow  OR  principle

#debt_drop %>% group_by(fy) %>% summarize(sum = sum(expenditure)) %>% arrange(-fy)


debt_keep <- exp_temp %>% 
  filter(fund != "0455" & (object == "8813" | object == "8800" )) 
# examine the debt costs we want to include

#debt_keep %>% group_by(fy) %>% summarize(sum = sum(expenditure)) %>% arrange(-fy) 


exp_temp <- anti_join(exp_temp, debt_drop) 
exp_temp <- anti_join(exp_temp, debt_keep)

debt_keep <- debt_keep %>%
  mutate(
    agency = ifelse(fund != "0455" & (object == "8813" | object == "8800"), "903", as.character(agency)),
    group = ifelse(fund != "0455" & (object == "8813" | object == "8800"), "903", as.character(group)),
    in_ff = ifelse(group == "903", 1, as.character(in_ff)))

debt_keep_yearly <- debt_keep %>% group_by(fy, group) %>% summarize(debt_cost = sum(expenditure,na.rm=TRUE)/1000000) %>% select(-group)

```


__Medicaid__ 

That portion of the Healthcare and Family Services (or Public Aid in earlier years, agency code 478) budget for Medical (appr_organization code 65) for awards and grants (object codes 4400 and 4900).  

Jan 2015 FF Documentation - "Typically, presentations of the Illinois budget—for example, the Illinois Office of the Comptroller’s Fiscal Focus reports—use "medical assistance payments" made by the Department of Healthcare and Family Services (DHFS) as a proxy for Medicaid expenditures but these numbers include both Medicaid and non-Medicaid expenditures such as state-funded medical programs (e.g. AllKids). Medicaid expenditures are not limited to medical assistance in DHFS - the program also covers clients of home and community-based care programs that are administered by the Department of Human Services and Department of Aging. Because Medicaid is an entitlement,
it is a reimbursement for medical care for qualifying individuals. Medical bills may be paid from designated funds by any of several agencies that have spending authority from those funds, as they provide health care services to individuals who qualify for Medicaid or another form of assistance. So, the numbers for medical assistance payments from DHFS may both over-state and underestimate the costs (and growth rate) of the actual Medicaid program. This is a legitimate concern, because the expenditure numbers for Illinois’ medical assistance payments do not correspond to the numbers disseminated by national Medicaid data sources"

- 2015 Documentation notes that federal cash receipts for Medicaid are used to estimate Medicaid expenditures from the General Revenue Fund (GRF) by dividing the cash receipts by the match rate. Medicaid expenditures from specially-designated funds are assumed to be the warrants issued from those funds. Then the two components—GRF and special funds—are added together for a total number. (Note that the number for FF Medicaid expenditures includes the state portion and the federal portion combined. Past code uses sequence variable also to identify CHIP and GRF (sequence = 20, 54, 61, 62, or 65).
 


- Department of Healthcare and Family Services administers medical assistance programs (e.g. Medicaid, CHIP, All Kids). The Fiscal Futures model moves Medicaid and CHIP expenditures to their own expenditure category. State funded programs remain in DHFS.


> Side note: State CURE healthcare provider relief will remain in the Medicaid expenditure category due to the nature of it being federal funds providing public health services and funding to locations that provide public services.

- Uses same appropriation name of "HEALTHCARE PROVIDER RELIEF" and fund == 0793 and obj_seq_type == 49000000. So can defend including healthcare provider relief as Medicaid expenditure. 




```{r Medicaid-check, include = FALSE, eval=FALSE}
medicaid_check <- exp_temp %>% filter(agency=="478" & (appr_org=="01" | appr_org == "65" | appr_org=="88") & (object=="4900" | object=="4400"))
# This includes the State CURE fund

medicaid_check %>% group_by(fy) %>% summarize(sum = sum(expenditure)) %>% arrange(-fy)  # looks good I think AWM

# if we want state CURE as public health or other group number
# exp_temp <- exp_temp %>% mutate(group = if_else(fund == "0324" & agency=="478" & appr_org == "65" & object=="4900", "478", as.character(group)))

medicaid_check %>% filter(fy>2020) %>% group_by(wh_approp_name,fy) %>% summarize(sum=sum(expenditure)) %>% arrange(wh_approp_name)
```

Federal Medical Assistance Program (FMAP): in 1965. The FMAP formula compares the state per-capita income to the national per-capita income. There is no cap on the dollar amount that the federal government pays, so the more that a state spends the more that it receives. a maximum of 83%. States with a higher per-capita income receive lower FMAP funding but no less than 50%, and the states that have a lower per-capita income receive higher FMAP funding. Those that need more, get more. 



### Add Other Fiscal Future group codes

```{r group-codes}


exp_temp <- exp_temp %>%
  #mutate(agency = as.numeric(agency) ) %>%
  # arrange(agency)%>%
  mutate(
    group = case_when(
      agency>"100"& agency<"200" ~ "910", # legislative
      
      agency == "528"  | (agency>"200" & agency<"300") ~ "920", # judicial
      pension>0  ~ "901", # pensions
      (agency>"309" & agency<"400") ~ "930",    # elected officers
      
      agency == "586" ~ "959", # create new K-12 group

      agency=="402" | agency=="418" | agency=="478" | agency=="444" | agency=="482" ~ as.character(agency), # aging, Children and Family Services (DCFS), Healthcare and Family Services (DHFS), human services (DHS), public health (DPH)
      T ~ as.character(group))
    ) %>%      

  
  mutate(group = case_when(
    agency=="478" & (appr_org=="01" | appr_org == "65" | appr_org=="88") & (object=="4900" | object=="4400") ~ "945", # separates CHIP from health and human services and saves it as Medicaid
    
    agency == "586" & fund == "0355" ~ "945",  # 586 (Board of Edu) has special education which is part of medicaid
    
    # OLD CODE: agency == "586" & appr_org == "18" ~ "945", # Spec. Edu Medicaid Matching
    
    agency=="425" | agency=="466" | agency=="546" | agency=="569" | agency=="578" | agency=="583" | agency=="591" | agency=="592" | agency=="493" | agency=="588" ~ "941", # public safety & Corrections
    
     agency=="406" | agency=="420" | agency=="494" |  agency=="557" ~ as.character(agency), # econ devt & infra, tollway
    
    agency=="511" | agency=="554" | agency=="574" | agency=="598" ~ "946",  # Capital improvement
    
    agency=="422" | agency=="532" ~ as.character(agency), # environment & nat. resources
    
    agency=="440" | agency=="446" | agency=="524" | agency=="563"  ~ "944", # business regulation
    
    agency=="492" ~ "492", # revenue
    
    agency == "416" ~ "416", # central management services
    agency=="448" & fy > 2016 ~ "416", #add DoIT to central management 
    
    T ~ as.character(group))) %>%
  
  
  mutate(group = case_when(
    # agency=="684" | agency=="691"  ~ as.character(agency), # moved under higher education in next line. 11/28/2022 AWM
    
    agency=="692" | agency=="695" | agency == "684" |agency == "691" | (agency>"599" & agency<"677") ~ "960", # higher education
    
    agency=="427"  ~ as.character(agency), # employment security
    
    agency=="507"|  agency=="442" | agency=="445" | agency=="452" |agency=="458" | agency=="497" ~ "948", # other departments
    
    # other boards & Commissions
    agency=="503" | agency=="509" | agency=="510" | agency=="565" |agency=="517" | agency=="525" | agency=="526" | agency=="529" | agency=="537" | agency=="541" | agency=="542" | agency=="548" |  agency=="555" | agency=="558" | agency=="559" | agency=="562" | agency=="564" | agency=="568" | agency=="579" | agency=="580" | agency=="587" | agency=="590" | agency=="527" | agency=="585" | agency=="567" | agency=="571" | agency=="575" | agency=="540" | agency=="576" | agency=="564" | agency=="534" | agency=="520" | agency=="506" | agency == "533" ~ "949", 
    
    # non-pension expenditures of retirement funds moved to "Other Departments"
    # should have removed pension expenditures already from exp_temp in Pensions step above
    agency=="131" | agency=="275" | agency=="589" |agency=="593"|agency=="594"|agency=="693" ~ "948",
    
    T ~ as.character(group))) %>%

  mutate(group_name = 
           case_when(
             group == "416" ~ "Central Management",
             group == "478" ~ "Healthcare and Family Services",
             group == "482" ~ "Public Health",
             group == "900" ~ "NOT IN FRAME",
             group == "901" ~ "STATE PENSION CONTRIBUTION",
             group == "903" ~ "DEBT SERVICE",
             group == "910" ~ "LEGISLATIVE"  ,
             group == "920" ~ "JUDICIAL" ,
             group == "930" ~ "ELECTED OFFICERS" , 
             group == "940" ~ "OTHER HEALTH-RELATED", 
             group == "941" ~ "PUBLIC SAFETY" ,
             group == "942" ~ "ECON DEVT & INFRASTRUCTURE" ,
             group == "943" ~ "CENTRAL SERVICES",
             group == "944" ~ "BUS & PROFESSION REGULATION" ,
             group == "945" ~ "MEDICAID" ,
             group == "946" ~ "CAPITAL IMPROVEMENT" , 
             group == "948" ~ "OTHER DEPARTMENTS" ,
             group == "949" ~ "OTHER BOARDS & COMMISSIONS" ,
             group == "959" ~ "K-12 EDUCATION" ,
             group == "960" ~ "UNIVERSITY EDUCATION" ,
             group == agency ~ as.character(group),
             TRUE ~ "Check name"),
         year = fy)

exp_temp %>% filter(group_name == "Check name")

#write_csv(exp_temp, "all_expenditures_recoded.csv")

```


All expenditures recoded but not aggregated: Allows for inspection of individual expenditures within larger categories. This stage of the data is extremely useful for investigating almost all questions we have about the data. 

Note that these are the raw figures BEFORE we take the additional steps:   

- Subtract tax refunds from tax revenues by revenue type.  

```{r}
exp_temp %>% filter(fy>2020 & fund == "0561") %>% group_by(wh_approp_name, fy) %>% summarize(sum=sum(expenditure)) %>% arrange(-sum)


```

SBE Federal Department of Education is fund 0561. Fund 0579 is the State Board of Education. 

## Modify  Revenue data


Revenue Categories NOT included in Fiscal Futures:   
- 32.  Garnishment-Levies.  (State is fiduciary, not beneficiary.)  
- 45.  Student Fees-Universities.  (Excluded from state-level budget.)  
- 51.  Retirement Contributions (of individuals and non-state entities).    
- 66.  Proceeds, Investment Maturities.  (Not sustainable flow.)  
- 72.  Bond Issue Proceeds.  (Not sustainable flow.)     
- 75.  Inter-Agency Receipts.  
- 79.  Cook County Intergovernmental Transfers.  (State is not beneficiary.)  
- 98.  Prior Year Refunds.   
- 99.  Statutory Transfers.   

**All Other Sources** 

Expanded to include the following smaller sources:  
- 30.  Horse Racing Taxes & Fees.  
- 60.  Other Grants and Contracts.  
- 63.  Investment Income.    


For aggregating revenue, use the rev_1998_2022 dataframe, join the funds_ab_in_2022 file to it, and then join the ioc_source_type file to the dataset. Remember: You need to update the funds_ab_in and ioc_source_type file every year! 

```{r, warning = FALSE, message=FALSE}
# fund info to revenue for all years
rev_temp <- inner_join(rev_1998_2022, funds_ab_in_2022, by = "fund") %>% arrange(source)

# need to update the ioc_source_type file every year! 
ioc_source_type <- readxl::read_xlsx("C:/Users/aleaw/OneDrive/Documents/PhD Fall 2021 - Spring 2022/Merriman RA/Fiscal Futures FY2022/Replication-Files/ioc_source_updated22_AWM.xlsx")

rev_temp <- left_join(rev_temp, ioc_source_type, by = "source")
# automatically used source, source name does not match for the join to work using source_name



# recodes old agency numbers to consistent agency number
rev_temp <- rev_temp %>% 
  mutate(agency = case_when(
    (agency=="438"| agency=="475" |agency == "505") ~ "440",
    # financial institution &  professional regulation &
     # banks and real estate  --> coded as  financial and professional reg
    agency == "473" ~ "588", # nuclear safety moved into IEMA
    (agency =="531" | agency =="577") ~ "532", # coded as EPA
    (agency =="556" | agency == "538") ~ "406", # coded as agriculture
    agency == "560" ~ "592", # IL finance authority (fire trucks and agriculture stuff)to state fire marshal
    agency == "570" & fund == "0011" ~ "494",   # city of Chicago road fund to transportation
    TRUE ~ (as.character(agency)))) 
```


#### Federal Revenue

The Fiscal Futures Model divides federal funds (IOC revenue type = 57) into Medicaid, Transportation, and All Other Federal Funds. These can include: Health and Human Services Grants, Federal Stimulus Package, Department of Education Grants, Department of Transportation Grants, Department of Agriculture Grants, TANF Grants, and Department of Labor Grants.

Federal Medicaid: DHFS receives money for Medicaid that is deposited into the General Revenue Fund. There are also Special State Funds used for Medicaid that receive specific revenues (e.g. Healthcare Provider Taxes) which are matched with Federal Funds. The federal receipts in these special funds are aggregated and added to the federal receipts in the GRF that are received by DHFS. 

Sources:

- 618 = Health and Human Services (not used)  
- 660 = HHS/Hospital Participation   
- 676 = Medical Assistance  
- 692 = Medical Assistance  
- 1552 = DHHS/ FFP-Medicaid Rehab Option
- 2306 = Enhanced Fed Fin PART-ARRA 
- 2076 = IDPH-HHS/CMS   
- 2364 = Department of Insurance (not used)

Other potential medicaid sources from CTRL-Fing "med" in revenue sources:   
Sources: 2104 = Medicare Part D & 675 = Medical Administration


Transportation: If Agency is 494 and considered Federal Revenue, then it is recoded to its own category of "Federal Transportation".

```{r create-rev-federal-transfers}
#rev_temp <- rev_temp %>% filter(in_ff==1)

rev_temp <- rev_temp %>% 
  mutate(
    rev_type = ifelse(rev_type=="57" & agency=="478" & (source=="0618"|source=="2364"|source=="0660"|source=="1552"| source=="2306"| source=="2076"|source=="0676"|source=="0692"), "58", rev_type),
    rev_type_name = ifelse(rev_type=="58", "Federal Medicaid Reimbursements", rev_type_name),
    rev_type = ifelse(rev_type=="57" & agency=="494", "59", rev_type),
    rev_type_name = ifelse(rev_type=="59", "Federal Transportation", rev_type_name),
    rev_type_name = ifelse(rev_type=="57", "Federal - Other", rev_type_name),
    rev_type = ifelse(rev_type=="6", "06", rev_type),
    rev_type = ifelse(rev_type=="9", "09", rev_type)) 

rev_temp %>% 
  filter(rev_type == "58" | rev_type == "59" | rev_type == "57") %>% 
  group_by(fy, rev_type, rev_type_name) %>% 
  summarise(receipts = sum(receipts, na.rm = TRUE)/1000000) %>% 
  ggplot() +
  geom_line(aes(x=fy, y=receipts,color=rev_type_name)) +
      theme_bw() +
  scale_y_continuous(labels = comma)+
  labs(title = "Federal to State Transfers", 
       y = "Millions of Dollars", x = "") + 
  theme(legend.position = "bottom", legend.title = element_blank()  )
```


Looking at Total Federal Revenue over the years:
```{r}
fedrev<- rev_temp %>% 
  filter(rev_type == "58" | rev_type == "59" | rev_type == "57") 

fedrev %>% 
  group_by(fy) %>% 
  summarise(receipts = sum(receipts, na.rm = TRUE)/1000000) %>% 
  ggplot() +
  geom_line(aes(x=fy, y=receipts)) +
      theme_bw() +
  scale_y_continuous(labels = comma)+
  labs(title = "All Federal Revenue", 
       y = "Millions of Dollars", x = "") + 
  theme(legend.position = "bottom", legend.title = element_blank()  )+
    scale_y_continuous(limits = c(0,45000))


fedrev %>% 
  filter(source_name_AWM != "FEDERAL STIMULUS PACKAGE") %>%
  group_by(fy) %>% 
  summarise(receipts = sum(receipts, na.rm = TRUE)/1000000) %>% 
  ggplot() +
  geom_line(aes(x=fy, y=receipts)) +
      theme_bw() +
  scale_y_continuous(labels = comma)+
  labs(title = "All Federal EXCEPT Federal Stimulus Package", 
       y = "Millions of Dollars", x = "",
       caption = "Note: Dropping Federal Stimulus Package revenue only removes the $3.5 billion from FY20, $0.3 billion from FY21, and 
       $8.5 billion from FY22. There is still over $12 billion more in Federal Revenue compared to FY19.") + 
  theme(legend.position = "bottom", legend.title = element_blank()  ) +
  scale_y_continuous(limits = c(0,45000))
```

Dropping Federal Stimulus Package from the revenue sources removes 3.5 billion in revenue from FY20 and nearly 8.5 billion from FY22. It also changes the revenue following the Great Recession in FY2009.

**Dropping State CURE Revenue**

The Fiscal Futures model focuses on sustainable revenue sources. To understand our fiscal gap and outlook, we need to exclude these one time revenues. GOMB has emphasized that they have allocated COVID dollars to one time expenditures (unemployment trust fund, budget stabilization fund, etc.). The fiscal gap, graphs,and CAGRs have been recalculated in the [Drop COVID Dollars] section below. In addition, an attempt at forecasting revenue and expenditures is also made after dropping the federal COVID dollars.

> NOTE: I have only dropped State and Local CURE revenue so far. Federal money went into other funds during the beginning of pandemic.

```{r}
# only labels observations where the source was specifically named "federal stimulous package"
# does not code other COVID legislative funds that went under other names and had more specific purposes.
rev_temp <- rev_temp %>% 
  mutate(covid_dollars = ifelse(source_name_AWM == "FEDERAL STIMULUS PACKAGE",1,0))
```


#### Health Insurance Premiums from Employees


Insurance premiums for employees is coded below but it is NOT used in the fiscal futures model. Employee and employer premiums are considered rev_51 and dropped from analysis in later step. 


0120 = ins prem-option life   
0120 = ins prem-optional life/univ   

0347 = optional health - HMO  
0348 = optional health - dental   
0349 = optional health - univ/local SI  
0350 = optional health - univ/local  
0351 = optional health - retirement  
0352 = optional health - retirement SI  
0353 = optional health - retire/dental   
0354 = optional health - retirement hmo    

2199-2209 = various HMOs, dental, health plans from Health Insurance Reserve (fund)

```{r step-3.6-insurance-premiums}

#collect optional insurance premiums to fund 0907 for use in eehc expenditure  
rev_temp <- rev_temp %>% 
  mutate(
    #variable not used in aggregates, but could be interesting for other purposes
    employee_premiums = ifelse(fund=="0907" & (source=="0120"| source=="0121"| (source>"0345" & source<"0357")|(source>"2199" & source<"2209")), 1, 0),
    
    # adds more rev_type codes
    rev_type = case_when(
      fund =="0427" ~ "12", # pub utility tax
      fund == "0742" | fund == "0473" ~ "24", # insurance and fees
      fund == "0976" ~ "36",# receipts from rev producing
      fund == "0392" |fund == "0723" ~ "39", # licenses and fees
      fund == "0656" ~ "78", #all other rev sources
      TRUE ~ as.character(rev_type)))
# if not mentioned, then rev_type as it was



# # optional insurance premiums = employee insurance premiums

# emp_premium <- rev_temp %>%
#   group_by(fy, employee_premiums) %>%
#   summarize(employee_premiums_sum = sum(receipts)/1000000) %>%
#   filter(employee_premiums == 1) %>%
#   rename(year = fy) %>% 
#   select(-employee_premiums)

emp_premium_long <- rev_temp %>%  filter(employee_premiums == 1)
# 381 observations have employee premiums == 1


# drops employee premiums from revenue
# rev_temp <- rev_temp %>% filter(employee_premiums != 1)
# should be dropped in next step since rev_type = 51


```

Employee premiums are dropped in the following steps. In FY21, employee premiums were subtracted from state healthcare costs on the expenditure side to calculate a "Net Healthcare Cost" but that methodology has been discontinued. Totals were practically unchanged: revenue from employee premiums is also very small. 


### Transfers in and Out: 

Funds that hold and disperse local taxes or fees are dropped from the analysis. Then other excluded revenue types are also dropped.

Drops Blank, Student Fees, Retirement contributions, proceeds/investments, bond issue proceeds, interagency receipts, cook IGT, Prior year refunds:

**I don't have much faith in the transfers in and out steps- AWM**

I am currently choosing to exclude the totals from this step. Overall, this decreases the total revenues in "All Other Revenues" by a few million dollars.

- in_from_out <- c("0847", "0867", "1175", "1176", "1177", "1178", "1181", "1182", "1582", "1592", "1745", "1982", "2174", "2264")   
- See the methodology document for the list of what these sources/funds are.   


```{r step-3.8}
rev_temp <- rev_temp %>% 
  filter(in_ff == 1) %>% 
  mutate(local = ifelse(is.na(local), 0, local)) %>% # drops all revenue observations that were coded as "local == 1"
  filter(local != 1)

# 1175 doesnt exist?
in_from_out <- c("0847", "0867", "1175", "1176", "1177", "1178", "1181", "1182", "1582", "1592", "1745", "1982", "2174", "2264")

# what does this actually include:
# all are items with rev_type = 75 originally. 
in_out_df <- rev_temp %>%
  mutate(infromout = ifelse(source %in% in_from_out, 1, 0)) %>%
  filter(infromout == 1)

rev_temp <- rev_temp %>% 
  mutate(rev_type_new = ifelse(source %in% in_from_out, "76", rev_type))
# if source contains any of the codes in in_from_out, code them as 76 (all other rev).
# I end up excluding rev_76 in later steps
```



```{r droprevtypes}
# revenue types to drop
drop_type <- c("32", "45", "51", 
               "66", "72", "75", "79", "98")

# drops Blank, Student Fees, Retirement contributions, proceeds/investments,
# bond issue proceeds, interagency receipts, cook IGT, Prior year refunds.


rev_temp <- rev_temp %>% filter(!rev_type_new %in% drop_type)
# keep observations that do not have a revenue type mentioned in drop_type

table(rev_temp$rev_type_new)

rev_temp %>% 
  group_by(fy, rev_type_new) %>% 
  summarize(total_reciepts = sum(receipts)/1000000) %>%
  pivot_wider(names_from = rev_type_new, values_from = total_reciepts, names_prefix = "rev_") 

# combines smallest 4  categories to to "Other"
# they were the 4 smallest in past years, are they still the 4 smallest? 

rev_temp <- rev_temp %>%  
 mutate(rev_type_new = ifelse(rev_type=="30" | rev_type=="60" | rev_type=="63" | rev_type=="76", "78", rev_type_new))


#table(rev_temp$rev_type_new)  # check work



rm(rev_1998_2022)
rm(exp_1998_2022)


write.csv(exp_temp, "allexp_fy22_recoded_02242023.csv")
write.csv(rev_temp, "rev_fy22_recoded_02242023.csv")

```



## Pivoting and Merging

- Local Government Transfers (exp_970) should be on the expenditure side  

- ~~State employer group insurance contributions should be dropped to avoid double counting both the state.~~   Do not do this. This was done for FY21 only and will not be done again.   
- ~~Subtract employee insurance premiums from State Employee Healthcare Expenditures (group == 904) - Employee Premiums = Actual state healthcare costs.~~   
- ~~ff_exp$exp904 − emp_premium$employee_premiums_sum = statehealthcarecosts~~

  - Did in FY21, but not doing again. Minor difference in fiscal gap overall from change in methodology.   


### Revenues

I chose to drop rev_76 for Transfers in and Out because I do not understand why that step occurs in the previously used Stata code. Rev_76 was created and included in rev_78 for All Other Revenues in old Stata code for years before FY21 but that method has been discontinued for FY22.      

```{r final-ffrev-table}
ff_rev <- rev_temp %>% 
  group_by(rev_type_new, fy) %>% 
  summarize(sum_receipts = sum(receipts, na.rm=TRUE)/1000000 ) %>%
  pivot_wider(names_from = "rev_type_new", values_from = "sum_receipts", names_prefix = "rev_")

ff_rev<- left_join(ff_rev, tax_refund)

#ff_rev <- left_join(ff_rev, pension2_fy22, by=c("fy" = "year"))

#ff_rev <- left_join(ff_rev, eehc2_amt) 
ff_rev <- mutate_all(ff_rev, ~replace_na(.,0))


ff_rev <- ff_rev %>%
  mutate(rev_02 = rev_02 - ref_02,
         rev_03 = rev_03 - ref_03,
         rev_06 = rev_06 - ref_06,
         rev_09 = rev_09 - ref_09,
         rev_21 = rev_21 - ref_21,
         rev_24 = rev_24 - ref_24,
         rev_35 = rev_35 - ref_35

      #   rev_78new = rev_78 #+ pension_amt #+ eehc
         ) %>% 
  select(-c(ref_02:ref_35, rev_99, rev_NA, rev_76#, pension_amt , rev_76,
          #  , eehc
            ))

ff_rev


```


Since I already pivot_wider()ed the table in the previous code chunk, I now change each column's name by using rename() to set new variable names. Ideally the final dataframe would have both the variable name and the variable label but I have not done that yet. 

```{r}
aggregate_rev_labels <- ff_rev %>%
  rename("INDIVIDUAL INCOME TAXES, gross of local, net of refunds" = rev_02,
         "CORPORATE INCOME TAXES, gross of PPRT, net of refunds" = rev_03,
         "SALES TAXES, gross of local share" = rev_06 ,
         "MOTOR FUEL TAX, gross of local share, net of refunds" = rev_09 ,
         "PUBLIC UTILITY TAXES, gross of PPRT" = rev_12,
         "CIGARETTE TAXES" = rev_15 ,
         "LIQUOR GALLONAGE TAXES" = rev_18,
         "INHERITANCE TAX" = rev_21,
         "INSURANCE TAXES&FEES&LICENSES, net of refunds" = rev_24 ,
         "CORP FRANCHISE TAXES & FEES" = rev_27,
       # "HORSE RACING TAXES & FEES" = rev_30,  # in Other
         "MEDICAL PROVIDER ASSESSMENTS" = rev_31 ,
         # "GARNISHMENT-LEVIES " = rev_32 , # dropped
         "LOTTERY RECEIPTS" = rev_33 ,
         "OTHER TAXES" = rev_35,
         "RECEIPTS FROM REVENUE PRODUCNG" = rev_36, 
         "LICENSES, FEES & REGISTRATIONS" = rev_39 ,
         "MOTOR VEHICLE AND OPERATORS" = rev_42 ,
         #  "STUDENT FEES-UNIVERSITIES" = rev_45,   # dropped
         "RIVERBOAT WAGERING TAXES" = rev_48 ,
         # "RETIREMENT CONTRIBUTIONS " = rev_51, # dropped
         "GIFTS AND BEQUESTS" = rev_54, 
         "FEDERAL OTHER" = rev_57 ,
         "FEDERAL MEDICAID" = rev_58, 
         "FEDERAL TRANSPORTATION" = rev_59 ,
         #"OTHER GRANTS AND CONTRACTS" = rev_60, #other
       # "INVESTMENT INCOME" = rev_63, # other
         # "PROCEEDS,INVESTMENT MATURITIES" = rev_66 , #dropped
         # "BOND ISSUE PROCEEDS" = rev_72,  #dropped
         # "INTER-AGENCY RECEIPTS" = rev_75,  #dropped
      #  "TRANSFER IN FROM OUT FUNDS" = rev_76,  #other
         "ALL OTHER SOURCES" = rev_78,
         # "COOK COUNTY IGT" = rev_79, #dropped
         # "PRIOR YEAR REFUNDS" = rev_98 #dropped
  ) 

aggregate_rev_labels

```



### Expenditures

Create exp_970 for all local government transfers (exp_971 + exp_972  + exp_975 + exp_976).

~~Create state employee healthcare costs that reflects the health costs minus the optional insurance premiums that came in (904_new=904−med_option_amt_recent).~~ Do not do this. This was done for FY21 only and will not be done again. Small differences in overall Fiscal Gap from methodology change.  

```{r}

ff_exp <- exp_temp %>% 
  group_by(fy, group) %>% 
  summarize(sum_expenditures = sum(expenditure, na.rm=TRUE)/1000000 ) %>%
  pivot_wider(names_from = "group", values_from = "sum_expenditures", names_prefix = "exp_")%>%
  
    left_join(debt_keep_yearly) %>%
  mutate(exp_903 = debt_cost) %>%

  #  left_join(healthcare_costs_yearly) %>%

  # join state employee healthcare and subtract employee premiums
  # left_join(emp_premium, by = c("fy" = "year")) %>%
#  mutate(exp_904_new = (`healthcare_cost` - `employee_premiums_sum`)) %>% # state employee healthcare premiums
  
 # left_join(retirement_contributions) %>%
  #    mutate(exp_901_new = exp_901 - contributions/1000000) %>% #employee pension contributions


  # join local transfers and create exp_970
  left_join(transfers) %>%
  mutate(exp_970 = exp_971 + exp_972  + exp_975 + exp_976)

ff_exp<- ff_exp %>% select(-c(debt_cost, exp_971:exp_976)) # drop unwanted columns
ff_exp
```



# Clean Table Outputs 

Create total revenues and total expenditures only:

- after aggregating expenditures and revenues, pivoting wider, then I want to drop the columns that I no longer want and then pivot_longer(). After pivoting_longer() and creating `rev_long` and `exp_long`, expenditures and revenues are in the same format and can be combined together for the totals and gap each year.   



```{r rev-exp-totals-long}
rev_long <- pivot_longer(ff_rev, rev_02:rev_78, names_to = c("type","Category"), values_to = "Dollars", names_sep = "_") %>% 
  rename(Year = fy) %>%
  mutate(Category_name = case_when(
    Category == "02" ~ "INDIVIDUAL INCOME TAXES" ,
    Category == "03" ~ "CORPORATE INCOME TAXES" ,
    Category == "06" ~ "SALES TAXES" ,
    Category == "09" ~ "MOTOR FUEL TAX" ,
    Category == "12" ~ "PUBLIC UTILITY TAXES" ,
    Category == "15" ~ "CIGARETTE TAXES" ,
    Category == "18" ~ "LIQUOR GALLONAGE TAXES" ,
    Category == "21" ~ "INHERITANCE TAX" ,
    Category == "24" ~ "INSURANCE TAXES&FEES&LICENSES" ,
    Category == "27" ~ "CORP FRANCHISE TAXES & FEES" ,
    Category == "30" ~ "HORSE RACING TAXES & FEES",  # in Other
    Category == "31" ~ "MEDICAL PROVIDER ASSESSMENTS" ,
    Category == "32" ~ "GARNISHMENT-LEVIES" , # dropped
    Category == "33" ~  "LOTTERY RECEIPTS" ,
    Category == "35" ~  "OTHER TAXES" ,
    Category == "36" ~  "RECEIPTS FROM REVENUE PRODUCING", 
    Category == "39" ~  "LICENSES, FEES & REGISTRATIONS" ,
    Category == "42" ~  "MOTOR VEHICLE AND OPERATORS" ,
    Category == "45" ~  "STUDENT FEES-UNIVERSITIES",   # dropped
    Category == "48" ~  "RIVERBOAT WAGERING TAXES" ,
    Category == "51" ~  "RETIREMENT CONTRIBUTIONS" , # dropped
    Category == "54" ~ "GIFTS AND BEQUESTS", 
    Category == "57" ~  "FEDERAL OTHER" ,
    Category == "58" ~  "FEDERAL MEDICAID", 
    Category == "59" ~  "FEDERAL TRANSPORTATION" ,
    Category == "60" ~  "OTHER GRANTS AND CONTRACTS", #other
    Category == "63" ~  "INVESTMENT INCOME", # other
    Category == "66" ~ "PROCEEDS,INVESTMENT MATURITIES" , #dropped
    Category == "72" ~ "BOND ISSUE PROCEEDS",  #dropped
    Category == "75" ~  "INTER-AGENCY RECEIPTS ",  #dropped
    Category == "76" ~  "TRANSFER IN FROM OUT FUNDS",  #other
    Category == "78" ~  "ALL OTHER SOURCES" ,
    Category == "79" ~   "COOK COUNTY IGT", #dropped
    Category == "98" ~  "PRIOR YEAR REFUNDS", #dropped
                 T ~ "Check Me!"

  ) )%>% 
  mutate(Category_name = str_to_title(Category_name))


exp_long <- pivot_longer(ff_exp, exp_402:exp_970 , names_to = c("type", "Category"), values_to = "Dollars", names_sep = "_") %>% 
  rename(Year = fy ) %>% 
  mutate(Category_name = 
           case_when(
             Category == "402" ~ "AGING" ,
             Category == "406" ~ "AGRICULTURE", 
             Category == "416" ~ "CENTRAL MANAGEMENT",
             Category == "418" ~ "CHILDREN AND FAMILY SERVICES", 
             Category == "420" ~ "COMMERCE AND ECONOMIC OPPORTUNITY",
             Category == "422" ~ "NATURAL RESOURCES" ,
             Category == "426" ~ "CORRECTIONS",
             Category == "427" ~ "EMPLOYMENT SECURITY" ,
             Category == "444" ~ "HUMAN SERVICES" ,
             Category == "448" ~ "Innovation and Technology", # AWM added fy2022
             Category == "478" ~ "FAMILY SERVICES net Medicaid", 
             Category == "482" ~ "PUBLIC HEALTH", 
             Category == "492" ~ "REVENUE", 
             Category == "494" ~ "TRANSPORTATION" ,
             Category == "532" ~ "ENVIRONMENTAL PROTECT AGENCY" ,
             Category == "557" ~ "IL STATE TOLL HIGHWAY AUTH" ,
             Category == "684" ~ "IL COMMUNITY COLLEGE BOARD", 
             Category == "691" ~ "IL STUDENT ASSISTANCE COMM" ,
             Category == "900" ~ "NOT IN FRAME",
             Category == "901" ~ "STATE PENSION CONTRIBUTION",
             Category == "903" ~ "DEBT SERVICE",
             Category == "904" ~ "State Employee Healthcare",
             Category == "910" ~ "LEGISLATIVE"  ,
             Category == "920" ~ "JUDICIAL" ,
             Category == "930" ~ "ELECTED OFFICERS" , 
             Category == "940" ~ "OTHER HEALTH-RELATED", 
             Category == "941" ~ "PUBLIC SAFETY" ,
             Category == "942" ~ "ECON DEVT & INFRASTRUCTURE" ,
             Category == "943" ~ "CENTRAL SERVICES",
             Category == "944" ~ "BUS & PROFESSION REGULATION" ,
             Category == "945" ~ "MEDICAID" ,
             Category == "946" ~ "CAPITAL IMPROVEMENT" , 
             Category == "948" ~ "OTHER DEPARTMENTS" ,
             Category == "949" ~ "OTHER BOARDS & COMMISSIONS" ,
             Category == "959" ~ "K-12 EDUCATION" ,
             Category == "960" ~ "UNIVERSITY EDUCATION",
             Category == "970" ~ "Local Govt Transfers",
             T ~ "CHECK ME!")
           ) %>% 
  mutate(Category_name = str_to_title(Category_name))

#write_csv(exp_long, "expenditures_recoded_long_FY22.csv")
#write_csv(rev_long, "revenue_recoded_long_FY22.csv")

aggregated_totals_long <- rbind(rev_long, exp_long)
aggregated_totals_long



year_totals <- aggregated_totals_long %>% 
  group_by(type, Year) %>% 
  summarize(Dollars = sum(Dollars, na.rm = TRUE)) %>% 
  pivot_wider(names_from = "type", values_from = Dollars) %>% 

  rename(
         Expenditures = exp,
         Revenue = rev) %>%  
  mutate(`Fiscal Gap` = round(Revenue - Expenditures))
# %>%  arrange(desc(Year))
# creates variable for the Gap each year

year_totals
```





# Graphs


Graphs made from `aggregated_totals_long` dataframe. 

```{r}

annotation <- data.frame(
  x = c(2004, 2017, 2019),
  y = c(60, 50, 5),  
  label = c("Expenditures","Revenue", "Fiscal Gap")
)

# with trend lines:
# year_totals %>%  
#   ggplot() +
#   # geom_smooth adds regression line, graphed first so it appears behind line graph
#   geom_smooth(aes(x = Year, y = Revenue/1000), color = "rosybrown2", alpha = 0.7, method = "lm", se = FALSE) + 
#   geom_smooth(aes(x = Year, y = Expenditures/1000), color = "gray", method = "lm", se = FALSE) +
#   
#   # line graph of revenue and expenditures
#   geom_line(aes(x = Year, y = Revenue/1000), color = "Black", size=1) +
#   geom_line(aes(x = Year, y = Expenditures/1000), color = "red", size=1) +
#   
#   # labels
#     theme_bw() +
#   scale_y_continuous(labels = comma)+
#   xlab("Year") + 
#   ylab("Billions of Dollars")  +
#   ggtitle("Illinois Expenditures and Revenue Totals, 1998-2022")



## Dashed line versions for expenditures: 

fiscal_gap <- year_totals %>%  
  ggplot() +
  geom_hline(yintercept = 0) +
  # geom_smooth adds regression line, graphed first so it appears behind line graph
  geom_smooth(aes(x = Year, y = Revenue/1000), color = "gray", alpha = 0.7, method = "lm", se = FALSE) + 
  #  scale_linetype_manual(values="dashed")+
  geom_smooth(aes(x = Year, y = Expenditures/1000), color ="rosybrown2", linetype = "dotted", method = "lm", se = FALSE, alpha = 0.7) +

  # line graph of revenue and expenditures
  geom_line(aes(x = Year, y = Revenue/1000), color = "Black", size=1) +
  geom_line(aes(x = Year, y = Expenditures/1000, linetype = "dashed"), color = "red", lwd=1) +
  geom_line(aes(x = Year, y = (`Fiscal Gap`/1000)), color = "gray") +
  
  geom_text(data = annotation, aes(x=x, y=y, label=label))+
  # labels
    theme_bw() +
    theme(legend.position = "none")+

    scale_linetype_manual(values = c("dashed", "dashed")) +

#  scale_y_continuous(labels = comma)+
  xlab("Year") + 
  ylab("Billions of Dollars")  +
  ggtitle("Illinois Expenditures and Revenue Totals, 1998-2022")

fiscal_gap



# annotation_billions <- data.frame(
#   x = c(2004, 2017, 2019),
#   y = c(60, 50, 5),  
#   label = c("Expenditures","Revenue", "Fiscal Gap"))


fiscal_gap2 <- year_totals %>%  
  ggplot() +
  geom_hline(yintercept = 0) +
  geom_line(aes(x = Year, y = Revenue/1000), color = "Black", lwd=1) +
  geom_line(aes(x = Year, y = Expenditures/1000, linetype = "dashed"), color = "red", lwd=1) +
  geom_line(aes(x = Year, y = (`Fiscal Gap`/1000)), color = "gray") +
  
  geom_text(data = annotation, aes(x=x, y=y, label=label))+
    theme_bw() +
  theme(legend.position = "none")+
    scale_linetype_manual(values = c("dashed", "dashed")) +

  scale_y_continuous(labels = comma)+
  xlab("Year") + 
  ylab("Billions of Dollars")  +
  ggtitle("Illinois Expenditures and Revenue Totals, 1998-2022")

fiscal_gap2


```


**Expenditure and revenues when focusing on largest categories and combining others into "All Other Expenditures(Revenues)":**

```{r}
exp_long %>%
  filter( Year == 2022) %>%
  mutate(rank = rank(Dollars),
        Category_name = ifelse(rank > 13, Category_name, 'All Other Expenditures')) %>%
 # select(-c(Year, Dollars, rank)) %>%
  arrange(desc(Dollars)) %>%
  ggplot() + 
  geom_col(aes(x = fct_reorder(Category_name, `Dollars`), y = `Dollars`), fill = "rosybrown2")+ 
  coord_flip() +
      theme_bw() +
    labs(title = "Expenditures for FY2022") +
    xlab("") +
  ylab("Millions of Dollars")

rev_long %>%
  filter( Year == 2022) %>%
  mutate(rank = rank(Dollars),
        Category_name = ifelse(rank > 10, Category_name, 'All Other Sources')) %>%
  arrange(desc(Dollars)) %>%
  ggplot() + 
  geom_col(aes(x = fct_reorder(Category_name, `Dollars`), y = `Dollars`), fill = "dark gray")+ 
  coord_flip() +
      theme_bw() +
    labs(title = "Revenues for FY2022") +
    xlab("") +
  ylab("Millions of Dollars")
```

