# Pension Costs Included within Department Expenditures {#sec-all-code-pensionchange}


**An Explanation of a Methodological Change in How We Categorize Some Pension Spending**

Fiscal Futures’ conceptual framework requires categorizing expenditures according to the function they support rather than the budgetary funds through which they are funneled. Because of this, past reports counted most pension spending as part of the expenditures for the employees of the relevant agency. For example, pension spending that went as benefits to human service employees was counted as spending on human services. However, because most teachers are not state employees, in past years, spending to support teacher pensions was treated differently.  We accounted for this spending as a separate expenditure item rather than including it as part of state spending on K-12 education.

On reflection, we decided to treat these pension payments in parallel with other pension spending and, therefore, include them as part of K-12 education spending. In order to maintain consistency across years we have recalculated all previous years using this methodology.

However, we thought that readers also might be interested in total pension spending during current and past years.  We have added a separate table showing this information but emphasize that pension spending is already incorporated in other spending categories and should not be added to the total shown in Table 1 of the report.


```{r setup, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(tidyverse)
library(formatR)
library(lubridate)
library(scales)
library(kableExtra)
library(ggplot2)
library(readxl)
library(janitor)
#devtools::install_github("CMAP-REPOS/cmapplot")
library(cmapplot)

theme_set(theme_classic() )

move_to_last <- function(df, n) df[c(setdiff(seq_len(nrow(df)), n), n), ]


current_year <- 2025 # fiscal year, not calendar year
past_year=current_year-1

rev_temp <- read_csv(paste0("../../Fiscal Futures IGPA/Fiscal-Future-Topics/data/FY", current_year, " Files/rev_temp.csv"))

exp_temp <- read_csv(paste0("../../Fiscal Futures IGPA/Fiscal-Future-Topics/data/FY", current_year, " Files/exp_temp.csv"))
```



```{r include=FALSE}
update_recessions <- function(url = NULL, quietly = FALSE){

  # Use default URL if user does not override
  if (is_null(url) | missing(url)) {
    url <- "https://data.nber.org/data/cycles/business_cycle_dates.json"
  }
  
   # locally bind variable names
  start_char <- end_char <- start_date <- end_date <- ongoing <- index <- peak <- trough <- NULL

  return(
    # attempt to download and format recessions table
    tryCatch({
      recessions <- jsonlite::fromJSON(url) %>%
        # drop first row trough
        dplyr::slice(-1) %>%
        # convert peaks and troughs...
        dplyr::mutate(
          # ...to R dates
          start_date = as.Date(peak),
          end_date = as.Date(trough),
          # ... and clean char strings
          start_char = format(start_date, "%b %Y"),
          end_char = format(end_date, "%b %Y")) %>%
        # confirm ascending and create row number
        dplyr::arrange(start_date) %>%
        mutate(index = row_number()) %>%
        mutate(
          # Flag unfinished recessions
          ongoing = case_when(
            is.na(end_date) & index == max(.$index) ~ T,
            TRUE ~ F),
          # set ongoing recession to arbitrary future date
          end_date = case_when(
            ongoing ~ as.Date("2200-01-01"),
            TRUE ~ end_date),
          # mark ongoing recession in char field
          end_char = case_when(
            ongoing ~ "Ongoing",
            TRUE ~ end_char)
          ) %>%
        # clean up
        select(start_char, end_char, start_date, end_date, ongoing)

      if (!quietly) {message("Successfully fetched from NBER")}

      # Return recessions
      recessions
    },
    error = function(cond){
      if (!quietly) message("WARNING: Fetch or processing failed. `NULL` returned.")
      return(NULL)
    }
    )
  )
}

#recess_table <- list(build_recessions(update))
recessions <- update_recessions()

```

## Modify Expenditure File

### Tax refunds

Aggregate expenditures: Save tax refunds as negative revenue. Code refunds to match the rev_type codes (02=income taxes, 03 = corporate income taxes, 06=sales tax, 09=motor fuel tax, 24=insurance taxes and fees, 35 = all other tax refunds).

```{r}
#| label: fig-tax-refunds
#| fig-cap:  Tax Refunds

tax_refund_long <- exp_temp %>%           # fund != "0401" # removes State Trust Funds
  filter(fund != "0401" &
           (object == "9900" | object=="9910"|object=="9921"|object=="9923"|object=="9925")) %>%
  # keeps these objects which represent revenue, insurance, treasurer,and financial and professional reg tax refunds
  mutate(refund = case_when(
    object == "9900" ~ "FY23_Rebates",
    fund=="0278" & sequence == "00" ~ "02", # for income tax refund
    fund=="0278" & sequence == "01" ~ "03", # tax administration and enforcement and tax operations become corporate income tax refund
     fund == "0278" & sequence == "02" ~ "02",
    object=="9921" ~ "21",                # inheritance tax and estate tax refund appropriation
    object=="9923" ~ "09",                # motor fuel tax refunds
    obj_seq_type == "99250055" ~ "06",    # sales tax refund
    fund=="0378" & object=="9925" ~ "24", # insurance privilege tax refund
   (fund=="0001" & object=="9925") | (object=="9925" & fund == "0384" & fy == 2023) ~ "35", # all other taxes
   # fund=="0001" & object=="9925" ~ "35", # all other taxes
      T ~ "CHECK"))                       # if none of the items above apply to the observations, then code them as CHECK 

    
exp_temp <- left_join(exp_temp, tax_refund_long) %>%
  mutate(refund = ifelse(is.na(refund),"not refund", as.character(refund)))

tax_refund <- tax_refund_long %>% 
  group_by(refund, fy)%>%
  summarize(refund_amount = sum(expenditure, na.rm = TRUE)/1000000) %>%
  pivot_wider(names_from = refund, values_from = refund_amount, names_prefix = "ref_") %>%
  mutate_all(~replace_na(.,0)) %>%
  arrange(fy)
# 
# tax_refund %>% 
#   pivot_longer(c(ref_06:ref_35, ref_FY23_Rebates), names_to = "Refund Type", values_to = "Amount") %>%
#   ggplot()+  
#   theme_classic()+
#   geom_line(aes(x=fy,y=Amount, group = `Refund Type`, color = `Refund Type`))+
#   labs(title = "Refund Types", 
#        caption = "Refunds are excluded from Expenditure totals and instead subtracted from Revenue totals") + 
#   labs(title = "Tax refunds", 
#        caption = "Rev_type codes: 02=income taxes, 03=corporate income taxes, 06=sales tax, 09=motor fuel tax, 
#        24=insurance taxes and fees, 35 = all other tax refunds." )

# remove the items we recoded in tax_refund_long
# exp_temp <- exp_temp %>% filter(refund == "not refund")


```

```{r}
# manually adds the abatements as expenditure item and keeps on expenditure side.
# otherwise ignored since it is in fund 0278 and 
exp_temp <- exp_temp %>%
  mutate(in_ff = ifelse(object == 9900, 1, in_ff))
```

### Pension Expenditures

State pension contributions for TRS and SURS are largely captured with object=4431. **(State payments into pension fund).** State payments to the following pension systems:

-   Teachers Retirement System (TRS) Agency 493 -->    K-12 Education (Group = 959)
-   New POB bond in 2019: Accelerated Bond Fund paid benefits in advance as lump sum   
-   State Employee Retirement System (SERS) Agency 589 -->    SERS Agency 589 - _Note: Object 4431 does not have SERS expenditures in it. Those are only in object 116X objects_   
-   State University Retirement System (SURS) Agency 693 -->  University Education (Group = 960)   
-   Judges Retirement System (JRS) Agency 275  --> Judicial (Group = 920)   
-   General Assembly Retirement System (GARS)  --> Legislative (Group 910)    


There are also "Other Post-Employment Benefits" (OPEBs). Expenditure object 4430 is for retirement benefits.

While it is good to know the overall cost of pensions for the state, if you want to know the true cost of providing services, pension and other benefit costs should be included in the department that is paying employees to provide those services. 


> Change in pension coding in chunk below:

```{r fig-pensions, fig.cap="Pensions"}


exp_temp <-  exp_temp %>% 
  arrange(fund) %>%
  mutate(pension = case_when( 
    
    ## Commented out line below:
    # (object=="4431") ~ 1, # 4431 = easy to find pension payments INTO fund
    

    
    (object=="1298" &  # Purchase of Investments, Normally excluded
       (fy==2010 | fy==2011) & 
       (fund=="0477" | fund=="0479" | fund=="0481")) ~ 3, #judges retirement OUT of fund
    # state borrowed money from pension funds to pay for core services during 2010 and 2011. 
    # used to fill budget gap and push problems to the future. 
    
    fund == "0319" ~ 4, # pension stabilization fund
    TRUE ~ 0) )

```

```{r fig-pensions-POB, fig.cap="Pension Expenditures"}

# special accounting of pension obligation bond (POB)-funded contributions to JRS, SERS, GARS, TRS 

exp_temp <- exp_temp %>% 
  # change object for 2010 and 2011, retirement expenditures were bond proceeds and would have been excluded
  mutate(object = ifelse((pension > 0 & in_ff == "0"), "4431", object)) %>% 
  # changes weird teacher & judge retirement system  pensions object to normal pension object 4431
  mutate(pension =  ifelse(pension > 0 & in_ff == "0", 6, pension)) %>% # coded as 6 if it was supposed to be excluded. 
  mutate(in_ff = ifelse(pension > 0, "1", in_ff))



# all other pensions objects  codes get agency code 901 for State Pension Contributions
exp_temp <- exp_temp %>% 
  mutate(agency = ifelse(pension > 0, "901", as.character(agency)),
         agency_name = ifelse(agency == "901", "State Pension Contributions", as.character(agency_name)))

```

Can also be thought of past commitments vs current contributions.
Cost of past commitments in the form of Pension benefits paid out, 



**Current Employees vs Retired Employees**

Current Employees: 
- Group Insurance Benefits  

Retired Employees:
- Deferred Compensation  
- Medicare Retirees and Survivors of State of Illinois Employees Group Insurance Program (SEGIP)    
  - Part of Medicare   
  
  
  
```{r}
exp_temp|>
  filter(fy==2024) |>
  filter((appr_org=="01" | appr_org == "65" | appr_org=="88") & (object=="4900" | object=="4400") )|> 
  group_by(agency, agency_name) |> # separates CHIP from health and human services and saves it as Medicaid
  summarize(expenditure = sum(expenditure))
```
  
  
  
Drop all cash transfers between funds, statutory transfers, and purchases of investments from expenditure data.


```{r drop-transfers}

transfers_drop <- exp_temp %>% filter(
  agency == "799" | # statutory transfers
           object == "1993" |  # interfund cash transfers
           object == "1298") # purchase of investments
transfers_drop # items being dropped, 
# always check to make sure you aren't accidentally dropping something of interest.

exp_temp <- anti_join(exp_temp, transfers_drop)


```

### State employee healthcare costs

> Commented out line of code that seperates healthcare costs. This should keep healthcare costs in the agency, similar to the change that was made for pensions.

    agency=="507"|  agency=="442" | agency=="445" | agency=="452" |agency=="458" | agency=="497" ~ "948", # other departments


It looks like healthcare costs shift to "Other Departments" which includes : 

- GOMB (507)  
- Human Rights (442)  
- Illinois Power Agency (445)  
- Labor (452)   
- State Lottery (458)   
- Veteran's Affairs (497)    

```{r}
exp_temp |> filter(org_name == "BUREAU OF BENEFITS") |> 
  group_by(fy, agency) |>
  summarize(expenditure = sum(expenditure) ) |>
  pivot_wider(names_from = "fy", values_from = "expenditure")

exp_temp |> 
  filter(org_name == "BUREAU OF BENEFITS") |> 
  group_by(fy, agency) |>
  summarize(expenditure = sum(expenditure) ) |>
  ggplot() + 
  geom_line(aes(x=fy, y=expenditure, group= agency, color = agency)) + 
  scale_y_continuous(labels = scales::dollar) +
  labs(title="Bureau of Benefits Expenditures")

exp_temp |> 
  filter(org_name == "BUREAU OF BENEFITS") |> 
  group_by(fy, object) |>
  summarize(expenditure = sum(expenditure) ) |>
  ggplot() + 
  geom_line(aes(x=fy, y=expenditure, group= object, color = object)) + 
  scale_y_continuous(labels = scales::dollar) +
  labs(title="Bureau of Benefits Expenditures")
```


```{r eehc1}
#if observation is a group insurance contribution, then the expenditure amount is set to $0 (essentially dropped from analysis)

# pretend eehc is named group_insurance_contribution or something like that
# eehc coded as zero implies that it is group insurance
# if eehc=0, then expenditures are coded as zero for group insurance to avoid double counting costs


exp_temp <- exp_temp %>% 
  mutate(eehc = ifelse(
    # group insurance contributions for 1998-2005 and 2013-present
   fund == "0001" & (object == "1180" | object =="1900") & agency == "416" & appr_org=="20", 0, 1) )%>% 
  
  mutate(eehc = ifelse(
    # group insurance contributions for 2006-2012
    fund == "0001" & object == "1180" & agency == "478" 
    & appr_org=="80", 0, eehc) )%>%
    
   # group insurance contributions from road fund
  # coded with 1900 for some reason??
    mutate(eehc = ifelse(
      fund == "0011" & object == "1900" & 
        agency == "416" & appr_org=="20", 0, eehc) ) %>%
  
  mutate(expenditure = ifelse(eehc=="0", 0, expenditure)) %>%

  mutate(agency = case_when(
    ## turns specific items into State Employee Healthcare (agency=904)
      fund=="0907" & (agency=="416" & appr_org=="20") ~ "904",   # central management Bureau of benefits using health insurance reserve
      fund=="0907" & (agency=="478" & appr_org=="80") ~ "904",   # agency = 478: healthcare & family services using health insurance reserve - stopped using this in 2012
       TRUE ~ as.character(agency))) %>%
  
  mutate(agency_name = ifelse(
    agency == "904", "STATE EMPLOYEE HEALTHCARE", as.character(agency_name)),
         in_ff = ifelse(agency == "904", 1, in_ff),
         group = ifelse(agency == "904", "904", as.character(agency)))  
# creates group variable

# Default group = agency number

healthcare_costs <- exp_temp %>% filter(group == "904")

```



```{r transfers-to-local}
exp_temp <- exp_temp %>% mutate(
  agency = case_when(fund=="0515" & object=="4470" & type=="08" ~ "971", # income tax to local governments
                     fund=="0515" & object=="4491" & type=="08" & sequence=="00" ~ "971", # object is shared revenue payments
                     fund=="0802" & object=="4491" ~ "972", #pprt transfer
                     fund=="0515" & object=="4491" & type=="08" & sequence=="01" ~ "976", #gst to local
                     fund=="0627" & object=="4472"~ "976" , # public transportation fund but no observations exist
                     fund=="0648" & object=="4472" ~ "976", # downstate public transportation, but doesn't exist
                     fund=="0515" & object=="4470" & type=="00" ~ "976", # object 4470 is grants to local governments
                    object=="4491" & (fund=="0188"|fund=="0189") ~ "976",
                     fund=="0187" & object=="4470" ~ "976",
                     fund=="0186" & object=="4470" ~ "976",
                    object=="4491" & (fund=="0413"|fund=="0414"|fund=="0415")  ~ "975", #mft to local
                  fund == "0952"~ "975", # Added Sept 29 2022 AWM. Transportation Renewal MFT
                    TRUE ~ as.character(agency)),
  
  agency_name = case_when(agency == "971"~ "INCOME TAX 1/10 TO LOCAL",
                          agency == "972" ~ "PPRT TRANSFER TO LOCAL",
                          agency == "975" ~ "MFT TO LOCAL",
                          agency == "976" ~ "GST TO LOCAL",
                          TRUE~as.character(agency_name)),
  group = ifelse(agency>"970" & agency < "977", as.character(agency), as.character(group)))

```

```{r fig-drop-local-transfers, fig.cap="Drop Transfers from State to Local Governments"}
transfers_long <- exp_temp %>% 
  filter(group == "971" |group == "972" | group == "975" | group == "976")


transfers <- transfers_long %>%
  group_by(fy, group ) %>%
  summarize(sum_expenditure = sum(expenditure)/1000000) %>%
  pivot_wider(names_from = "group", values_from = "sum_expenditure", names_prefix = "exp_" )

exp_temp <- anti_join(exp_temp, transfers_long)


dropped_inff_0 <- exp_temp %>% filter(in_ff == 0)

exp_temp <- exp_temp %>% filter(in_ff == 1) # drops in_ff = 0 funds AFTER dealing with net-revenue above

```



```{r debt-service}
debt_drop <- exp_temp %>% 
  filter(object == "8841" |  object == "8811")  
# escrow  OR  principle

#debt_drop %>% group_by(fy) %>% summarize(sum = sum(expenditure)) %>% arrange(-fy)


debt_keep <- exp_temp %>% 
  filter(fund != "0455" & (object == "8813" | object == "8800" )) 
# examine the debt costs we want to include

#debt_keep %>% group_by(fy) %>% summarize(sum = sum(expenditure)) %>% arrange(-fy) 


exp_temp <- anti_join(exp_temp, debt_drop) 
exp_temp <- anti_join(exp_temp, debt_keep)

debt_keep <- debt_keep %>%
  mutate(
    agency = ifelse(fund != "0455" & (object == "8813" | object == "8800"), "903", as.character(agency)),
    group = ifelse(fund != "0455" & (object == "8813" | object == "8800"), "903", as.character(group)),
    in_ff = ifelse(group == "903", 1, as.character(in_ff)))

debt_keep_yearly <- debt_keep %>% 
  group_by(fy, group) %>% 
  summarize(debt_cost = sum(expenditure,na.rm=TRUE)/1000000) %>% 
  select(-group)

```

### Add Other Fiscal Future group codes

> Commented out line that creates Other departments from a few agencies. Keeping them seperate to see where healthcare costs go. 


    agency=="507"|  agency=="442" | agency=="445" | agency=="452" |agency=="458" | agency=="497" ~ "948", # other departments



```{r group-codes}


exp_temp <- exp_temp %>%
  #mutate(agency = as.numeric(agency) ) %>%
  # arrange(agency)%>%
  mutate(
    group = case_when(
      agency>"100"& agency<"200" ~ "910", # legislative
      
      agency == "528"  | (agency>"200" & agency<"300") ~ "920", # judicial
      
      ######################################################
      # Not used if we are not separating pension costs!!
      # pension > 0  ~ "901", # pensions
      
      ## New CODE: April 23rd, 2025:
      agency == "593" ~ "959", #  TRS becomes part of K-12 costs
      agency == "594" ~ "959",   # TRS
      agency == "589" ~ "948", # SERS becomes part of "Other Agencies"
      agency == "693" ~ "960", # SURS becomes part of group 960
      agency == "275" ~ "920",  # JRS becomes part of group 920
      agency == "131" ~  "910", # GARS becomes part of Group 910
      ######################################################
      
      (agency>"309" & agency<"400") ~ "930",    # elected officers: Governor, lt gov, attorney general, sec. of state, comptroller, treasurer
      
      agency == "586" ~ "959", # create new K-12 group

      agency=="402" | agency=="418" | agency=="478" | agency=="444" | agency=="482" ~ as.character(agency), # aging, CFS, HFS, human services, public health
      
      T ~ as.character(group))
    ) %>%      

  
  mutate(group = case_when(
    agency=="478" & (appr_org=="01" | appr_org == "65" | appr_org=="88") & (object=="4900" | object=="4400") ~ "945", # separates CHIP from health and human services and saves it as Medicaid
    
    agency == "586" & fund == "0355" ~ "945",  # 586 (Board of Edu) has special education which is part of medicaid
    
    # OLD CODE: agency == "586" & appr_org == "18" ~ "945", # Spec. Edu Medicaid Matching
    
    agency=="425" | agency=="466" | agency=="546" | agency=="569" | agency=="578" | agency=="583" | agency=="591" | agency=="592" | agency=="493" | agency=="588" ~ "941", # public safety & Corrections
    
    agency=="420" | agency=="494" |  agency=="406" | agency=="557" ~ as.character(agency), # econ devt & infra, tollway
    
    agency=="511" | agency=="554" | agency=="574" | agency=="598" ~ "946",  # Capital improvement
    
    agency=="422" | agency=="532" ~ as.character(agency), # environment & nat. resources
    
    agency=="440" | agency=="446" | agency=="524" | agency=="563"  ~ "944", # business regulation
    
    agency=="492" ~ "492", # revenue
    
    agency == "416" ~ "416", # central management services
    agency=="448" & fy > 2016 ~ "416", #add DoIT to central management 
    
    T ~ as.character(group))) %>%
  
  
  mutate(group = case_when(
    # agency=="684" | agency=="691"  ~ as.character(agency), # moved under higher education in next line. 11/28/2022 AWM
    
    agency=="692" | agency == "693" | agency=="695" | agency == "684" |agency == "691" | (agency>"599" & agency<"677") ~ "960", # higher education
    
    agency=="427"  ~ as.character(agency), # employment security
    
############################ 
# Leaving these agencies as their own agency number for now. Had been coded to "Other departments" Group 948
# - GOMB (507)  
# - Human Rights (442)  
# - Illinois Power Agency (445)  
# - Labor (452)   
# - State Lottery (458)   
# - Veteran's Affairs (497) 

  #     agency=="507" | agency=="442" | agency=="445" | agency=="452" |agency=="458" | agency=="497" ~ as.character(agency), # Were included within "other departments"
    
    agency=="507"|  agency=="442" | agency=="445" | agency=="452" |agency=="458" | agency=="497" ~ "948", # other departments
###########################################
    

# other boards & Commissions
    agency=="503" | agency=="509" | agency=="510" | agency=="565" |agency=="517" | agency=="525" | agency=="526" | agency=="529" | agency=="537" | agency=="541" | agency=="542" | agency=="548" |  agency=="555" | agency=="558" | agency=="559" | agency=="562" | agency=="564" | agency=="568" | agency=="579" | agency=="580" | agency=="587" | agency=="590" | agency=="527" | agency=="585" | agency=="567" | agency=="571" | agency=="575" | agency=="540" | agency=="576" | agency=="564" | agency=="534" | agency=="520" | agency=="506" | agency == "533" ~ "949", 
    
  #   # Other Departments
  #   agency=="131" |
  # #  agency=="275" | #JRS
  # #  agency=="589" | #SERS
  #  # agency=="593"|  # TRS
  # #  agency=="594"| # Also TRS
  # #  agency=="693"   #SURS
  #  ~ "948",
    
    T ~ as.character(group))) %>%
  
  mutate(group_name = 
           case_when(
             group == "416" ~ "Central Management",
             group == "442" ~ "Human Rights",
             group == "445"  ~ "Illinois Power Agency",
             group == "452" ~ "Labor",
             group == "458" ~ "State Lottery",
             group == "489" ~ "SERS",
             group == "478" ~ "Healthcare and Family Services",
             group == "482" ~ "Public Health",
             group == "497" ~ "Veteran's Affairs",
             group == "507" ~ "GOMB",
             group == "900" ~ "NOT IN FRAME",
             group == "901" ~ "STATE PENSION CONTRIBUTION",
             group == "903" ~ "DEBT SERVICE",
             group == "910" ~ "LEGISLATIVE"  ,
             group == "920" ~ "JUDICIAL" ,
             group == "930" ~ "ELECTED OFFICERS" , 
             group == "940" ~ "OTHER HEALTH-RELATED", 
             group == "941" ~ "PUBLIC SAFETY" ,
             group == "942" ~ "ECON DEVT & INFRASTRUCTURE" ,
             group == "943" ~ "CENTRAL SERVICES",
             group == "944" ~ "BUS & PROFESSION REGULATION" ,
             group == "945" ~ "MEDICAID" ,
             group == "946" ~ "CAPITAL IMPROVEMENT" , 
             group == "948" ~ "OTHER DEPARTMENTS" ,
             group == "949" ~ "OTHER BOARDS & COMMISSIONS" ,
             group == "959" ~ "K-12 EDUCATION" ,
             group == "960" ~ "UNIVERSITY EDUCATION" ,
             group == agency ~ as.character(agency_name),
             TRUE ~ "Check name"),
         year = fy)

exp_temp %>% filter(group_name == "Check name")

write_csv(exp_temp, file = paste0("./data/FY", current_year, " Files/all_expenditures_recoded_pensionrecode.csv"))

```

::: callout-important
All expenditures recoded but not aggregated: Allows for inspection of individual expenditures within larger categories. This stage of the data is extremely useful for investigating how individual items have been coded before they are aggregated into larger categories.
:::

## Modify Revenue data

```{r rev-recode-agencies, warning = FALSE, message=FALSE}

# recodes old agency numbers to consistent agency number
rev_temp <- rev_temp %>% 
  mutate(agency = case_when(
    (agency=="438"| agency=="475" |agency == "505") ~ "440",
    # financial institution &  professional regulation &
     # banks and real estate  --> coded as  financial and professional reg
    agency == "473" ~ "588", # nuclear safety moved into IEMA
    (agency =="531" | agency =="577") ~ "532", # coded as EPA
    (agency =="556" | agency == "538") ~ "406", # coded as agriculture
    agency == "560" ~ "592", # IL finance authority (fire trucks and agriculture stuff)to state fire marshal
    agency == "570" & fund == "0011" ~ "494",   # city of Chicago road fund to transportation
    TRUE ~ (as.character(agency)))) 
```



```{r fig-create-rev-federal-transfers}
#rev_temp <- rev_temp %>% filter(in_ff==1)

rev_temp <- rev_temp %>% 
  mutate(
    rev_type = ifelse(rev_type=="57" & agency=="478" & (source=="0618"|source=="2364"|source=="0660"|source=="1552"| source=="2306"| source=="2076"|source=="0676"|source=="0692"), "58", rev_type),
    rev_type_name = ifelse(rev_type=="58", "Federal Medicaid Reimbursements", rev_type_name),
    rev_type = ifelse(rev_type=="57" & agency=="494", "59", rev_type),
    rev_type_name = ifelse(rev_type=="59", "Federal Transportation", rev_type_name),
    rev_type_name = ifelse(rev_type=="57", "Federal - Other", rev_type_name),
    rev_type = ifelse(rev_type=="6", "06", rev_type),
    rev_type = ifelse(rev_type=="9", "09", rev_type)) 

```

```{r coding-stimuluspackage-dollars}
rev_temp <- rev_temp %>% mutate(covid_dollars = ifelse(source_name_AWM == "FEDERAL STIMULUS PACKAGE",1,0))

rev_temp %>% filter(source_name_AWM == "FEDERAL STIMULUS PACKAGE") %>%
  group_by(fy) %>% summarize(Received = sum(receipts))
```

### Health Insurance Premiums from Employees

Insurance premiums for employees is coded below but it is NOT used in the fiscal futures model. Employee and employer premiums are considered rev_51 and dropped from analysis in later step.

-   0120 = ins prem-option life
-   0120 = ins prem-optional life/univ
-   0347 = optional health - HMO
-   0348 = optional health - dental
-   0349 = optional health - univ/local SI
-   0350 = optional health - univ/local
-   0351 = optional health - retirement
-   0352 = optional health - retirement SI
-   0353 = optional health - retire/dental
-   0354 = optional health - retirement hmo
-   2199-2209 = various HMOs, dental, health plans from Health Insurance Reserve (fund)

```{r insurance-premiums}

#collect optional insurance premiums to fund 0907 for use in eehc expenditure  
rev_temp <- rev_temp %>% 
  mutate(
    employee_premiums = ifelse(fund=="0907" & (source=="0120"| source=="0121"| (source>"0345" & source<"0357")|(source>"2199" & source<"2209")), 1, 0),
    
    # adds more rev_type codes
    rev_type = case_when(
      fund =="0427" ~ "12", # pub utility tax
      fund == "0742" | fund == "0473" ~ "24", # insurance and fees
      fund == "0976" ~ "36",# receipts from rev producing
      fund == "0392" |fund == "0723" ~ "39", # licenses and fees
      fund == "0656" ~ "78", #all other rev sources
      TRUE ~ as.character(rev_type)))
# if not mentioned, then rev_type as it was



# # optional insurance premiums = employee insurance premiums

# emp_premium <- rev_temp %>%
#   group_by(fy, employee_premiums) %>%
#   summarize(employee_premiums_sum = sum(receipts)/1000000) %>%
#   filter(employee_premiums == 1) %>%
#   rename(year = fy) %>% 
#   select(-employee_premiums)

emp_premium_long <- rev_temp %>%  filter(employee_premiums == 1)
# 381 observations have employee premiums == 1


# drops employee premiums from revenue
# rev_temp <- rev_temp %>% filter(employee_premiums != 1)
# should be dropped in next step since rev_type = 51


```


### Transfers in and Out:

Funds that hold and disperse local taxes or fees are dropped from the analysis. Then other excluded revenue types are also dropped.

Drops Blank, Student Fees, Retirement contributions, proceeds/investments, bond issue proceeds, interagency receipts, cook IGT, Prior year refunds:


```{r}
rev_temp <- rev_temp %>% 
  filter(in_ff == 1) %>% 
  mutate(local = ifelse(is.na(local), 0, local)) %>% # drops all revenue observations that were coded as "local == 1"
  filter(local != 1)

# 1175 doesnt exist?
in_from_out <- c("0847", "0867", "1175", "1176", "1177", "1178", "1181", "1182", "1582", "1592", "1745", "1982", "2174", "2264")

# what does this actually include:
# all are items with rev_type = 75 originally. 
in_out_df <- rev_temp %>%
  mutate(infromout = ifelse(source %in% in_from_out, 1, 0)) %>%
  filter(infromout == 1)

rev_temp <- rev_temp %>% 
   mutate(rev_type_new = ifelse(source %in% in_from_out, "76", rev_type))
# if source contains any of the codes in in_from_out, code them as 76 (all other rev).
# I end up excluding rev_76 in later steps
```

> Are the 4 smallest categories from past years still the 4 smallest categories? Check it each year.

```{r droprevtypes}
# revenue types to drop
drop_type <- c("32", "45", "51", 
               "66", "72", "75", "76", "79", "98", "99")

# drops Blank, Student Fees, Retirement contributions, proceeds/investments,
# bond issue proceeds, interagency receipts, cook IGT, Prior year refunds.


rev_temp <- rev_temp %>% filter(!rev_type_new %in% drop_type)
# keep observations that do not have a revenue type mentioned in drop_type

table(rev_temp$rev_type_new)

rev_temp %>% 
  group_by(fy, rev_type_new) %>% 
  summarize(total_reciepts = sum(receipts)/1000000) %>%
  pivot_wider(names_from = rev_type_new, values_from = total_reciepts, names_prefix = "rev_") 

# combines smallest 4  categories to to "Other"
# they were the 4 smallest in past years, are they still the 4 smallest? 

# rev_temp <- rev_temp %>%  
#  mutate(rev_type_new = ifelse(rev_type=="30" | rev_type=="60" | rev_type=="63" ,
#                             ##  | rev_type=="76", 
#                               "78", rev_type_new))


#table(rev_temp$rev_type_new)  # check work



rm(rev_1998_2022)
rm(exp_1998_2022)


write_csv(exp_temp, file = paste0("./data/FY", current_year, " Files/exp_pensionrecode.csv"))
write_csv(rev_temp, file = paste0("./data/FY", current_year, " Files/rev_pensionrecode.csv"))


```

## Pivoting and Merging

-   Local Government Transfers (exp_970) should be on the expenditure side

### Revenues

<!--- If there are NA rev types, it will cause the code to  break when binding dataframes together later --->


```{r code--check, include=FALSE}
rev_temp %>% filter(is.na(rev_type))

```

```{r}
#| label: tbl-final-ffrev-table
#| tbl-cap: "Pivoted Revenue Table ($ Millions) - Intermediate Step. Not actually used for anything other than to have output in same format as old STATA output to make it easily comparable."


ff_rev <- rev_temp %>% 
  group_by(rev_type_new, fy) %>% 
  summarize(sum_receipts = sum(receipts, na.rm=TRUE)/1000000 ) %>%
  pivot_wider(names_from = "rev_type_new", values_from = "sum_receipts", names_prefix = "rev_")


ff_rev <- mutate_all(ff_rev, ~replace_na(.,0))

# 
# ff_rev <- ff_rev %>%
#   mutate(rev_02 = rev_02 - ref_02,
#          rev_03 = rev_03 - ref_03,
#          rev_06 = rev_06 - ref_06,
#          rev_09 = rev_09 - ref_09,
#          rev_21 = rev_21 - ref_21,
#          rev_24 = rev_24 - ref_24,
#          rev_35 = rev_35 - ref_35
# 
#       #   rev_78new = rev_78 #+ pension_amt #+ eehc
#          ) %>% 
#   select(-c(ref_02:ref_35, rev_99, rev_NA, rev_76
#             #, ref_CHECK#, pension_amt , rev_76,
#           #  , eehc
#             ))
# 
# ff_rev



#noproblem <- c(0)  # if ref_CHECK = $0, then there is no problem. :) 
# 
# if((sum(ff_rev$ref_CHECK) == 0 )){
# 
# ff_rev <- ff_rev %>%
#   
#   mutate(rev_02 = rev_02 - ref_02,
#          rev_03 = rev_03 - ref_03,
#          rev_06 = rev_06 - ref_06,
#          rev_09 = rev_09 - ref_09,
#          rev_21 = rev_21 - ref_21,
#          rev_24 = rev_24 - ref_24,
#          rev_35 = rev_35 - ref_35
#          ) %>% 
#   select(-c(ref_02:ref_35, rev_99, rev_76, ref_CHECK )) 
# }else{"You have a problem! Check what revenue items did not have rev codes (causing it to be coded as rev_NA) or the check if there were refunds that were not assigned revenue codes (tax_refunds_long objects)"}

ff_rev %>%  mutate_all(., ~round(.,digits=0))
```


```{r}
#| label: tbl-labeled-final-ffrev-table
#| tbl-cap: "Aggregated Revenue Categories ($ Millions), with old labels. Not actually used for anything other than to have output that was easiy comparable to old STATA output from past years."
#| include: false

# Since I already pivot_wider()ed the table in the previous code chunk, I now change each column's name by using rename() to set new variable names. Ideally the final dataframe would have both the variable name and the variable label but I have not done that yet.


aggregate_rev_labels <- ff_rev %>%
  rename("INDIVIDUAL INCOME TAXES, gross of local, net of refunds" = rev_02,
         "CORPORATE INCOME TAXES, gross of PPRT, net of refunds" = rev_03,
         "SALES TAXES, gross of local share" = rev_06 ,
         "MOTOR FUEL TAX, gross of local share, net of refunds" = rev_09 ,
         "PUBLIC UTILITY TAXES, gross of PPRT" = rev_12,
         "CIGARETTE TAXES" = rev_15 ,
         "LIQUOR GALLONAGE TAXES" = rev_18,
         "INHERITANCE TAX" = rev_21,
         "INSURANCE TAXES&FEES&LICENSES, net of refunds" = rev_24 ,
         "CORP FRANCHISE TAXES & FEES" = rev_27,
        "HORSE RACING TAXES & FEES" = rev_30,  # in Other
         "MEDICAL PROVIDER ASSESSMENTS" = rev_31 ,
         # "GARNISHMENT-LEVIES " = rev_32 , # dropped
         "LOTTERY RECEIPTS" = rev_33 ,
         "OTHER TAXES" = rev_35,
         "RECEIPTS FROM REVENUE PRODUCNG" = rev_36, 
         "LICENSES, FEES & REGISTRATIONS" = rev_39 ,
         "MOTOR VEHICLE AND OPERATORS" = rev_42 ,
         #  "STUDENT FEES-UNIVERSITIES" = rev_45,   # dropped
         "RIVERBOAT WAGERING TAXES" = rev_48 ,
         # "RETIREMENT CONTRIBUTIONS " = rev_51, # dropped
         "GIFTS AND BEQUESTS" = rev_54, 
         "FEDERAL OTHER" = rev_57 ,
         "FEDERAL MEDICAID" = rev_58, 
         "FEDERAL TRANSPORTATION" = rev_59 ,
         "OTHER GRANTS AND CONTRACTS" = rev_60, #other
        "INVESTMENT INCOME" = rev_63, # other
         # "PROCEEDS,INVESTMENT MATURITIES" = rev_66 , #dropped
         # "BOND ISSUE PROCEEDS" = rev_72,  #dropped
         # "INTER-AGENCY RECEIPTS" = rev_75,  #dropped
       # "TRANSFER IN FROM OUT FUNDS" = rev_76,  # dropped
         "ALL OTHER SOURCES" = rev_78,
         # "COOK COUNTY IGT" = rev_79, #dropped
         # "PRIOR YEAR REFUNDS" = rev_98 #dropped
  ) 

aggregate_rev_labels %>% mutate_all(., ~round(., digits = 0))

```

### Expenditures

Create exp_970 for all local government transfers (exp_971 + exp_972 + exp_975 + exp_976).

```{r}
#| label: tbl-ffexp-notlabeled
#| tbl-cap: "Pivoted Expenditure Categories ($ Millions). Intermediate step. Not actually used for anything other than having output similar to past STATA output."
#| include: false

ff_exp <- exp_temp %>% 
  group_by(fy, group) %>% 
  summarize(sum_expenditures = sum(expenditure, na.rm=TRUE)/1000000 ) %>%
  pivot_wider(names_from = "group", values_from = "sum_expenditures", names_prefix = "exp_")%>%
  
    left_join(debt_keep_yearly) %>%
  rename(exp_903 = debt_cost) %>%
 #  join local transfers and create exp_970
  left_join(transfers) %>%
  mutate(exp_970 = exp_971 + exp_972  + exp_975 + exp_976)

 ff_exp<- ff_exp %>% 
  select(-c(exp_971:exp_976)) # drop unwanted columns that are already included in exp_970

ff_exp # not labeled
```

```{r}
#| label: tbl-final-table-with-labels
#| tbl-cap: "Final Expenditure Categories, with Fiscal Futures Grouped Expenditure Categories. Intermediate step. Not actually used for anything other than having output similar to past STATA output."
#| include: false

aggregate_exp_labeled <- exp_temp %>%
  group_by(year, group_name) %>%
  summarize(sum_expenditure = sum(expenditure)/1000000) %>%
  arrange(year) %>%
  pivot_wider(names_from = "group_name", values_from = "sum_expenditure")

aggregate_exp_labeled %>% mutate_all(., ~round(., digits = 0))
```

# Graphs and Tables {#sec-graphs-and-tables}

Create total revenues and total expenditures only:

-   after aggregating expenditures and revenues, pivoting wider, then I want to drop the columns that I no longer want and then pivot_longer(). After pivoting_longer() and creating `rev_long` and `exp_long`, expenditures and revenues are in the same format and can be combined together for the totals and gap each year.


```{r}
#| label: tbl-rev-and-exp-totals-long
#| tbl-cap: "Long Version of Data that has Revenue and Expenditures in One Dataframe. Creates `expenditures_recoded_long_pensionchange_FY`, `revenues_recoded_long_pensionchange_FY` and `aggregated_totals_pensionchange` which are exported as CSVs."
#| tbl-cap-location: top

rev_long <- pivot_longer(ff_rev, rev_02:rev_78, names_to = c("type","Category"), values_to = "Dollars", names_sep = "_") %>% 
  rename(Year = fy) %>%
  mutate(Category_name = case_when(
    Category == "02" ~ "INDIVIDUAL INCOME TAXES" ,
    Category == "03" ~ "CORPORATE INCOME TAXES" ,
    Category == "06" ~ "SALES TAXES" ,
    Category == "09" ~ "MOTOR FUEL TAX" ,
    Category == "12" ~ "PUBLIC UTILITY TAXES" ,
    Category == "15" ~ "CIGARETTE TAXES" ,
    Category == "18" ~ "LIQUOR GALLONAGE TAXES" ,
    Category == "21" ~ "INHERITANCE TAX" ,
    Category == "24" ~ "INSURANCE TAXES&FEES&LICENSES" ,
    Category == "27" ~ "CORP FRANCHISE TAXES & FEES" ,
    Category == "30" ~ "HORSE RACING TAXES & FEES",  # in Other
    Category == "31" ~ "MEDICAL PROVIDER ASSESSMENTS" ,
    Category == "32" ~ "GARNISHMENT-LEVIES" , # dropped
    Category == "33" ~  "LOTTERY RECEIPTS" ,
    Category == "35" ~  "OTHER TAXES" ,
    Category == "36" ~  "RECEIPTS FROM REVENUE PRODUCING", 
    Category == "39" ~  "LICENSES, FEES & REGISTRATIONS" ,
    Category == "42" ~  "MOTOR VEHICLE AND OPERATORS" ,
    Category == "45" ~  "STUDENT FEES-UNIVERSITIES",   # dropped
    Category == "48" ~  "RIVERBOAT WAGERING TAXES" ,
    Category == "51" ~  "RETIREMENT CONTRIBUTIONS" , # dropped
    Category == "54" ~ "GIFTS AND BEQUESTS", 
    Category == "57" ~  "FEDERAL OTHER" ,
    Category == "58" ~  "FEDERAL MEDICAID", 
    Category == "59" ~  "FEDERAL TRANSPORTATION" ,
    Category == "60" ~  "OTHER GRANTS AND CONTRACTS", #other
    Category == "63" ~  "INVESTMENT INCOME", # other
    Category == "66" ~ "PROCEEDS,INVESTMENT MATURITIES" , #dropped
    Category == "72" ~ "BOND ISSUE PROCEEDS",  #dropped
    Category == "75" ~  "INTER-AGENCY RECEIPTS ",  #dropped
    Category == "76" ~  "TRANSFER IN FROM OUT FUNDS",  #other
    Category == "78" ~  "ALL OTHER SOURCES" ,
    Category == "79" ~   "COOK COUNTY IGT", #dropped
    Category == "98" ~  "PRIOR YEAR REFUNDS", #dropped
    T ~ "Check Me!"
    
  ) )%>% 
  mutate(Category_name = str_to_title(Category_name))


exp_long <- pivot_longer(ff_exp, exp_402:exp_970 , names_to = c("type", "Category"), values_to = "Dollars", names_sep = "_") %>% 
  rename(Year = fy ) %>% 
  mutate(Category_name = 
           case_when(
             
             Category == "131" ~ "GARS",
             Category == "275" ~ "JRS",
             Category == "402" ~ "AGING" ,
             Category == "406" ~ "AGRICULTURE", 
             Category == "416" ~ "CENTRAL MANAGEMENT",
             Category == "418" ~ "CHILDREN AND FAMILY SERVICES", 
             Category == "420" ~ "COMMERCE AND ECONOMIC OPPORTUNITY",
             Category == "422" ~ "NATURAL RESOURCES" ,
             Category == "426" ~ "CORRECTIONS",
             Category == "427" ~ "EMPLOYMENT SECURITY" ,
             
             Category == "442" ~ "Human Rights" ,
             
             Category == "444" ~ "Human Services" ,
             Category == "445" ~ "IL Power Agency" ,
             Category == "448" ~ "Innovation and Technology", # AWM added fy2022
             
             Category == "452" ~ "Labor" ,
             
             Category == "458" ~ "State Lottery" ,
             
             Category == "478" ~ "FAMILY SERVICES net Medicaid",
             Category == "482" ~ "PUBLIC HEALTH", 
             Category == "492" ~ "REVENUE", 
             Category == "493" ~  "Teacher Retirmeent System (TRS)",
             Category == "494" ~ "TRANSPORTATION" ,
          
             Category == "489" ~ "SERS",
             Category == "507" ~ "GOMB",
             Category == "497" ~ "VETERNS' AFFAIRS" ,
             Category == "532" ~ "ENVIRONMENTAL PROTECT AGENCY" ,
             Category == "557" ~ "IL STATE TOLL HIGHWAY AUTH" ,
             Category == "589" ~ "State Employment Retirement System (SERS)",
             Category == "684" ~ "IL COMMUNITY COLLEGE BOARD", 
             Category == "691" ~ "IL STUDENT ASSISTANCE COMM" ,
             Category == "693" ~ "SURS",
             Category == "900" ~ "NOT IN FRAME",
             Category == "901" ~ "STATE PENSION CONTRIBUTION",
             Category == "903" ~ "DEBT SERVICE",
             Category == "904" ~ "State Employee Healthcare",
             Category == "910" ~ "LEGISLATIVE"  ,
             Category == "920" ~ "JUDICIAL" ,
             Category == "930" ~ "ELECTED OFFICERS" , 
             Category == "940" ~ "OTHER HEALTH-RELATED", 
             Category == "941" ~ "PUBLIC SAFETY" ,
             Category == "942" ~ "ECON DEVT & INFRASTRUCTURE" ,
             Category == "943" ~ "CENTRAL SERVICES",
             Category == "944" ~ "BUS & PROFESSION REGULATION" ,
             Category == "945" ~ "MEDICAID" ,
             Category == "946" ~ "CAPITAL IMPROVEMENT" , 
             Category == "948" ~ "OTHER DEPARTMENTS" ,
             Category == "949" ~ "OTHER BOARDS & COMMISSIONS" ,
             Category == "959" ~ "K-12 EDUCATION" ,
             Category == "960" ~ "UNIVERSITY EDUCATION",
             Category == "970" ~ "Local Govt Transfers",
             T ~ "CHECK ME!")
           ) %>% 
  mutate(Category_name = str_to_title(Category_name))

write_csv(exp_long, paste0("data/FY", current_year, " Files/expenditures_recoded_long_pensionchange_FY", current_year, ".csv"))

write_csv(rev_long, paste0("data/FY", current_year, " Files/revenue_recoded_long_pensionchange_FY", current_year, ".csv"))


# combine revenue and expenditures into one data frame
aggregated_totals_long <- rbind(rev_long, exp_long)


aggregated_totals_long %>% mutate(`Dollars (Millions)` = round(Dollars, digits = 0)) %>% select(-Dollars) %>%
  select(Year, Category_name, `Dollars (Millions)`, type, Category)

write_csv(aggregated_totals_long, paste0("data/FY", current_year, " Files/aggregated_totals_pensionchange", current_year, ".csv"))

```


<!--- Oh hey, there were NAs when updating the files for FY25 because I hadn't assigned revenue or expenditure categories to the new funds and sources and it did indeed break! --->


```{r}
#| label: tbl-year-totals-table-withgap
#| tbl-cap-location: top

year_totals <- aggregated_totals_long %>% 
  group_by(type, Year) %>% 
  summarize(Dollars = sum(Dollars, na.rm = TRUE)) %>% 
  pivot_wider(names_from = "type", values_from = Dollars) %>% 

  rename(Expenditures = exp,
         Revenue = rev) %>%  
  mutate(`Fiscal Gap` = Revenue - Expenditures)
# %>%  arrange(desc(Year))
# creates variable for the Gap each year

year_totals  %>%  
  mutate_all(., ~round(., digits = 0)) %>%
  kbl(caption = "Fiscal Gap for each Fiscal Year ($ Millions)") %>% 
  kable_styling(bootstrap_options = c("striped"))  %>%
  kable_classic() %>%   
  add_footnote(c("Values include State CURE dollars (SLFRF)") )

```

Graphs made from `aggregated_totals_long` dataframe.

### Fiscal Gap Graph

```{r}
#| label: fig-fiscal-gap
#| fig-cap: "Fiscal Gap Comparison"
#| fig-subcap: 
#| - "Fiscal Gap With Trend Lines"
#| - "Fiscal Gap Without Trend Lines"
#| fig-cap-location: top

## Adjust x and y coordinates to move placement of text
annotation <- data.frame(
  x = c(2004, 2017, 2019),
  y = c(60, 50, 5),  
  label = c("Expenditures","Revenue", "Fiscal Gap")
)


annotation_nums <- data.frame(
  x = c(2022, 2022, 2023),
  y = c(91, 120, -5),  
  label = c(
   year_totals$Expenditures[year_totals$Year==current_year]/1000,
  year_totals$Revenue[year_totals$Year==current_year]/1000, 
  year_totals$`Fiscal Gap`[year_totals$Year==current_year]/1000)
)

## Dashed line versions for expenditures: 
fiscal_gap <-   
  ggplot(data = year_totals, aes(x=Year, y = Revenue/1000)) +
  geom_recessions(text = FALSE, update = recessions)+

  # geom_smooth adds regression line, graphed first so it appears behind line graph
  geom_smooth(aes(x = Year, y = Revenue/1000), color = "gray", alpha = 0.7, method = "lm", se = FALSE) + 
  #  scale_linetype_manual(values="dashed")+
  geom_smooth(aes(x = Year, y = Expenditures/1000), color ="rosybrown2", linetype = "dotted", method = "lm", se = FALSE, alpha = 0.7) +

  # line graph of revenue and expenditures
  geom_line(aes(x = Year, y = Revenue/1000), color = "Black", size=1) +
  geom_line(aes(x = Year, y = Expenditures/1000, linetype = "dashed"), color = "red", lwd=1) +
  geom_line(aes(x = Year, y = (`Fiscal Gap`/1000)), color = "darkgray", lwd = 1) +
    geom_hline(yintercept = 0) +

  geom_text(data = annotation, aes(x=x, y=y, label=label,
  parse = TRUE))+
  # labels
    theme_classic() +
    theme(legend.position = "bottom", legend.title = element_blank())+
    scale_linetype_manual(values = c("dashed", "dashed")) +
  scale_x_continuous(expand = c(0,0)) +

#  scale_y_continuous(labels = comma)+
  xlab("Year") + 
  ylab("Billions of Dollars")  +
  ggtitle(paste0("Illinois Expenditures and Revenue Totals, 1998-",current_year))

fiscal_gap

# annotation_billions <- data.frame(
#   x = c(2004, 2017, 2019),
#   y = c(60, 50, 5),  
#   label = c("Expenditures","Revenue", "Fiscal Gap"))


fiscal_gap2 <- ggplot(data = year_totals, aes(x=Year, y = Revenue/1000)) +
  geom_recessions(text = FALSE, update_recessions = recessions)+
  geom_line(aes(x = Year, y = Revenue/1000,  color = "Revenue"), lwd = 1, label = "Revenue") +
  geom_line(aes(x = Year, y = Expenditures/1000, color = "Expenditures"),  linetype = "dotted", lwd = 1, label = "Expenditures") +
  geom_line(aes(x = Year, y = (`Fiscal Gap`/1000)), color = "darkgray", lwd=1) +
  
  geom_text(data = annotation, aes(x=x, y=y, label=label)) + ## Word locations and text
  geom_text(data = annotation_nums, aes(x = x, y = y, label = scales::dollar(label, accuracy = 0.01L)), size = 3) +  ## Number locations and text
  theme_classic() +
  theme(legend.position = "bottom", legend.title = element_blank()) +
  scale_color_manual(values = c("Revenue" = "black", "Expenditures" = "red")) + 
  geom_hline(yintercept = 0) +

  scale_y_continuous(#labels = comma, 
                     limits = c(-12, 120), 
                     breaks = c(-10, 20, 40, 60, 80, 100, 120), 
                     minor_breaks = c(-10, 0, 10, 30, 50, 70, 90, 110))+
  scale_x_continuous(expand = c(0,0), 
                     limits = c(1998, current_year)
                     ) +
 # scale_color_manual(values = c("red" = "Expenditures", "black" = "Revenue")) + 

  xlab("Year") + 
  ylab("Billions of Dollars")  +
  ggtitle(paste0("Illinois Expenditures and Revenue Totals, 1998-",current_year))


fiscal_gap2

```

```{r FIGURE1-EXPORT}
#| include: false

fiscal_gap2 <- ggplot(data = year_totals, aes(x=Year, y = Revenue/1000)) +
  geom_recessions(text = FALSE, update = recessions)+
  geom_line(aes(x = Year, y = Revenue/1000, color = "Revenue"), lty = "solid"
                ,  lwd = 1) +
  geom_line(aes(x = Year, y = Expenditures/1000, color = "Expenditures"), lty = "dashed"
                , lwd = .7 ) +
  geom_line(aes(x = Year, y = (`Fiscal Gap`/1000), color = "Fiscal Gap"),  lwd=1) +
  
  geom_text(data = annotation, aes(x=x, y=y, label=label))+
  geom_text(data = annotation_nums, aes(x = x+1, y = y, label = scales::dollar(label, accuracy=0.01L)), size = 3) + 
theme_classic() +
theme(
  legend.position = "bottom",
  legend.title = element_blank(),  
  panel.background = element_rect(fill='transparent'), #transparent panel bg
  plot.background = element_rect(fill='transparent', color=NA), #transparent plot bg
  panel.grid.major = element_blank(), #remove major gridlines
  panel.grid.minor = element_blank(), #remove minor gridlines
  legend.background = element_rect(fill='transparent'), #transparent legend bg
)+
#scale_linetype_manual(values = c( "dashed", "solid", "solid" )) +
#scale_linetype_manual(labels =c("Expenditures", "Revenue")) +
  geom_hline(yintercept = 0) +
  scale_color_manual(values = c( "Expenditures" = "red", "Revenue" = "black", "Fiscal Gap" = "darkgray"),     
                     guide = guide_legend(override.aes = list(linetype = c( "dashed", "solid", "solid")))
   ) + 
  scale_y_continuous(limits = c(-12, 120), 
                     breaks = c(-10, 20, 40, 60, 80, 100, 120))+
  scale_x_continuous(expand = c(0,0), 
                     limits = c(1998, current_year), 
                     breaks = c(1998, 2000, 2005, 2010, 2015, 2020, current_year) ) +
  xlab("Year") + 
  ylab("Billions of Dollars") 

fiscal_gap2


# ggsave(plot = fiscal_gap2, filename= "./paper-figures/Figure1-fiscalgap.eps", width = 6, height = 4)

#  ggsave(plot = fiscal_gap2, filename= "./paper-figures/Figure1-fiscalgap.png", width = 6, height = 4)

```

**Expenditure and revenue amounts in billions of dollars:**

```{r}
#| label: fig-bar-graphs-fy23
#| fig-cap: FY23 Totals
#| fig-subcap: 
#| - "FY24 Expenditures"
#| - "FY24 Revenue Sources"

exp_long %>%
  filter(Year == current_year) %>%
  arrange(desc(`Dollars`)) %>%
  ggplot() + 
  geom_col(aes(x = fct_reorder(Category_name, `Dollars`), y = (`Dollars`/1000), fill = "red"))+ 
  coord_flip() +
      theme_classic()+
  theme(legend.position = "none") +
      labs(title = paste0("Expenditures for ", current_year))+    xlab("Expenditure Categories") +
  ylab("Billions of Dollars") 

rev_long %>%
  filter(Year == current_year) %>%
  arrange(desc(`Dollars`)) %>%
  ggplot() + 
  geom_col(aes(x = fct_reorder(Category_name, `Dollars`), y = (`Dollars`/1000)))+ 
  coord_flip() +
    theme_classic() +
    theme(legend.position = "none") +
      labs(title = paste0("Revenue for ", current_year))+
    xlab("Revenue Categories") +
  ylab("Billions of Dollars") 

```

**Expenditure and revenues when focusing on largest categories and combining others into "All Other Expenditures(Revenues)":**

```{r}
#| label: fig-bargraphs-largest-expandrev-cats
#| fig-cap: 
#| fig-subcap: 
#| - "Largest Expenditures for FY2024"
#| - "Largest Revenue Sources for FY2024"

exp_long %>%
  filter( Year == current_year) %>%
  mutate(rank = rank(Dollars),
        Category_name = ifelse(rank > 13, Category_name, 'All Other Expenditures')) %>%
 # select(-c(Year, Dollars, rank)) %>%
  arrange(desc(Dollars)) %>%
  ggplot() + 
  geom_col(aes(x = fct_reorder(Category_name, `Dollars`), y = `Dollars`/1000), fill = "rosybrown2") + 
  coord_flip() +
      theme_classic() +
      labs(title = paste0("Expenditures for ", current_year))+    xlab("") +
  ylab("Billions of Dollars")

rev_long %>%
  filter( Year == current_year) %>%
  mutate(rank = rank(Dollars),
        Category_name = ifelse(rank > 10, Category_name, 'All Other Sources')) %>%
  arrange(desc(Dollars)) %>%
  ggplot() + 
  geom_col(aes(x = fct_reorder(Category_name, `Dollars`/1000), y = `Dollars`/1000), fill = "dark gray")+ 
  coord_flip() +
      theme_classic() +
      labs(title = paste0("Revenue for ", current_year)) +  
  xlab("") +
  ylab("Billions of Dollars")
```


### Top 3 Revenues {#sec-top-3-revenues}

```{r}
#| label: fig-top3-revenues
#| fig-cap: Top 3 Revenue Sources (Own-Source Revenues only)
#| fig-cap-location: top

annotation <- data.frame(
  x = c(2013, 2018, 2013),
  y = c(16, 10, 5),  
  label = c("Individual Income Tax", "Sales Tax", "Corporate Income Tax")
)

top3 <- rev_long  %>% 
  filter(Category == "02" | Category == "03" | Category == "06")

top3 <- ggplot(data = top3, aes(x=Year, y=Dollars/1000))+
      geom_recessions(text = FALSE, update = recessions)+
  geom_line(aes(x=Year, y=Dollars/1000, color = Category_name)) + 
  geom_text(data = annotation, aes(x=x, y=y, label=label)) +
  theme_classic() +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(labels = comma) +
  scale_linetype_manual(values = c("dotted", "dashed", "solid")) +
  theme(legend.position = "none") +
  labs(title = "Top 3 Own Source Revenues", 
       subtitle = "Individual Income Taxes, Sales Tax, and Corporate income taxes",
       y = "Billions of Nominal Dollars") 
  
top3

```



### Own Source and Fed Transfers {#sec-own-source-and-fed-transfers}

```{r}
#| label: tbl-ownsource-and-fed-revenue
#| tbl-cap: "Own Source and Federal Revenue"
#| tbl-location: margin
 

ownsource_rev <- rev_long %>%
  filter(!Category %in% c("57", "58", "59")) %>%
  group_by(Year) %>% 
  summarize(Dollars = sum(Dollars))

ownsource_rev
# ownsource_rev %>% 
#   ggplot()+geom_line(aes(x=Year, y=Dollars)) + 
#   labs(title = "Own Source Revenues", subtitle = "Total own source revenue", y = "Millions of Dollars")

```

```{r}
fed_rev <- ff_rev %>% 
  select(fy, rev_57, rev_58, rev_59) %>%
  mutate(fed_total = rev_57+rev_58+rev_59)


annotation <- data.frame(
  x = c(2013, 2015),
  y = c(50, 25),  
  label = c("Own Source Revenue", "Federal Revenue")
)

annotation_nums <- data.frame(
  x = c(2023, 2023),
  y = c(50, 25),  
  label = c("Own Source Revenue", "Federal Revenue")
)

ownsource_graph <- ggplot(ownsource_rev, aes(x=Year, y=Dollars/1000)) + 
  geom_recessions(text = FALSE, update_recessions = recessions)+
  geom_line(data = ownsource_rev, aes(x=Year, y=Dollars/1000), color = "Red") + 
  geom_line(data = fed_rev, aes(x=fy, y=fed_total/1000), color = "Black") + 
    geom_text(data = annotation, aes(x=x, y=y, label=label))+
    scale_y_continuous(labels = comma)+
  scale_x_continuous(expand = c(0,0)) + 
  theme(legend.position = "none") +

  theme_classic()+
  labs(title = "Own Source Revenue and Federal Revenue, 1998-2023", 
  y = "Billions of Dollars")
ownsource_graph

# ggsave(plot = ownsource_graph, file = "Figure4.eps")

```

```{r FIGURE2-EXPORT, include = FALSE}

# Same as above but doesn't have title. 
# Made for fy23 report and design team.


annotation_nums <- data.frame(
  x = c(2022, 2022.5),
  y = c(80, 30),  
  label = c("$?", "$?")
)


ownsource_graph <- ggplot(ownsource_rev, aes(x=Year, y=Dollars/1000)) + 
  geom_recessions(text = FALSE, update = recessions)+
  geom_line(data = ownsource_rev, aes(x=Year, y=Dollars/1000), color = "Red") + 
  geom_line(data = fed_rev, aes(x=fy, y=fed_total/1000), color = "Black") + 
    geom_text(data = annotation, aes(x=x, y=y, label=label))+
    geom_text(data = annotation_nums, aes(x=x, y=y, label=label))+
    scale_y_continuous(labels = comma)+
  scale_x_continuous(expand = c(0,0), limits = c(1998, 2024)) + 
  theme_classic()+
  theme(legend.position = "none",
    panel.background = element_rect(fill='transparent'), #transparent panel bg
    plot.background = element_rect(fill='transparent', color=NA), #transparent plot bg
    panel.grid.major = element_blank(), #remove major gridlines
    panel.grid.minor = element_blank(), #remove minor gridlines
    legend.background = element_rect(fill='transparent'), #transparent legend bg
  )+
  labs(
    #title = "Own Source Revenue and Federal Revenue, 1998-2023", 
  y = "Billions of Dollars")
ownsource_graph

# ggsave(plot = ownsource_graph, file = "./paper-figures/Figure2.eps", limitsize=FALSE)
# ggsave(plot = ownsource_graph, file = "./paper-figures/Figure2.png", limitsize = FALSE)

```

# Change from Previous Year

Each year, you will need to update the CAGR formulas! Change the filter() year.

`calc_cagr` is a function created for calculating the CAGRs for different spans of time.

```{r}
exp_totals <- ff_exp %>% 
  rowwise() %>% 
  mutate(exp_TOTALS = sum(across(exp_402:exp_970), na.rm=TRUE))

rev_totals <- ff_rev %>%    
  rowwise() %>% 
  mutate(rev_TOTALS = sum(across(rev_02:rev_78), na.rm=TRUE))



rev_long <- pivot_longer(rev_totals, rev_02:rev_TOTALS, names_to = c("type","Category"), values_to = "Dollars", names_sep = "_") %>% 
  rename(Year = fy) %>%
  mutate(Category_name = case_when(
    Category == "02" ~ "INDIVIDUAL INCOME TAXES" ,
    Category == "03" ~ "CORPORATE INCOME TAXES" ,
    Category == "06" ~ "SALES TAXES" ,
    Category == "09" ~ "MOTOR FUEL TAX" ,
    Category == "12" ~ "PUBLIC UTILITY TAXES" ,
    Category == "15" ~ "CIGARETTE TAXES" ,
    Category == "18" ~ "LIQUOR GALLONAGE TAXES" ,
    Category == "21" ~ "INHERITANCE TAX" ,
    Category == "24" ~ "INSURANCE TAXES&FEES&LICENSES" ,
    Category == "27" ~ "CORP FRANCHISE TAXES & FEES" ,
    Category == "30" ~ "HORSE RACING TAXES & FEES",  # in Other
    Category == "31" ~ "MEDICAL PROVIDER ASSESSMENTS" ,
    Category == "32" ~ "GARNISHMENT-LEVIES" , # dropped
    Category == "33" ~  "LOTTERY RECEIPTS" ,
    Category == "35" ~  "OTHER TAXES" ,
    Category == "36" ~  "RECEIPTS FROM REVENUE PRODUCING", 
    Category == "39" ~  "LICENSES, FEES & REGISTRATIONS" ,
    Category == "42" ~  "MOTOR VEHICLE AND OPERATORS" ,
    Category == "45" ~  "STUDENT FEES-UNIVERSITIES",   # dropped
    Category == "48" ~  "RIVERBOAT WAGERING TAXES" ,
    Category == "51" ~  "RETIREMENT CONTRIBUTIONS" , # dropped
    Category == "54" ~ "GIFTS AND BEQUESTS", 
    Category == "57" ~  "FEDERAL OTHER" ,
    Category == "58" ~  "FEDERAL MEDICAID", 
    Category == "59" ~  "FEDERAL TRANSPORTATION" ,
    Category == "60" ~  "OTHER GRANTS AND CONTRACTS", #other
    Category == "63" ~  "INVESTMENT INCOME", # other
    Category == "66" ~ "PROCEEDS,INVESTMENT MATURITIES" , #dropped
    Category == "72" ~ "BOND ISSUE PROCEEDS",  #dropped
    Category == "75" ~  "INTER-AGENCY RECEIPTS ",  #dropped
    Category == "76" ~  "TRANSFER IN FROM OUT FUNDS",  #other
    Category == "78" ~  "ALL OTHER SOURCES" ,
    Category == "79" ~   "COOK COUNTY IGT", #dropped
    Category == "98" ~  "PRIOR YEAR REFUNDS", #dropped
    Category == "TOTALS" ~ "Total",
    T ~ "CHECK ME"
    
  ) ) %>% 
  select(-type, -Category) %>%  # drop extra columns type and Category number
  group_by(Year, Category_name) %>%
  summarise(Dollars= round(sum(Dollars, na.rm=TRUE), digits=2)) %>% 
  mutate(Category_name = str_to_title(Category_name))

# creates wide version of table where each revenue source is a column
revenue_wide2 <- rev_long %>% 
  pivot_wider(names_from = Category_name, 
              values_from = Dollars) %>%
  #  relocate("Other Revenue Sources **", .after = last_col()) %>%
  relocate("Total", .after =  last_col())
```


```{r}

exp_long <- pivot_longer(exp_totals, exp_402:exp_TOTALS , names_to = c("type", "Category"), values_to = "Dollars", names_sep = "_") %>%

  rename(Year = fy ) %>% 
  mutate(Category_name = 
           case_when(
             Category == "402" ~ "AGING" ,
             Category == "406" ~ "AGRICULTURE", 
             Category == "416" ~ "Central Management",
             Category == "418" ~ "CHILDREN AND FAMILY SERVICES", 
             Category == "420" ~ "Community Development",
             Category == "422" ~ "NATURAL RESOURCES" ,
             Category == "426" ~ "CORRECTIONS",
             Category == "427" ~ "EMPLOYMENT SECURITY" ,
             Category == "442" ~ "Human Rights" ,
             
             Category == "444" ~ "Human Services" ,
             Category == "445" ~ "IL Power Agency" ,
             Category == "452" ~ "Labor" ,
             Category == "458" ~ "State Lottery" ,
             
             Category == "478" ~ "HEALTHCARE & FAM SER NET OF MEDICAID", 
             Category == "482" ~ "PUBLIC HEALTH", 
             Category == "492" ~ "REVENUE", 
             Category == "494" ~ "Transportation" ,
             Category == "497" ~ "Veterans' Affairs" ,
             Category == "507" ~ "GOMB",

             Category == "532" ~ "ENVIRONMENTAL PROTECT AGENCY" ,
             Category == "557" ~ "Tollway" ,
             Category == "901" ~ "State Pension Contribution",
             Category == "903" ~ "Debt Service",
             Category == "904" ~ "State Employee Healthcare",
             Category == "910" ~ "LEGISLATIVE"  ,
             Category == "920" ~ "JUDICIAL" ,
             Category == "930" ~ "ELECTED OFFICERS" , 
             Category == "941" ~ "Public Safety" ,
             Category == "942" ~ "ECON DEVT & INFRASTRUCTURE" ,
             Category == "943" ~ "CENTRAL SERVICES",
             Category == "944" ~ "BUS & PROFESSION REGULATION" ,
             Category == "945" ~ "Medicaid" ,
             Category == "946" ~ "Capital Improvement" , 
             Category == "948" ~ "OTHER DEPARTMENTS" ,
             Category == "949" ~ "OTHER BOARDS & COMMISSIONS" ,
             Category == "959" ~ "K-12 Education" ,
             Category == "960" ~ "UNIVERSITY EDUCATION",
             Category == "970" ~ "Local Govt Revenue Sharing",
             Category == "TOTALS" ~ "Total",
             T ~ "CHECK ME" 
            #  T ~ "All Other Expenditures **")
           )) %>% 
    select(-type, -Category) %>% 

   # mutate(Category_name = ifelse(Dollars < 1, "Other Departments", Category_name)) |>

  group_by(Year, Category_name) %>% 
  summarise(Dollars = round(sum(Dollars, na.rm=TRUE), digits=2)) %>% 
  mutate(Category_name = str_to_title(Category_name))

expenditure_wide2 <- exp_long %>% 
  pivot_wider(names_from = Category_name, 
              values_from = Dollars) %>%
  relocate("Total", .after =  last_col())



```

Things to do when updating the code:

-   Each year, you need to increase the cagr value by 1. The value should be the (current year - 1998). For FY23, this is 2023-1998 = 25. So all cagr values that were 24 will be changed to 25.


```{r}


# function for calculating the CAGR
calc_cagr <- function(df, n) {
  df <- df %>%
    arrange(Category_name, Year) %>%
    group_by(Category_name) %>%
    mutate(cagr = ((`Dollars` / lag(`Dollars`, n)) ^ (1 / n)) - 1,
           cagr = ifelse(is.na(cagr), 0, cagr)
)

  return(df)
}

cagr_calculations <- function(df){ # This works for one variable at a time
  
  df <- df
  
  cagr_max <- calc_cagr(df, 26) %>% 
    summarize(cagr_max = round(sum(cagr*100, na.rm = TRUE), 2))

  # Update year in the filter() and summarize() commands to current year.
  cagr_10 <- calc_cagr(df, 10) %>% 
    filter(Year == current_year) %>%
    summarize(cagr_10 = case_when(Year == current_year ~ round(sum(cagr*100, na.rm = TRUE), 2)))

  cagr_5 <- calc_cagr(df, 5) %>% 
    filter(Year == current_year) %>%
    summarize(cagr_5 = case_when(Year == current_year ~ round(sum(cagr*100, na.rm = TRUE), 2)))

  cagr_3 <- calc_cagr(df, 3) %>% 
    filter(Year == current_year) %>%
    summarize(cagr_3 = case_when(Year == current_year ~ round(sum(cagr*100, na.rm = TRUE), 2)))

  cagr_2 <- calc_cagr(df, 2) %>% 
    filter(Year == current_year) %>%
    summarize(cagr_2 = case_when(Year == current_year ~ round(sum(cagr*100, na.rm = TRUE), 2)))

  cagr_1 <- calc_cagr(df, 1) %>% 
    filter(Year == current_year) %>%
    summarize(cagr_1 = case_when(Year == current_year ~ round(sum(cagr*100, na.rm = TRUE), 2)))
  
    # Combine all into one tibble
  result <- data.frame(cagr_max, cagr_10, cagr_5, cagr_3, cagr_2, cagr_1)
  return(result)
  }
```


```{r }
#| label: tbl-good-exp-CAGR-withTotals
#| tbl-cap: Expenditure Category CAGRs with Total CAGR (Ordered Alphabetically)
#| tbl-cap-location: top


CAGR_expenditures_summary_tot <- cagr_calculations(exp_long)   |>
  select(-c(Category_name.1, Category_name.2, Category_name.3, Category_name.4, Category_name.5 )) %>% 
  rename("Expenditure Category" = Category_name, "1 Year CAGR" = cagr_1, "2 Year CAGR" = cagr_2, "3 Year CAGR" = cagr_3, "5 Year CAGR" = cagr_5, "10 Year CAGR" = cagr_10, "26 Year CAGR" = cagr_max )


totalrow <- which(grepl("Total", CAGR_expenditures_summary_tot$`Expenditure Category`))

CAGR_expenditures_summary_tot <- move_to_last(CAGR_expenditures_summary_tot, totalrow) 

lastrow = nrow(CAGR_expenditures_summary_tot)


CAGR_expenditures_summary_tot %>%   
  kbl(caption = "CAGR Calculations for All Expenditure Categories" , row.names=FALSE) %>% 
  kable_classic() %>%
  row_spec(lastrow, bold = T, color = "black", background = "gray")
```

```{r }
#| label: tbl-RevCAGRs-allcats
#| tbl-cap: Revenue Category CAGRs with Total CAGR (Ordered Alphabetically)
#| tbl-cap-location: top

CAGR_revenue_summary_tot <- cagr_calculations(rev_long) %>%   
  select(-c(Category_name.1, Category_name.2, Category_name.3, Category_name.4, Category_name.5 )) %>% 
  rename("Revenue Category" = Category_name, "1 Year CAGR" = cagr_1, "2 Year CAGR" = cagr_2, "3 Year CAGR" = cagr_3, "5 Year CAGR" = cagr_5, "10 Year CAGR" = cagr_10, "26 Year CAGR" = cagr_max )

CAGR_revenue_summary_tot <- move_to_last(CAGR_revenue_summary_tot, 1)

totalrow <- which(grepl("Total", CAGR_revenue_summary_tot$`Revenue Category`))

CAGR_revenue_summary_tot <- move_to_last(CAGR_revenue_summary_tot, totalrow)

lastrow = nrow(CAGR_revenue_summary_tot)

CAGR_revenue_summary_tot %>% 
  kbl(caption = "CAGR Calculations for All Revenue Sources (Ordered Alphabetical)", row.names = FALSE) %>% 
   kable_classic() %>%
    row_spec(lastrow, bold = T, color = "black", background = "gray")
```


```{r}
#| label: tbl-AppendixItem1
#| tbl-cap: "Yearly Change in Revenues - All FF Categories, Ordered from Largest to Smallest Revenue Amount"


first_year = as.numeric(1998)
n_year_change = as.numeric(current_year-1998)


revenue_change2 <- rev_long %>%
  filter(Year >= past_year | Year == first_year) %>%
  pivot_wider(names_from = Year , values_from = Dollars,   names_prefix = "Dollars_") %>%
  mutate(
    "Current FY ($ billions)" = round(Dollars_2025/1000, digits = 2),
    "Past FY ($ billions)" = round(Dollars_2024/1000, digits = 2),
        "FY 1994 ($ billions)" = round(Dollars_1998/1000, digits = 2),
    "1-Year Change" = percent((Dollars_2025 -Dollars_2024)/Dollars_2024, accuracy = .01)) |>
  left_join(CAGR_revenue_summary_tot, by = c("Category_name" = "Revenue Category")) %>% 
    arrange(-`Current FY ($ billions)`)%>%
    mutate(`26 Year CAGR` = percent(`26 Year CAGR`/100, accuracy=.01)) %>%

  rename(  
          "Revenue Category" = Category_name ) %>%
  select(-c( Dollars_1998, Dollars_2025, Dollars_2024, `1 Year CAGR`:`10 Year CAGR`))


allother_row <- which(grepl("All Other", revenue_change2$`Revenue Category`))

revenue_change2 <- move_to_last(revenue_change2, allother_row) # Move "All Other" to 2nd to last row

totalrow <- which(grepl("Total", revenue_change2$`Revenue Category`))
revenue_change2 <- move_to_last(revenue_change2, totalrow) # Move "Total" to last row

lastrow = nrow(revenue_change2)

revenue_change2 %>%   
  filter(!is.na(`Revenue Category`)) %>%

  kbl(caption = "Table 1. Yearly Change in Revenue", row.names = FALSE) %>% 
   kable_classic() %>%
    row_spec(lastrow, bold = T, color = "black", background = "gray")
```


```{r}
#| label: tbl-AppendixItem2
#| tbl-cap: "Yearly Change in Expenditures - All FF Categories, Ordered from Largest to Smallest Expenditure Amount"

expenditure_change2 <- exp_long %>%
  group_by(Year, Category_name) |>
  summarize(Dollars = sum(Dollars, na.rm=TRUE)) |>
  ungroup() |>
  filter(Year >= past_year | Year == first_year) %>%
  pivot_wider(names_from = Year , values_from = Dollars,   names_prefix = "Dollars_") %>%
  mutate(
    "FY 2025 ($ billions)" = round(Dollars_2025/1000, digits = 2),
    "FY 2024 ($ billions)" = round(Dollars_2024/1000, digits = 2),
    "FY 1998 ($ billions)" = round(Dollars_1998/1000, digits = 2),
    "1-Year Change" = percent((Dollars_2025 -Dollars_2024)/Dollars_2024, accuracy = .01)) |>
  left_join(CAGR_expenditures_summary_tot, by = c("Category_name" = "Expenditure Category")) %>% 
  arrange(-`FY 2025 ($ billions)`)%>%
    mutate(`26 Year CAGR` = percent(`26 Year CAGR`/100, accuracy=.01)) %>%

  select(-c( Dollars_1998, Dollars_2025, Dollars_2024, `1 Year CAGR`:`10 Year CAGR`)) %>%

  rename("Expenditure Category" = Category_name ) # |> filter(!is.na(`Expenditure Category`))


allother_row <- which(grepl("All Other", expenditure_change2$`Expenditure Category`))

expenditure_change2 <- move_to_last(expenditure_change2, allother_row) # Move "All Other" to 2nd to last row

totalrow <- which(grepl("Total", expenditure_change2$`Expenditure Category`))
expenditure_change2 <- move_to_last(expenditure_change2, totalrow) # Move "Total" to last row

lastrow = nrow(expenditure_change2)

expenditure_change2 %>% 
  kbl(row.names = FALSE) %>% 
  kable_classic() %>%
    row_spec(lastrow, bold = T, color = "black", background = "gray")

           
```

## Summary Tables - Largest Categories

The 10 largest revenue sources and 15 largest expenditure sources remain separate categories and all other smaller sources/expenditures are combined into "All Other Revenues (Expenditures)". These condensed tables are typically used in the Fiscal Futures articles. They were manually created in past years but this hopefully automates the process a bit until final formatting stages.

-   take ff_rev and ff_exp data frames, which were in wide format, pivot them longer and mutate the Category_name variable to nicer labels. Keep largest categories separate and aggregate the rest.
-   You need to manually comment out the categories that are not the largest each year. Check and compare to the previous years largest categories!

```{r }
#| label: tbl-top-rev-CAGRs
#| tbl-cap: Largest Revenue Categories with CAGRs

n_categories <- 10+1 # (Top 10 and then Total )

rev_majorcats <- rev_long |> 
  filter(Year == current_year | Year == first_year
         ) |>
  arrange(desc(Dollars)) |>
  slice(1:n_categories)

rev_long_majorcats <- rev_long |>
  mutate(Category_name = ifelse(Category_name %in% rev_majorcats$Category_name, Category_name, "All Other Sources"),
         Category_name = ifelse(Category_name == "Total", "Total Revenue", Category_name)) |>
  group_by(Year, Category_name) |> 
  summarize(Dollars = sum(Dollars, na.rm=TRUE))


# creates wide version of table where each revenue source is a column
revenue_wide_majorcats <- rev_long_majorcats %>% 
  pivot_wider(names_from = Category_name, 
              values_from = Dollars) %>%
  relocate("All Other Sources", .after = last_col()) %>%
  relocate("Total Revenue", .after =  last_col())
```

```{r}
#| label: tbl-top10-rev-CAGRs
#| tbl-cap: Top 10 Revenue Sources with CAGRs
#| tbl-cap-location: top


CAGR_revenue_majorcats_tot <- cagr_calculations(rev_long_majorcats) %>%   
  select(-c(Category_name.1, Category_name.2, Category_name.3, Category_name.4, Category_name.5 )) %>% 
  rename("Revenue Category" = Category_name, "1 Year CAGR" = cagr_1, "2 Year CAGR" = cagr_2, "3 Year CAGR" = cagr_3, "5 Year CAGR" = cagr_5, "10 Year CAGR" = cagr_10, "26-Year CAGR" = cagr_max )

allother_row <- which(grepl("All Other", CAGR_revenue_majorcats_tot$`Revenue Category`))

CAGR_revenue_majorcats_tot <- move_to_last(CAGR_revenue_majorcats_tot, allother_row) # Move "All Other" to 2nd to last row

totalrow <- which(grepl("Total", CAGR_revenue_majorcats_tot$`Revenue Category`))
CAGR_revenue_majorcats_tot <- move_to_last(CAGR_revenue_majorcats_tot, totalrow) # Move "Total" to last row

lastrow = nrow(CAGR_revenue_majorcats_tot)

CAGR_revenue_majorcats_tot %>% 
  kbl(caption = "CAGR Calculations for Largest Revenue Sources", row.names = FALSE) %>% 
  kable_classic() %>%
    row_spec(lastrow, bold = T, color = "black", background = "gray")
```

```{r }
#| label: tbl-top10-rev-yearlychange
#| tbl-cap: Top 10 Revenue Sources with CAGRs
#| tbl-cap-location: top

###### Yearly change summary table for Top 10 Revenues #####
revenue_change_majorcats <- rev_long_majorcats %>%
  #select(-c(Category)) %>%
  filter(Year >= past_year |  Year == first_year) %>%
  pivot_wider(names_from = Year , values_from = Dollars,   names_prefix = "Dollars_") %>%
  mutate(
    "FY 2025 ($ billions)" = round(Dollars_2025/1000, digits = 2),
    "FY 2024 ($ billions)" = round(Dollars_2024/1000, digits = 2),
    "FY 1998 ($ billions)" = round(Dollars_1998/1000, digits = 2),
    "1-Year Change" = percent((Dollars_2025 -Dollars_2024)/Dollars_2024, accuracy = .01),
  )  %>%
  left_join(CAGR_revenue_majorcats_tot, by = c("Category_name" = "Revenue Category")
            ) %>% 
  arrange(-`FY 2024 ($ billions)`)%>%
  mutate(`26-Year CAGR` = percent(`26-Year CAGR`/100, accuracy=.01)) %>%
  select(-c(Dollars_1998, Dollars_2025, Dollars_2024, `1 Year CAGR`:`10 Year CAGR`
            )) %>%
  rename("Revenue Category" = Category_name )

allother_row <- which(grepl("All Other", revenue_change_majorcats$`Revenue Category`))

revenue_change_majorcats <- move_to_last(revenue_change_majorcats, allother_row) # Move "All Other" to 2nd to last row

totalrow <- which(grepl("Total", revenue_change_majorcats$`Revenue Category`))
revenue_change_majorcats <- move_to_last(revenue_change_majorcats, totalrow) # Move "Total" to last row

lastrow = nrow(revenue_change_majorcats)

revenue_change_majorcats%>% 
  kbl(caption = "Yearly Change in Revenue for Main Revenue Sources", row.names = FALSE, align = "l") %>% 
   kable_classic() %>%
    row_spec(lastrow, bold = T, color = "black", background = "gray")

```


```{r }
#| label: tbl-top-exp-CAGRs
#| tbl-cap: Largest Expenditure Categories with CAGRs

n_categories <- 9 + 1 # (Top 9 and then Total )

exp_majorcats <- exp_long |> 
  filter(Year == current_year | Year == first_year
         ) |>
  arrange(desc(Dollars)) |>
  slice(1:n_categories) # keep top 10 largest categories or categories larger than 2 billion for final table in report (not a set rule, changes each year depending what the focus of the report is or what is highlighted.)


exp_long_majorcats <- exp_long |>
  mutate(Category_name = ifelse(Category_name %in% exp_majorcats$Category_name, Category_name, "All Other Expenditures **"),
         Category_name = ifelse(Category_name == "Total", "Total Expenditures", Category_name)) |>
  group_by(Year, Category_name) |> 
  summarize(Dollars = sum(Dollars, na.rm=TRUE))

# expenditure_wide_majorcats <- exp_long_majorcats %>% 
#   pivot_wider(names_from = Category_name, 
#               values_from = Dollars) %>%
#   relocate("All Other Expenditures **", .after = last_col()) %>%
#   relocate("Total Expenditures", .after =  last_col())



# CAGR values for largest expenditure categories and combined All Other Expenditures


CAGR_expenditures_majorcats_tot <- cagr_calculations(exp_long_majorcats) |>
  select(-c(Category_name.1, Category_name.2, Category_name.3, Category_name.4, Category_name.5 )) %>% 
  rename("Expenditure Category" = Category_name, "1 Year CAGR" = cagr_1, "2 Year CAGR" = cagr_2, "3 Year CAGR" = cagr_3, "5 Year CAGR" = cagr_5, "10 Year CAGR" = cagr_10,"26-Year CAGR" = cagr_max )

allother_row <- which(grepl("Other", CAGR_expenditures_majorcats_tot$`Expenditure Category`))

CAGR_expenditures_majorcats_tot <- move_to_last(CAGR_expenditures_majorcats_tot, allother_row) # Move "All Other" to 2nd to last row

totalrow <- which(grepl("Total", CAGR_expenditures_majorcats_tot$`Expenditure Category`))
CAGR_expenditures_majorcats_tot <- move_to_last(CAGR_expenditures_majorcats_tot, totalrow) # Move "Total" to last row

lastrow = nrow(CAGR_expenditures_majorcats_tot)

CAGR_expenditures_majorcats_tot%>%   
  kbl(caption = "CAGR Calculations for Largest Expenditure Categories" , row.names=FALSE) %>% 
     kable_classic()  %>%
    row_spec(lastrow, bold = T, color = "black", background = "gray")


# Yearly change for Top n largest expenditure categories
expenditure_change_majorcats <- exp_long_majorcats %>%
  filter(Year >= past_year |
           Year == first_year) %>%
  pivot_wider(names_from = Year , values_from = Dollars,   names_prefix = "Dollars_") %>%
  mutate("FY 2025 ($ Billions)" = round(Dollars_2025/1000, digits = 2),
         "FY 2024 ($ Billions)" = round(Dollars_2024/1000, digits = 2),
         "FY 1998 ($ Billions)" = round(Dollars_1998/1000, digits = 2),
         "1-Year Change" = percent((Dollars_2025 -Dollars_2024)/Dollars_2024, accuracy = .01),
         )  %>%
 left_join(CAGR_expenditures_majorcats_tot, by = c("Category_name" = "Expenditure Category")) %>% 
  arrange(-`FY 2025 ($ Billions)`)%>%
  mutate(`26-Year CAGR` = percent(`26-Year CAGR`/100, accuracy=.01)) %>%
  select(-c(Dollars_1998, Dollars_2025, Dollars_2024, `1 Year CAGR`:`10 Year CAGR`
            )) %>%
  rename(
   # "1-Year Change" = `1 Year CAGR`,
    "26-Year Change" = `26-Year CAGR`, 
          "Expenditure Category" = Category_name )

allother_row <- which(grepl("All Other", expenditure_change_majorcats$`Expenditure Category`))

expenditure_change_majorcats <- move_to_last(expenditure_change_majorcats, allother_row) # Move "All Other" to 2nd to last row

totalrow <- which(grepl("Total", expenditure_change_majorcats$`Expenditure Category`))
expenditure_change_majorcats <- move_to_last(expenditure_change_majorcats, totalrow) # Move "Total" to last row

lastrow = nrow(expenditure_change_majorcats)

expenditure_change_majorcats %>% 
  kbl(caption = "Yearly Change in Expenditures", row.names = FALSE, align = "l") %>% 
  kable_classic() %>%
    row_spec(lastrow, bold = T, color = "black", background = "gray")

```



Export summary file with Totals

```{r}
#install.packages("openxlsx")
library(openxlsx)

dataset_names <- list(#'Aggregate Revenues' = revenue_wide2, 
                      #'Aggregate Expenditures' = expenditure_wide2, 


                      
                      'Table 1' = revenue_change_majorcats, #Top categories with yearly change, 23 yr cagr
                      'Table 2' = expenditure_change_majorcats,
                      
                      'Appendix 1' = revenue_change2,
                      'Appendix 2' = expenditure_change2,
                      
                      'CAGR Rev-MajorCats' = CAGR_revenue_majorcats_tot, # Categories Match Table 1 in paper
                      'CAGR Exp-MajorCats' = CAGR_expenditures_majorcats_tot,  
                                            
                      'Fiscal Gap' = year_totals,    # Total Revenue, Expenditure, and Fiscal gap per year
                      
                      'aggregated_totals_long' = aggregated_totals_long # all data in long format. Good for creating pivot tables in Excel
                      )

write.xlsx(dataset_names, file = paste0("data/FY", current_year, " Files/summary_file_FY25_PensionChange_2025_07_016.xlsx"))
```

Saves main items in one excel file named `summary_file.xlsx`. Delete `eval=FALSE` to run on local computer.
