# Federal COVID Money Received

```{r warning=FALSE, message=FALSE, echo=FALSE}
knitr::opts_chunk$set(warning=FALSE, message = FALSE)
library(tidyverse)
library(haven)

library(lubridate)
library(smooth)
library(forecast)
library(scales)

library(ggplot2)
library(readxl)
library(tidyverse)
library(data.table)
library(quantmod)
library(geofacet)
library(fredr)
library(sf)
library(usmap)
library(janitor)

library(RColorBrewer)
library(networkD3)
library(ggsankey)
library(ggalluvial)
```

```{r}

rev_temp <- read_csv("rev_temp.csv")
fedrev<- rev_temp %>% 
  filter(rev_type == "58" | rev_type == "59" | rev_type == "57") 

fedrev %>% 
  group_by(fy) %>% 
  summarise(receipts = sum(receipts, na.rm = TRUE)/1000000) %>% 
  ggplot() +
  geom_line(aes(x=fy, y=receipts)) +
      theme_bw() +
  scale_y_continuous(labels = comma)+
  labs(title = "All Federal Revenue", 
       y = "Millions of Dollars", x = "") + 
  theme(legend.position = "bottom", legend.title = element_blank()  )+
    scale_y_continuous(limits = c(0,45000))


fedrev %>% 
  filter(source_name_AWM != "FEDERAL STIMULUS PACKAGE") %>%
  group_by(fy) %>% 
  summarise(receipts = sum(receipts, na.rm = TRUE)/1000000) %>% 
  ggplot() +
  geom_line(aes(x=fy, y=receipts)) +
      theme_bw() +
  scale_y_continuous(labels = comma)+
  labs(title = "All Federal EXCEPT Federal Stimulus Package", 
       y = "Millions of Dollars", x = "",
       caption = "Note: Dropping Federal Stimulus Package revenue only removes the $3.5 billion from FY20, $0.3 billion from FY21, and $8.5 billion from FY22. There is still over $12 billion more in Federal Revenue compared to FY19.") + 
  theme(legend.position = "bottom", legend.title = element_blank()  ) +
  scale_y_continuous(limits = c(0,45000))
```

## Revenue Graphs

All graphs have two versions: one without labels and one with labels. This is done so that labels can be added manually in Publisher after the final style is chosen. These labels can then have fund totals added too.

```{r with-DCEO}
sankeydata <- read_csv("sankeyattempt2022.csv")

sankeydata2 <- sankeydata %>% select(Federal, StateFunds, StFund, Expenditures, value, Notes) %>%
  filter(StFund == "Total") %>%
  mutate(StateFunds = factor(StateFunds, levels = c("Total_received_fy21", "Total_received_fy22")),
                  Federal = as.factor(Federal),
                  Expenditures = factor(Expenditures, levels = c("Community Development", "Transit", "Public Health", "Medicaid", "K-12", "Federal Other")),
         Notes = factor(Notes, levels = c("CARES", "CRRSA", "ARPA", "Other *")))


sankey4 <- sankeydata2
ggplot(sankey4, 
       aes(y = value, axis3 = Federal, axis2 = StateFunds, axis1=Expenditures, label = "stratum")) +
  geom_flow(aes(fill = Notes), color = "black") +
 # guides(fill = FALSE) +   
  geom_stratum()+
 # geom_text(stat = "stratum", label.strata = TRUE) + 
coord_flip()+
   scale_fill_brewer(palette = "YlOrRd", direction = -1)+
  theme_void() +
     scale_x_discrete( limits = c("Federal Funds", "State Funds", "Expenditure Categories") )+ 
  #    geom_text_repel(stat = "stratum", aes(label = after_stat(stratum)), size = 4)
  ggtitle("")



# sankey4 <- sankeydata %>% select(Federal, StateFunds, StFund, Expenditures, value, Notes) %>%
#   filter(StFund == "Total") %>%
#   mutate(StateFunds = factor(StateFunds, levels = c("Total_received_fy21", "Total_received_fy22")),
#                   Federal = as.factor(Federal),
#                   Expenditures = as.factor(Expenditures) )

ggplot(sankey4, 
       aes(y = value, 
           axis3 = Federal, axis2 = StateFunds, axis1=Expenditures, label = "stratum")) +
  geom_flow(aes(fill = Notes), color = "black", reverse=FALSE) +
 # guides(fill = FALSE) +   
  geom_stratum(reverse=FALSE)+
 # geom_text(stat = "stratum", label.strata = TRUE) + 
coord_flip()+
   scale_fill_brewer(palette = "YlOrRd", direction = -1)+
  theme_void() +
     scale_x_discrete( limits = c("Federal Funds", "State Funds", "Expenditure Categories") )+ 
      geom_text(stat = "stratum", aes(label = after_stat(stratum)), size = 2, reverse=FALSE)
 # ggtitle("With Labels")
```

```{r}
sankey <- sankeydata %>% select(Federal, StateFunds, StFund, Expenditures, stfundname, value, Notes) %>%
    filter(StFund == "Total") %>%

  mutate(StateFunds = factor(StateFunds, levels = c("Total_received_fy21", "Total_received_fy22")),
         stfundname = as.factor(stfundname),
         Federal = as.factor(Federal),
         
                   Expenditures = factor(Expenditures, levels = c("Transit", "Medicaid", "Public Health",  "K-12", "Federal Other", "Community Development")),
         Notes = factor(Notes, levels = c( "Other *", "CARES","CRRSA", "ARPA")))


ggplot(sankey, 
       aes(y = value, axis3 = StateFunds, axis2 = stfundname, axis1=Expenditures, label = "stratum")) +
  geom_flow(aes(fill = stfundname), color = "black") +
  # guides(fill = FALSE) +   
  geom_stratum()+
  # geom_text(stat = "stratum", label.strata = TRUE) + 
  coord_flip()+
   scale_fill_brewer(palette = "YlOrRd", direction = -1)+
  theme_void() +
     scale_x_discrete( limits = c("Federal Funds", "State Funds", "Expenditure Categories") )+ 
  #    geom_text_repel(stat = "stratum", aes(label = after_stat(stratum)), size = 4)
  ggtitle("")


ggplot(sankey, 
       aes(y = value, axis3 = StateFunds, axis2 = stfundname, axis1=Expenditures, label = "stratum")) +
  geom_flow(aes(fill = stfundname), color = "black",reverse=FALSE) +
  guides(fill = FALSE) +   
  geom_stratum(reverse=FALSE)+
 # geom_text(stat = "stratum", label.strata = TRUE) + 
coord_flip()+
   scale_fill_brewer(palette = "YlOrRd", direction = -1)+
  theme_void() +
     scale_x_discrete( limits = c("Federal Funds", "State Funds", "Expenditure Categories") )+ 
      geom_text(stat = "stratum", aes(label = after_stat(stratum)), size = 2, reverse=FALSE)
 # ggtitle("With Labels")
```

```{r}

ggplot(sankey, 
       aes(y = value, axis3 = Notes, axis2 = stfundname, axis1=Expenditures, label = "stratum")) +
  geom_flow(aes(fill = stfundname), color = "black",reverse=FALSE) +
 # guides(fill = FALSE) +   
  geom_stratum(reverse=FALSE)+
 # geom_text(stat = "stratum", label.strata = TRUE) + 
coord_flip()+
   scale_fill_brewer(palette = "YlOrRd", direction = -1)+
  theme_void() +
     scale_x_discrete( limits = c("Federal Funds", "State Funds", "Expenditure Categories") ) #+ 
    #  geom_text(stat = "stratum", aes(label = after_stat(stratum)), size = 2)
 # ggtitle("With Labels")


ggplot(sankey, 
       aes(y = value, axis3 = Notes, axis2 = stfundname, axis1=Expenditures, label = "stratum")) +
  geom_flow(aes(fill = stfundname), color = "black", reverse=FALSE) +
 # guides(fill = FALSE) +   
  geom_stratum(reverse=FALSE)+
 # geom_text(stat = "stratum", label.strata = TRUE) + 
coord_flip()+
   scale_fill_brewer(palette = "YlOrRd", direction = -1)+
  theme_void() +
     scale_x_discrete( limits = c("Federal Funds", "State Funds", "Expenditure Categories") )+ 
      geom_text(stat = "stratum", aes(label = after_stat(stratum)), size = 2,reverse=FALSE)
 # ggtitle("With Labels")
```

```{r}

sankey <- sankeydata %>% select(Federal, StateFunds, StFund, Expenditures, stfundname, value, Notes) %>%
    filter(StFund == "Total") %>%

  mutate(StateFunds = factor(StateFunds, levels = c( "Total_received_fy21", "Total_received_fy22")),
         stfundname = as.factor(stfundname),
         Federal = as.factor(Federal),
         
                   Expenditures = factor(Expenditures, levels = c("Community Development", "Public Health", "Medicaid", "K-12", "Federal Other", "Transit")),
         Notes = factor(Notes, levels = c(  "CARES", "Other *","CRRSA", "ARPA")))


ggplot(sankey, 
       aes(y = value, axis3 = StateFunds, axis2 = Notes, axis1=Expenditures, label = "stratum")) +
  geom_flow(aes(fill = stfundname), color = "black") +
 # guides(fill = FALSE) +   
  geom_stratum()+
 # geom_text(stat = "stratum", label.strata = TRUE) + 
coord_flip()+
   scale_fill_brewer(palette = "YlOrRd", direction = -1)+
  theme_void() +
     scale_x_discrete( limits = c("Federal Funds", "State Funds", "Expenditure Categories") ) #+ 
    #  geom_text(stat = "stratum", aes(label = after_stat(stratum)), size = 2)
 # ggtitle("With Labels")


ggplot(sankey, 
       aes(y = value, axis3 = StateFunds, axis2 = Notes, axis1=Expenditures, label = "stratum")) +
  geom_flow(aes(fill = stfundname), color = "black", reverse=FALSE) +
 # guides(fill = FALSE) +   
  geom_stratum(reverse=FALSE)+
 # geom_text(stat = "stratum", label.strata = TRUE) + 
coord_flip()+
   scale_fill_brewer(palette = "YlOrRd", direction = -1)+
  theme_void() +
     scale_x_discrete( limits = c("Federal Funds", "State Funds", "Expenditure Categories") ) + 
     geom_text(stat = "stratum", aes(label = after_stat(stratum)), size = 2,  reverse=FALSE)

 # ggtitle("With Labels")

ggplot(sankey, 
       aes(y = value, 
           axis3 = Notes, axis2 = StateFunds, axis1=Expenditures, label = "stratum")) +
  geom_flow(aes(fill = stfundname), color = "black", reverse=FALSE) +
 # guides(fill = FALSE) +   
  geom_stratum(reverse=FALSE)+
 # geom_text(stat = "stratum", label.strata = TRUE) + 
coord_flip()+
   scale_fill_brewer(palette = "YlOrRd", direction = -1)+
  theme_void() +
     #scale_x_discrete( limits = c("Federal Funds", "State Funds", "Expenditure Categories") )+  # doesn't work
      geom_text(stat = "stratum", aes(label = after_stat(stratum)), size = 2, reverse = FALSE)
 # ggtitle("With Labels")
```

> Note: CARES funds were originally received and spent in Disaster Response Funds within Public Safety category. Remaining CARES funds were transfered to State CURE fund for FY22.

> Revenue from Local Cure is the Local Government Transfers.

Using new file called federalcoviddollars.xlsx:

```{r}
library(readxl)
sankeydata <- read_excel("federalcoviddollars.xlsx")
sankey <- sankeydata %>% 
  #select(Federal, StateFunds, StFund, Expenditures, stfundname, value, Notes) %>%
  #  filter(StFund == "Total") %>%

  mutate(StateFunds = factor(FY, levels = c( "FY21", "FY22")),
        Legislation = factor(Legislation2, levels = c( "Other", "CARES", "CRRSA", "ARPA"))
)
         
                 #  Expenditures = factor(Expenditures, levels = c("Community Development", "Public Health", "Medicaid", "K-12", "Federal Other", "Transit")),
      #   Notes = factor(Notes, levels = c(  "CARES", "Other *","CRRSA", "ARPA")))


ggplot(sankey, 
       aes(y = `Dollars Received`, axis3 = FY, axis2 = Recipient, axis1=FF_Cat, label = "stratum")) +
  geom_flow(aes(fill = Legislation), color = "black", reverse=FALSE) +
 # guides(fill = FALSE) +   
  geom_stratum(reverse=FALSE)+
coord_flip()+
   scale_fill_brewer(palette = "YlOrRd", direction = -1)+
  theme_void() +
   #  scale_x_discrete( limits = c("Federal Funds", "State Funds", "Expenditure Categories") ) #+ 
     geom_text(stat = "stratum", aes(label = after_stat(stratum)), size = 2, reverse=FALSE)
 # ggtitle("With Labels")

sankey %>% group_by(FY)%>%summarize(TotalReceived=sum(`Dollars Received`))

sankey %>% group_by(Legislation)%>%summarize(TotalReceived=sum(`Dollars Received`))


sankey %>% group_by(Recipient)%>%summarize(TotalReceived=sum(`Dollars Received`))

sankey %>% group_by(Legislation, Recipient)%>%summarize(TotalReceived=sum(`Dollars Received`)) %>% pivot_wider(names_from = Legislation, values_from = TotalReceived)

sankey %>% group_by(Legislation, FF_Cat)%>%summarize(TotalReceived=sum(`Dollars Received`)) %>% pivot_wider(names_from = Legislation, values_from = TotalReceived)

sankey %>% group_by(FY, FF_Cat)%>%summarize(TotalReceived=sum(`Dollars Received`)) %>% pivot_wider(names_from = FY, values_from = TotalReceived)



```

## Expenditures Graphs

Federal expenditures from CURE and other major funds:

```{r}
cure_exp <- read_xlsx("fedCUREexpenditures.xlsx")

cure_exp <- cure_exp %>%
    #filter(StFund == "Total_spent") %>%

  mutate(
         FY_Spent = factor(FY_Spent, levels = c("2021", "2022", "2023")),
         FY_Received = factor(FY_Received, levels = c("2020", "2021", "2022")))

ggplot(cure_exp, 
       aes(y = `FY Expenditures`, axis3 = FY_Spent, axis2 = Agency, axis1=FF_Cat, label = "stratum")) +
  geom_flow(aes(fill = Law), color = "black", reverse=FALSE) +
  # guides(fill = FALSE) +   
  geom_stratum(reverse=FALSE)+
 geom_text(stat = "stratum", label.strata = TRUE, reverse=FALSE) + 
  coord_flip()+
   scale_fill_brewer(palette = "YlOrRd", direction = -1)+
  theme_void() + 
  ggtitle("")

ggplot(cure_exp, 
       aes(y = `FY Expenditures`, axis3 = `Federal Funds`, axis2 = FY_Spent, axis1=Fund, label = "stratum")) +
  geom_flow(aes(fill = Law), color = "black", reverse=FALSE) +
  # guides(fill = FALSE) +   
  geom_stratum(reverse=FALSE)+
 geom_text(stat = "stratum", label.strata = TRUE, reverse=FALSE) + 
  coord_flip()+
   scale_fill_brewer(palette = "YlOrRd", direction = -1)+
  theme_void() + 
        geom_text(stat = "stratum", aes(label = after_stat(stratum)), size = 2, reverse=FALSE)+

  ggtitle("")

ggplot(cure_exp, 
       aes(y = `FY Expenditures`,axis3 = `Federal Funds`, axis4 = `FY_Received`,axis2 = FY_Spent, axis1=FF_Cat, label = "stratum")) +
  geom_flow(aes(fill = Law), color = "black", reverse=FALSE) +
  # guides(fill = FALSE) +   
  geom_stratum(reverse=FALSE)+
 geom_text(stat = "stratum", label.strata = TRUE, reverse=FALSE) + 
  coord_flip()+
   scale_fill_brewer(palette = "YlOrRd", direction = -1)+
  theme_void() + 
        geom_text(stat = "stratum", aes(label = after_stat(stratum)), size = 2, reverse=FALSE)+

  ggtitle("")

ggplot(cure_exp, 
       aes(y = `FY Expenditures`,axis3 = `Federal Funds`,axis2 = Agency, axis1=FF_Cat, label = "stratum")) +
  geom_flow(aes(fill = Law), color = "black", reverse=FALSE) +
  # guides(fill = FALSE) +   
  geom_stratum(reverse=FALSE)+
 geom_text(stat = "stratum", label.strata = TRUE, reverse=FALSE) + 
  coord_flip()+
   scale_fill_brewer(palette = "YlOrRd", direction = -1)+
  theme_void() + 
        geom_text(stat = "stratum", aes(label = after_stat(stratum)), size = 2, reverse=FALSE)+

  ggtitle("")

ggplot(cure_exp, 
       aes(y = `FY Expenditures`,axis3 = `Law`,axis2 = Agency, axis1=FF_Cat, label = "stratum")) +
  geom_flow(aes(fill = `Federal Funds`), color = "black", reverse=FALSE) +
  # guides(fill = FALSE) +   
  geom_stratum(reverse=FALSE)+
 geom_text(stat = "stratum", label.strata = TRUE, reverse=FALSE) + 
  coord_flip()+
   scale_fill_brewer(palette = "YlOrRd", direction = -1)+
  theme_void() + 
        geom_text(stat = "stratum", aes(label = after_stat(stratum)), size = 2, reverse=FALSE)+

  ggtitle("")
```

Dates on top are Fiscal Year received. Dates in the middle of the graph are Fiscal Year expenditures. Federal funds for COVID recovery have been received and spent in different years. This makes calculating and comparing the fiscal gap difficult.

```{r works, warning=FALSE, message=FALSE, eval=FALSE}
sankey2 <- read_csv("sankeyattempt.csv") %>% select(source, target, value)
# From these flows we need to create a node data frame: it lists every entities involved in the flow

nodes <- data.frame(
  name=c(as.character(sankey2$source), 
  as.character(sankey2$target)) %>% unique()
)
 
# With networkD3, connection must be provided using id, not using real name like in the links dataframe.. So we need to reformat it.
sankey2$IDsource <- match(sankey2$source, nodes$name)-1 
sankey2$IDtarget <- match(sankey2$target, nodes$name)-1
 

p <- sankeyNetwork(Links = sankey2, Nodes = nodes,
              Source = "IDsource", Target = "IDtarget",
              Value = 'value', NodeID = "name", nodeWidth = 30,
              sinksRight=TRUE)
              
              #align = "center", units = "Billions", title = "Federal Funds to State Expenditures FY 2021", showNodeValues = TRUE)
p

# Add a 'group' column to the nodes data frame:
nodes$group <- as.factor(c("a","b","b","b","b", "b","b", "c", "c", "c", "c", "c"))
 
# Give a color for each group:
my_color <- 'd3.scaleOrdinal() .domain(["a", "b", "c"]) .range(["#69b3a2", "steelblue", "purple"])'


sankeyNetwork(Links = sankey2, Nodes = nodes,
              Source = "IDsource", Target = "IDtarget",
              Value = "value", NodeID = "name",
              colourScale = my_color, NodeGroup = "group", nodeWidth = 40, fontSize = 14, units = "Billions of Dollars"
              )
```

[ggsankey info and example](https://cheatography.com/seleven/cheat-sheets/ggalluvial/)

> Note: TANF, SNAP, Increased Matching Grant, Medicare Continuous Coverage Requirement, all count as Federal Medicaid Revenue.
