# State Employee Healthcare Discussion

```{r warning = FALSE, message=FALSE}

library(tidyverse)
library(haven)
library(formatR)
library(lubridate)
library(smooth)
library(forecast)
library(scales)
library(kableExtra)
library(ggplot2)
library(readxl)
library(tidyverse)
library(data.table)
library(quantmod)
library(geofacet)
library(janitor)


knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

```

State Employee Health Care = Sum of expenditures for "health care coverage as elected by members per state employees group insurance act." The payments are made from the Health Insurance Reserve Fund. Employee contributions are not considered a revenue source or an expenditure in our analysis.

Funding for the State Employees Group Insurance plan originates from two funds. The Health Insurance Reserve Fund (HIRF) and the Group Insurance Premium Fund (GIPF). Contributions and payment for Health coverage benefits are deposited INTO HIRF and contributions for life insurance are deposited into the GIPF.

HIRF is the fund mainly used to administer the group insurance program. Funding for HIRF comes from several different revenue sources, the General Revenue Fund (GRF), Road Fund, reimbursements, university funds, and misc funds. [CGFA Report](https://cgfa.ilga.gov/Upload/fy2011stateemployees'groupinsurance.pdf).

**Coding details**

In FY2013, the Local Government Health Insurance fund was transferred to the department of Central Management Services (agency changes from 478 to 416 in data.)

Employer group insurance contributions for health insurance are excluded to avoid double counting the cost of healthcare provision. All employer group insurance contributions are coded as object = 1180. BUT the last two fiscal years were coded as 1900 instead of 1180 for lump sums instead of employer contributions --- anything to do with pandemic money categorization?

-   Fund = 0907 = health insurance reserve, in_ff = 1
-   Fund = 0457 is "Group insurance premium", in_ff = 1
-   Fund = 0193 is "Local govt health insurance reserve", in=ff = 0
-   fund = 0477 is "Community College Health Insurance", in=ff = 0.
    -   had large amount in early years
-   Fund = 9939 is "group self-insurers' insolv", in_ff = 1
-   Fund = 0940 is Self-Insurers security, in_ff = 0
-   Fund = 0739 is Group Workers Comp Pool Insol, in_ff = 1
    -   eehc = 0 means it is NOT a state healthcare cost but it is an employer contribution of some type to some fund

    -   eehc = 1 means it is a state employee healthcare cost and it is an employer contribution to health insurance

```{r healthcare-employer-contributions}
exp_temp <- read_csv("exp_temp.csv")
rev_temp <- read_csv("rev_temp.csv")


health_ins_reserve <- exp_temp %>% filter(fund == "0907") %>%  group_by(fy) %>% 

    summarize(fund_0907 = sum(expenditure)) 

health_ins_reserve %>% 

  ggplot(aes(x=fy, y=fund_0907)) + geom_line() + labs(title="Health Insurance Reserve", subtitle = "Sum of expenditures from fund 907")

# object 1180 is inconsistently coded over time form the IOC 

# object 1180 should be employer contributions to healthcare group insurance

employer_contributions <- exp_temp %>% filter(object == "1180") %>% group_by(fy) %>% summarize(object1180 = sum(expenditure)) 

employer_contributions%>% 

  ggplot(aes(x=fy, y=object1180)) + geom_line() + labs(title="Employer Contributions to Healthcare Group Insurance, IOC Object 1180")

employer_contributions2 <- exp_temp %>% filter(object == "1180" & fund=="0001") %>% group_by(fy) %>% summarize(object1180 = sum(expenditure)) 

employer_contributions2 %>% 

  ggplot(aes(x=fy, y=object1180)) + geom_line() + labs(title="Employer Contributions to Healthcare Group Insurance", subtitle = "IOC Object 1180 from Fund 001")

# examine combined group insurance totals per year

group_ins2 <- exp_temp %>% 

  mutate(eehc = ifelse(

    # group insurance contributions for 1998-2005 and 2013-present

    # CMS took over health insurance in 2013

   fund == "0001" & (object == "1180" | object =="1900") & agency == "416" & appr_org=="20", 1, 0) )%>% 

  mutate(eehc = ifelse(

    # group insurance contributions for 2006-2012

    # health insurance was in healthcare and family services, agency 478 for a few years

    fund == "0001" & object == "1180" & agency == "478" & appr_org=="80", 1, eehc) )%>%

  filter(eehc == 1) %>% 

    group_by(fy) %>% 

    summarize(dropped_group_premiums = sum(expenditure, na.rm=TRUE))

group_ins2 %>% ggplot(aes(x=fy, y=dropped_group_premiums)) + geom_line() + labs(title="Employer Healthcare Group Insurance Contributions", subtitle= " - Dropped from analysis to avoid double counting healthcare expenditures", caption = "Objects 1180 and 1900 from fund 0001. See code for additional coding details.")
```

If observation is a group insurance contribution, then the expenditure amount is set to \$0 (essentially dropped from analysis).

### Health Insurance Premiums - Revenue Side

Employee insurance premiums for healthcare are a revenue source for the state in the IOC data but are NOT included in the Fiscal Futures analysis and fiscal gap calculations.

Source \#'s:

-   0120 = ins prem-option life
-   0120 = ins prem-optional life/univ
-   0347 = optional health - HMO
-   0348 = optional health - dental
-   0349 = optional health - univ/local SI
-   0350 = optional health - univ/local
-   0351 = optional health - retirement
-   0352 = optional health - retirement SI
-   0353 = optional health - retire/dental
-   0354 = optional health - retirement hmo
-   2199-2209 = various HMOs, dental, health plans from Health Insurance Reserve (fund)

```{r employee-insurance-premiums}

health_insurance_fund_rev<- rev_temp %>% 

  filter(fund=="0907") %>% 

    group_by(fy) %>%

  summarize(health_ins_rev = sum(receipts)) 

health_insurance_fund_rev %>% 

  ggplot(aes(x=fy, y = health_ins_rev)) + 

  geom_line() + labs( title = "Health insurance fund - All revenue, Fund 0907")

#collect optional insurance premiums to fund 0907 for use in eehc expenditure  

employee_health_premiums <- rev_temp %>% 

  mutate(employee_premiums = ifelse(

    fund=="0907" & (source=="0120"| source=="0121"| (source>"0345" & source<"0357")|(source>"2199" & source<"2209")), 1, 0)) %>%

  filter(employee_premiums == 1)

# optional insurance premiums = employee insurance premiums

emp_premium <- employee_health_premiums %>%

  group_by(fy) %>%

  summarize(employee_premiums_sum = sum(receipts))

  

emp_premium %>% ggplot(aes(x=fy, y = employee_premiums_sum)) + 

  geom_line() + labs( title = "Employee health insurance premiums")

# contributions and benefits paid comparison

ggplot()+

  #  geom_line(data=group_ins, aes(x=fy, y=object1180, color='Group Insurance1')) +

      geom_line(data=health_insurance_fund_rev, aes(x=fy, y=health_ins_rev, color='Health Insurance Fund - All Revenue')) +

 geom_line(data = emp_premium, aes(x=fy, y = employee_premiums_sum, color = 'Revenue from Employee Premiums')) +

    geom_line(data=health_ins_reserve, aes(x=fy, y=fund_0907, color='Cost of Provision')) +

    geom_line(data=employer_contributions, aes(x=fy, y=object1180, color='Group Insurance-Object1180')) +

 #   geom_line(data=employer_contributions2, aes(x=fy, y=object1180, color='Employer Contributions-General Fund')) +

  geom_line(data=group_ins2, aes(x=fy, y=dropped_group_premiums, color='Group Insurance - 1180 & 1900')) +

  #geom_line(data= healthcare_costs, aes(x=fy, y = cost_of_provision, color = 'Healthcare Costs'))+ 

  scale_color_manual(values = c(

    'Cost of Provision' = 'darkblue',

    'Health Insurance Fund - All Revenue' = 'light green',

    'Revenue from Employee Premiums' = 'dark green',

    'Group Insurance - 1180 & 1900' = 'blue',

    'Group Insurance-Object1180' = 'light blue'

   #     'Employer Contributions-General Fund' = 'light blue'

)) +

  labs(title="Healthcare costs and group insurance contributions", 

       caption = "Healthcare costs and group insurance contributions", 

       y = "Dollars", x = "")

exp_temp <- exp_temp %>% 

  mutate(eehc = ifelse(object == "1180", 1, 0)) %>%

  mutate(eehc = ifelse((eehc == 1 & in_ff =="0"), 2, eehc))

#%>%mutate(in_ff = ifelse(eehc == 2, "1", in_ff) ) %>% filter(eehc ==2 )

table(exp_temp$eehc)
```
