# K-12 Spending & ESSER Funds

```{r warning=FALSE, message=FALSE, echo=FALSE}

library(tidyverse)
library(haven)
library(formatR)
library(scales)
library(kableExtra)
library(ggplot2)
library(readxl)
library(tidyverse)
library(data.table)
library(janitor)
library(cmapplot)
library(ggrepel)

knitr::opts_chunk$set(warning = FALSE, message = FALSE)
first_year = 1998
current_year = 2025

rev_temp <- read_csv(paste0("data/FY", current_year, " Files/rev_temp.csv"))

exp_temp <- read_csv(paste0("data/FY", current_year, " Files/exp_temp.csv"))

alea_theme <- function() {
  font <-"Whitney"
  
  ggplot2::theme(
    legend.position = "right",
    legend.title = element_blank(),
    
    panel.background = ggplot2::element_blank(),
    panel.grid.minor.x = ggplot2::element_blank(),
    panel.grid.major.y = element_line(color = "grey"),
    panel.grid.minor.y = element_line(color = "grey", 
                                      linetype = "dashed"),
     panel.grid.major.x = ggplot2::element_blank(),
    axis.ticks = element_line(color = "gray"),
    axis.ticks.x = element_blank()
  )
  
}

theme_set(alea_theme())

exp_temp <- exp_temp %>%   filter(agency != "799") |> 
  mutate(fund_name = fund_name_ab)

rev_temp <- rev_temp %>% 
  filter(agency != "799"  & rev_type != "75")  %>% mutate(fund_name = fund_name_ab)
```

```{r include=FALSE}

update_recessions <- function(url = NULL, quietly = FALSE){

  # Use default URL if user does not override
  if (is_null(url) | missing(url)) {
    url <- "https://data.nber.org/data/cycles/business_cycle_dates.json"
  }
  
   # locally bind variable names
  start_char <- end_char <- start_date <- end_date <- ongoing <- index <- peak <- trough <- NULL

  return(
    # attempt to download and format recessions table
    tryCatch({
      recessions <- jsonlite::fromJSON(url) %>%
        # drop first row trough
        dplyr::slice(-1) %>%
        # convert peaks and troughs...
        dplyr::mutate(
          # ...to R dates
          start_date = as.Date(peak),
          end_date = as.Date(trough),
          # ... and clean char strings
          start_char = format(start_date, "%b %Y"),
          end_char = format(end_date, "%b %Y")) %>%
        # confirm ascending and create row number
        dplyr::arrange(start_date) %>%
        mutate(index = row_number()) %>%
        mutate(
          # Flag unfinished recessions
          ongoing = case_when(
            is.na(end_date) & index == max(.$index) ~ T,
            TRUE ~ F),
          # set ongoing recession to arbitrary future date
          end_date = case_when(
            ongoing ~ as.Date("2200-01-01"),
            TRUE ~ end_date),
          # mark ongoing recession in char field
          end_char = case_when(
            ongoing ~ "Ongoing",
            TRUE ~ end_char)
          ) %>%
        # clean up
        select(start_char, end_char, start_date, end_date, ongoing)

      if (!quietly) {message("Successfully fetched from NBER")}

      # Return recessions
      recessions
    },
    error = function(cond){
      if (!quietly) message("WARNING: Fetch or processing failed. `NULL` returned.")
      return(NULL)
    }
    )
  )
}

#recess_table <- list(build_recessions(update))
recessions <- update_recessions()

```


-   \$3.4266 billion Total spent through FY23 from ARP and ESSER Funds. - 1.694 billion spent in 2023 (ARP = ESSER III) but that money arrived in 2022 (=\> Fiscal Gap Implications!)

-   \$2.34 Billion of ARP Emergency Relief Funds (aka ESSER III) spent in 2022 and 2023 combined

-   583 million for ESSER II in 2021

-   376 million with ESSER I in 2021 and 128 million in 2020.

-   ISBE received around an additional \$7.9 billion total from ESSER compared to its normal federal funding because of COVID

-   At end of State's FY23, \$3.43/\$7.9 billion = \~ 50% spent

-   As of Dec.7th, it has spent 74.1% of these funds ("ESSER Spending Dashboard" 2023)

-   640 million allocated for FY24 (according to CPS, I think they have different fiscal years though...)

```{r}
#| layout-ncol: 2
#| column: screen
#|
rev_temp <- rev_temp %>% 
  mutate(
    rev_type = ifelse(rev_type=="57" & agency=="478" & (source=="0618"|source=="2364"|source=="0660"|source=="1552"| source=="2306"| source=="2076"|source=="0676"|source=="0692"), "58", rev_type),
    rev_type_name = ifelse(rev_type=="58", "Federal Medicaid Reimbursements", rev_type_name),
    rev_type = ifelse(rev_type=="57" & agency=="494", "59", rev_type),
    rev_type_name = ifelse(rev_type=="59", "Federal Transportation", rev_type_name),
    rev_type_name = ifelse(rev_type=="57", "Federal - Other", rev_type_name),
    rev_type = ifelse(rev_type=="6", "06", rev_type),
    rev_type = ifelse(rev_type=="9", "09", rev_type)) 

rev_temp %>% 
  filter(rev_type == "57") %>% 
  group_by(fy, rev_type, rev_type_name) %>% 
  summarise(receipts = sum(receipts, na.rm = TRUE)/1000000000) %>% 
  ggplot(aes(x=fy, y=receipts, color=rev_type_name)) +
  geom_recessions(xformat = "numeric",text = FALSE)+
    geom_smooth(aes(x=fy, y = receipts), method = lm, se = FALSE, lty = "dashed", color = "red", alpha = .5, lwd = .5) + 

  geom_line(aes(x=fy, y=receipts,color=rev_type_name), lwd = 1.5) +
  scale_y_continuous(labels = comma)+
  scale_x_continuous(expand = c(0,.5), n.breaks = 6) +
  labs(title = "Other Federal Revenue
       with federal stimulus funds", 
       caption = "These values include stimulus funds after the Great Recession and the COVID pandemic response.",
       y = "Billions of Dollars", x = "") + 
  theme(legend.position = "bottom", legend.title = element_blank()  )

rev_temp %>% 
  filter(rev_type == "57") %>% 
  filter(source_name_AWM != "FEDERAL STIMULUS PACKAGE" & source_name_AWM != "STATE CURE") %>%
  group_by(fy, rev_type, rev_type_name) %>% 
  summarise(receipts = sum(receipts, na.rm = TRUE)/1000000000) %>% 
  ggplot(aes(x=fy, y=receipts, color=rev_type_name) )+
  geom_recessions(xformat = "numeric",text = FALSE)+
  geom_line(aes(x=fy, y=receipts, color=rev_type_name), lwd = 1.5) +
  geom_smooth(aes(x=fy, y = receipts), method = lm, se = FALSE, lty = "dashed", color = "red", alpha = .5, lwd = .5) + 
  scale_y_continuous(limits = c(0,20), labels = comma)+
  scale_x_continuous(expand = c(0,.5), n.breaks = 6) +
  labs(title = "Federal to State Transfers: All Other Fed Rev 
       without Federal Stimulus $", 
       caption = "These values do NOT include stimulus funds after the Great Recession and the COVID pandemic response.",
       y = "Billions of Dollars", x = "") + 
  theme(legend.position = "bottom", legend.title = element_blank()  )
```

```{r}
#| label: fig-k12-federal-revenue

rev_temp %>% 
  filter(rev_type == "57" #fund_name == "SBE FEDERAL DEPT OF EDUCATION"
         ) %>% 
  group_by(fy) %>% 
  summarise(receipts = sum(receipts, na.rm = TRUE)) %>%
  mutate(adj_receipts = case_when(
    fy==2020 ~ receipts - 677964975, # ESSER 1 from CARES
    fy==2022 ~ receipts - 7353697183, # ARPA ESSER3 + CRRSA ESSER2 $ to ISBE
  )
         )
  


rev_temp %>% 
  filter(fund_name == "SBE FEDERAL DEPT OF EDUCATION") %>% 
  group_by(fy, rev_type, rev_type_name) %>% 
  summarise(receipts = sum(receipts, na.rm = TRUE)/1000000000) %>% 
  ggplot(aes(x=fy, y=receipts, color=rev_type_name)) +
  geom_recessions(xformat = "numeric",text = FALSE)+
    geom_smooth(aes(x=fy, y = receipts), method = lm, se = FALSE, lty = "dashed", color = "red", alpha = .5, lwd = .5) + 
  geom_line(aes(x=fy, y=receipts,color=rev_type_name), lwd = 1.5) +
  scale_y_continuous(limits = c(0,4), labels = comma)+
  scale_x_continuous(expand = c(0,.5), n.breaks = 6) +
  labs(title = "K-12 Federal Revenue (all revenue received)", 
       caption = "All funds received by the SBE fund",
       y = "Billions of Dollars", x = "") + 
  theme(legend.position = "bottom", legend.title = element_blank()  )



rev_temp %>% 
  filter(rev_type == "57" & fund_name == "SBE FEDERAL DEPT OF EDUCATION") %>% 
  filter(source_name_AWM != "FEDERAL STIMULUS PACKAGE") %>%
  group_by(fy, rev_type, rev_type_name) %>% 
  summarise(receipts = sum(receipts, na.rm = TRUE)/1000000000) %>% 
  ggplot(aes(x=fy, y=receipts, color=rev_type_name) )+
  geom_recessions(xformat = "numeric",text = FALSE)+
  geom_line(aes(x=fy, y=receipts, color=rev_type_name), lwd = 1.5) +
  geom_smooth(aes(x=fy, y = receipts), method = lm, se = FALSE, lty = "dashed", color = "red", alpha = .5, lwd = .5) + 
  scale_y_continuous(limits= c(0,4),labels = comma)+
  scale_x_continuous(expand = c(0,0)) +
  labs(title = "Federal Revenue for K-12", 
       caption = "These values do NOT include stimulus funds after the Great Recession and the COVID pandemic response.",
       y = "Billions of Dollars", x = "") + 
  theme(legend.position = "bottom", legend.title = element_blank()  )
```

```{r}
K12rev <- rev_temp %>% 
  filter( fund_name == "SBE FEDERAL DEPT OF EDUCATION" &  agency_name == "STATE BOARD OF EDUCATION") %>%
  group_by(fy) %>% 
  summarize(receipts =round(sum(receipts)/1000000)) %>% 
  arrange(-receipts)

K12exp <- exp_temp %>% 
  filter( fund_name == "SBE FEDERAL DEPT OF EDUCATION" & agency_name == "STATE BOARD OF EDUCATION") %>%
  group_by(fy) %>% 
  summarize(expenditure =round(sum(expenditure)/1000000)) %>% 
  arrange(-expenditure)

ggplot() +
  geom_line(data = K12rev, aes(x=fy, y = receipts), lty = 1 ) +
  geom_line(data = K12exp, aes(x=fy, y = expenditure), lty = 2) +
  scale_x_continuous(expand = c(0,.5), n.breaks = 6)
```

```{r}
#| collapse: false
#| code-fold: false

# looks at ISBE federal funding after 2018
# gives a sort of base comparison of funding before pandemic
# ESSER revenue sources had multiple source names so lots of scrolling and CTRL-F searching was done
rev_temp %>% 
  filter(
  #rev_type == "57" & 
           fy > 2018 & fund_name == "SBE FEDERAL DEPT OF EDUCATION") %>%
  group_by(source_name_AWM , fund_name, fy) %>% 
  summarize(receipts =round(sum(receipts)/1000000)) %>% 
  arrange(-receipts)
```

```{r}
#| fig-cap: "All Federal Revenue"
#| layout-ncol: 2
#| fig-show: hold
#| fig-column: screen

fedrev<- rev_temp %>% 
  filter(rev_type == "58" | rev_type == "59" | rev_type == "57") 

fedrev %>% 
  group_by(fy) %>% 
  summarise(receipts = sum(receipts, na.rm = TRUE)/1000000000) %>% 
  ggplot() +
  geom_line(aes(x=fy, y=receipts)) +
  labs(title = "All Federal Revenue w/ Fed. Stimulus Package", y = "Billions of Dollars", x = "") + 
  theme(legend.position = "bottom", legend.title = element_blank()  )+
  scale_x_continuous(expand = c(0,.5), n.breaks = 6) +
  scale_y_continuous(limits = c(0,45), labels = comma)


fedrev %>% 
  filter(source_name_AWM != "FEDERAL STIMULUS PACKAGE") %>%
  group_by(fy) %>% 
  summarise(receipts = sum(receipts, na.rm = TRUE)/1000000000) %>% 
  ggplot() +
  geom_line(aes(x=fy, y=receipts)) +
  labs(title = "All Federal Rev EXCEPT Federal Stimulus Package", 
       y = "Billions of Dollars", x = "",
       caption = "Note: Dropping Federal Stimulus Package revenue only removes the $3.5 billion from FY20, $0.3 billion from FY21, and 
       $8.5 billion from FY22. There is still over $12 billion more in Federal Revenue compared to FY19.") + 
  theme(legend.position = "bottom", legend.title = element_blank()  ) +
  scale_x_continuous(expand = c(0,.5), n.breaks = 6) +
  scale_y_continuous(limits = c(0,45), labels = comma)
```

```{r}
#| label: tbl-esserperyear
#| code-fold: false
#| tbl-cap: "All Appropriations made by State Board of Education - funds spent per state fiscal year"

## Looking at ESSER funds spent per year ## 

exp_temp %>% 
  filter(fy >2018 & fund_name == "SBE FEDERAL DEPT OF EDUCATION" & 
           agency_name == "STATE BOARD OF EDUCATION") %>% 
  group_by(wh_approp_name) %>% 
  summarize(expenditures = round(sum(expenditure)/1000000, digits = 2))
```

Key words and phrases from the appropriations above were used to identify appropriations that specifically used ESSER funds.

```{r}
K12_ESSER_words <- c("CRRSA","ESSER","EMER R", "EMR R", "CARES", "AMER R", "EMER ED")

K12 <- exp_temp %>% 
  filter(agency_name == "STATE BOARD OF EDUCATION") %>%
  mutate(ESSERfunds = case_when(
    str_detect(wh_approp_name, "CRRSA") ~ "ESSER",
    str_detect(wh_approp_name, "ESSER") ~ "ESSER",
    str_detect(wh_approp_name, "EMER R") ~ "ESSER",
    str_detect(wh_approp_name, "EMR R") ~ "ESSER",
    str_detect(wh_approp_name, "CARES") ~ "ESSER",
    str_detect(wh_approp_name, "AMER R") ~ "ESSER",
    str_detect(wh_approp_name, "EMER ED") ~ "ESSER",
    
    TRUE ~ 'not_esser')) %>%
  filter(ESSERfunds == "ESSER")
```

```{r}
#| tbl-column: margin
#| label: tbl-table3
#| tbl-cap: "Amount of ESSER money spent total (all years summed together)"
#| collapse: false
#| code-fold: false


# amount of ESSER money spent total (all years summed together)

K12 %>% 
  summarize(esser_spent = sum(expenditure))
```

```{r}
#| label: tbl-fig3
#| code-fold: false
#| collapse: false
#| tbl-column: margin
#| tbl-cap: "ESSER spent each year"

# grouped by fiscal year
K12 %>% 
  group_by(fy) %>% 
  summarize(esser_spent = sum(expenditure))
```

```{r}
#| collapse: false
#| label: tbl-table4
#| code-fold: false
#| tbl-cap: ESSER items grouped by fiscal year and appropriation description



# Grouped by fiscal year and appropriation description
K12 %>% 
  group_by(fy, wh_approp_name) %>% 
  summarize(esser_spent = sum(expenditure))
```
