# Medicaid

```{r warning = FALSE, message=FALSE}

library(tidyverse)
library(haven)
library(formatR)
library(lubridate)
library(smooth)
library(forecast)
library(scales)
library(kableExtra)
library(ggplot2)
library(readxl)
library(tidyverse)
library(data.table)
library(quantmod)
library(geofacet)
library(janitor)


knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

exp_temp <- read_csv("exp_temp.csv")
rev_temp <- read_csv("rev_temp.csv")

```

## Federal Medicaid Reimbursements and Medicaid Costs

```{r}
medicaid_cost <- exp_temp %>% 
  filter(agency=="478" & (appr_org=="01" | appr_org == "65" | appr_org=="88") & (object=="4900" | object=="4400")) %>% 
  group_by(fy) %>% 
  summarize(sum=sum(expenditure))

med_reimburse <- rev_temp %>% 
  filter(rev_type=="57" & agency=="478" & (source=="0618"|source=="2364"|source=="0660"|source=="1552"| source=="2306"| source=="2076"|source=="0676"|source=="0692")) %>% 
  group_by(fy) %>% 
  summarize(sum=sum(receipts))



ggplot()+
  geom_line(data=medicaid_cost, aes(x=fy, y=sum), color = "red") + 
  geom_line(data=med_reimburse, aes(x=fy, y = sum), color="black") + 
  labs(title = "Medicaid reimbursements and Medicaid expenditures", 
       caption = "Medicaid expenditures include funds provided to medical providers. ")
```

**Medicaid.**That portion of the Healthcare and Family Services (or Public Aid in earlier years, agency code 478) budget for Medical (appr_organization code 65) for awards and grants (object codes 4400 and 4900).

> State CURE revenue will remain in the Medicaid category due to the nature of it being federal funds providing public health services and funding to locations that provide public services.

-   Uses same appropriation name of "HEALTHCARE PROVIDER RELIEF" and fund == 0793 and obj_seq_type == 49000000. So can defend the "mistake" of including healthcare provider relief as Medicaid expenditure.

```{r Medicaid-check, include = FALSE, eval=FALSE}
medicaid_check <- exp_temp %>% filter(agency=="478" & (appr_org=="01" | appr_org == "65" | appr_org=="88") & (object=="4900" | object=="4400"))

# This includes the State CURE fund

medicaid_check %>% group_by(fy) %>% summarize(sum = sum(expenditure)) %>% arrange(-fy)  # looks good I think AWM

# if we want state CURE as public health or other group number

# exp_temp <- exp_temp %>% mutate(group = if_else(fund == "0324" & agency=="478" & appr_org == "65" & object=="4900", "478", as.character(group)))

medicaid_check %>% filter(fy>2020) %>% group_by(wh_approp_name,fy) %>% summarize(sum=sum(expenditure)) %>% arrange(wh_approp_name)

```

Federal Medical Assistance Program (FMAP): in 1965. The FMAP formula compares the state per-capita income to the national per-capita income. There is no cap on the dollar amount that the federal government pays, so the more that a state spends the more that it receives. a maximum of 83%. States with a higher per-capita income receive lower FMAP funding but no less than 50%, and the states that have a lower per-capita income receive higher FMAP funding. Those that need more, get more.
