# Local Transfers

Separate transfers to local from parent agencies that come from DOR(492) or Transportation (494). Treats muni revenue transfers as expenditures, not negative revenue.

The share of certain taxes levied state-wide at a common rate and then transferred to local governments. (Purely local-option taxes levied by specific local governments with the state acting as collection agent are NOT included.)

The six corresponding revenue items are:

-   Local share of Personal Income Tax
    -   Individual Income Tax Pass-Through New 2021 (source 2582).
-   Local share of General Sales Tax
-   Personal Property Replacement Tax on Business Income
-   Personal Property Replacement Tax on Public Utilities
-   Local share of Motor Fuel Tax
-   Transportation Renewal Fund 0952

Until Dec 18. 2022, Local CURE was being aggregated into Revenue totals since the agency was the Department of Revenue. However the \$371 million expenditure is for "LOC GOVT ARPA" and the revenue source that is Local CURE is also \$371 million. Since it cancels out and is just passed through the state government, I am changing changing the fund_ab_in file so that in_ff=0 for the Local CURE fund. It also inflates the department of revenue expenditures in a misleading way when the expense is actually a transfer to local governments.

-   Dropping Local CURE fund from analysis results in a \$371 million decrease in the department of Revenue (where the Local Government ARPA transfer money). The appropriation for it was over \$740 million so some will probably be rolled over to FY23 too.

-   In the FY21 New and Reused Funds word document, 0325 Local CURE is described as *"Created as a federal trust fund. The fund is established to receive transfers from either the disaster response and recovery fund or the state cure fund of federal funds received by the state. These transfers, subject to appropriation, will provide for the administration and payment of grants and expense reimbursements to units of local government. Revenues should be under Federal Other and expenditures under Commerce and Economic Opportunity."* - Changed to Exclude.

```{r warning = FALSE, message=FALSE}

library(tidyverse)
library(haven)
library(formatR)
library(lubridate)
library(smooth)
library(forecast)
library(scales)
library(kableExtra)
library(ggplot2)
library(readxl)
library(tidyverse)
library(data.table)
library(quantmod)
library(geofacet)
library(janitor)


knitr::opts_chunk$set(warning = FALSE, message = FALSE)

exp_temp <- read_csv("exp_temp.csv")
rev_temp <- read_csv("rev_temp.csv")
```

```{r}
exp_temp <- exp_temp %>% 
  mutate(
    agency = case_when(
      fund=="0515" & object=="4470" & type=="08" ~ "971", # income tax to local governments
      
      fund=="0515" & object=="4491" & type=="08" & sequence=="00" ~ "971", # object is shared revenue payments
      
      fund=="0802" & object=="4491" ~ "972", #pprt transfer
      
      fund=="0515" & object=="4491" & type=="08" & sequence=="01" ~ "976", #gst to local
      
      fund=="0627" & object=="4472"~ "976" , # public transportation fund but no observations exist
      
      fund=="0648" & object=="4472" ~ "976", # downstate public transportation, but doesn't exist
      
      fund=="0515" & object=="4470" & type=="00" ~ "976", # object 4470 is grants to local governments
      
      object=="4491" & (fund=="0188"|fund=="0189") ~ "976",
      
      fund=="0187" & object=="4470" ~ "976",
      
      fund=="0186" & object=="4470" ~ "976",
      
      object=="4491" & (fund=="0413"|fund=="0414"|fund=="0415")  ~ "975", #mft to local
      fund == "0952"~ "975", # Added Sept 29 2022 AWM. Transportation Renewal MFT
      TRUE ~ as.character(agency)),
    
    agency_name = case_when(
      agency == "971"~ "INCOME TAX 1/10 TO LOCAL",
      agency == "972" ~ "PPRT TRANSFER TO LOCAL",
      agency == "975" ~ "MFT TO LOCAL",
      agency == "976" ~ "GST TO LOCAL",
      TRUE~as.character(agency_name))) %>% 
  mutate(group = ifelse(agency>"970" & agency < "977", as.character(agency), ""))


transfers_long <- exp_temp %>% 
  filter(group == "971" |group == "972" | group == "975" | group == "976")
```

```{r graph-local-transfers}

transfers_long %>% 
  group_by(fy, group ) %>%
  summarize(sum_expenditure = sum(expenditure)/1000000) %>%
  pivot_wider(names_from = "group", values_from = "sum_expenditure", names_prefix = "exp_" )

transfers_long %>% 
  group_by(agency_name, group, fy) %>% 
  summarize(expenditure = sum(expenditure, na.rm=TRUE) )%>% 
  ggplot() + 
  geom_line(aes(x=fy, y = expenditure, color=agency_name)) + 
  labs(title = "Transfers to Local Governments", caption = "Data Source: Illinois Office of the Comptroller")

```

Large increases in local transfers were driven by large increases in overall tax revenue collected by the state.
